### 2011-06-29 projects/slrsim/paper/slrsim.tex
Add manuscript tex files.

\documentclass{mbe}
\usepackage{natbib}
\setlength{\bibsep}{0.0pt}

\begin{document}

\newcommand{\pranka}{PRANK\_A}
\newcommand{\prankc}{PRANK\_C}
\newcommand{\omg}{\bm{\omega}}
\newcommand{\tpr}{TPR$_{1\%}$}

\title{The effects of alignment error and alignment filtering on the sitewise detection of positive selection}

\author{{Gregory Jordan and Nick Goldman}\\
\affiliation{European Bioinformatics Institute, Wellcome Trust Genome Campus, Hinxton, Cambridgeshire, CB10 1SD, United Kingdom}}
\authorlist{Jordan and Goldman}
\shorttitle{Alignment eror and detecting sitewise positive selection}

\mbeabstract{%

The effects of alignment error on many types of evolutionary inference
are not well understood. Alignment filters are commonly applied to try
to reduce error rates resulting from misalignment, but neither the
quantitative impact of misalignment nor the ability of filters to
minimize such error has been systematically studied. We performed
simulation experiments to quantify the level of false positive and
false negative for sitewise detection of positive selection due to
misalignment and the ability of alignment filters to minimize
alignment-induced error. We find that while misalignment causes few
false positives when analyzing mammalian-like datasets with an
effective aligner, analyses at vertebrate-like or greater divergences
with either a poor aligner or a small number of sequences results in
high false positive rates.

}

\keywords{Positive selection, multiple alignment, alignment filtering, likelihood.}

\email{greg@ebi.ac.uk}

\mbestyle

\section*{Introduction}

Large-scale comparative genomics has been widely employed to identify
shared and unique biological features between organisms of interest to
the biomedical sciences, with applications ranging from evolutionary
annotation transfer \citep{Bork1998Predicting} to phylogenetics
\citep{Murphy2007Using,Prasad2008Confirming} and population genetic
inference \citep{Hurst2009Genetics}. The decreasing cost of DNA
sequencing has triggered a striking increase in the number of model
and non-model organisms with planned genome sequencing projects,
suggesting that the range and scale of comparative genomics
applications will continue to expand
\citep{Green20072x,2007Identification}.

The study of protein evolutionary rates and selective pressures in
particular has flourished as a result of the growth in comparative
genomics datasets. Clusters of closely-related genome sequences across
a wide taxonomic range have led to a better understanding of which
aspects of molecular evolution are variable and which are constant
\citep{Wolf2009Nonlinear}, and an increased sampling of species should
continue to boost the power and accuracy of individual analyses within
a given clade. This is especially beneficial for the calculation of
spatially precise evolutionary estimates, as additional species
sampling has been shown to be an effective means of boosting the
accuracy and power of sitewise detection of positive selection and
evolutionary constraint
\citep{Anisimova2001Accuracy,Massingham2005Detecting}. Site-specific
evolutionary estimates have proved especially valuable when analyzed
in conjunction with other protein-based datasets such as protein
structural features \citep{Lin2007Proportion,Ramsey2011Relationship},
human population diversity \citep{2010Map}, and human disease
mutations \citep{Arbiza2006Selective}.

A major concern in the sitewise detection of protein positive
selection is that the effect of alignment error is not well
characterized. Intuitively, one might expect alignment error to result
mainly in an increased number of false positives, as the spurious
alignment of nonhomologous codons on average would result in a high
number of apparent non-synonymous substitutions and a low number of
synonymous substitutions (since a randomly chosen pair of codons are
more likely to be nonsynonymous than synonymous). However, false
negatives may also be introduced, either through over-alignment of
synonymous codons (increasing the synonymous substitution rate) or
non-alignment of homologous codons (causing reduced power due to less
evolutionary information). Since different aligners employ a variety
of algorithms, evolutionary models, and heuristic optimizations
\citep{Notredame2007Recent}, each program may be more or less prone to different
types of alignment error, causing potentially large variations in the
nature and magnitude of its impact on the detection of positive
selection.

The protein structure and the evolutionary divergence of a dataset may
also contribute to the effects of alignment error. Differently
structured protein regions show variable tolerance to biological
indels, with indels more common in extracellular and transmembrane
proteins than in highly folded enzymes and housekeeping genes
\citep{delaChaux2007DNA} . This suggests that well-folded protein
regions will experience fewer biological indels -- and will therefore
be less susceptible to alignment error -- than less structured
regions. Furthermore, the evolutionary divergence of a dataset affects
the power of sitewise inference and the prevalence of indels in
different ways. Because maximum-likelihood methods for detecting
positive selection require data in the form of observed substitution
events, they show little power at low divergence and their highest
power at intermediate to high divergence levels
\citep{Anisimova2001Accuracy}. On the other hand, the number of indels
in a dataset should increase monotonically with divergence. These two
trends suggest that the overall power will be low at both extremes of
divergence, with little inference power at low divergences (due to the
scarcity of data in the form of observed substitutions) and an
overwhelming amount of alignment error at high divergences (due to the
large number of indel events). It is important to note, however, that
the majority of genes in many biological clades of interest (such as
mammals, vertebrates, fruit flies, and yeast) fall within the middle
range of divergences where sitewise methods are at their most powerful
and where multiple alignment is a difficult---but not
hopeless---problem. As such, it is important to seek an understanding
of the impact of alignment error on overall error rates within this
important range of divergence levels.

A number of empirical analyses have established that errors in gene
sequencing, annotation and alignment can contribute to errors in
downstream evolutionary analyses such as phylogeny inference
\citep{Wong2008Alignment} et and estimates of positive selection
\citep{Schneider2009Estimates,Markova-Raina2011High} . Most recently,
Markova-Raina et al. \citeyearpar{Markova-Raina2011High} showed that
the detection of positively-selected sites and genes in Drosophila
genomes is highly sensitive to aligner choice, with PRANK's codon
model consistently producing alignments with the lowest amount of
positive selection. Still, according to the authors' manual inspection
of alignments, even positively-selected sites identified with PRANK
alignments contained a sizable proportion of apparent false positives.

A limitation to the analysis of error in empirical datasets is the
lack of a benchmark set of true alignments and positively-selected
sites. Markova-Raina et al. \citeyearpar{Markova-Raina2011High} used
the expected general effect of alignment error (an increase in false
positives due to misalignment of non-homologous codons) as a proxy by
which to compare different methods, allowing for the conclusion that
PRANK was the least error-prone aligner in their analysis. However,
the absolute number of false positives remained uncertain due to the
possible conflation of multiple sources of error: in addition to
alignment error, the authors noted that gene mis-annotation was
responsible for many apparent false positives, and there was also an
expected error rate from the likelihood inference method itself. This
limitation leaves important and interesting questions regarding the
nature of misalignment error and its quantitative impact on the
detection of positive selection left unanswered.

Controlled simulation experiments provide a natural framework for
investigating error rates in detail, allowing one to pinpoint the
source of errors in multi-step analyses such as alignment followed by
evolutionary inference. This approach has widely been employed in
assessing the robustness of phylogenetic inference methods to
misalignment
\citep{Dwivedi2009Phylogenetic,Ogden2006Multiple,Loytynoja2008PhylogenyAware}
but those results cannot be easily extrapolated to the analysis of
sitewise selective pressures. More recently, Fletcher and Yang
\citeyearpar{Fletcher2010Effect} performed a series of simulation
experiments investigating alignment error in the use of branch-site
models to detect positive selection in genes. Their results showed
that most aligners caused false positives by over-aligning codons, and
that datasets from mammalian and vertebrate gene families contain
enough evolutionary divergence to make errors resulting from
misalignment a legitimate concern.

Reflecting a widespread awareness of the problem of misalignment,
methods for identifying and removing uncertain or unreliable alignment
regions have been commonly used in phylogenetic and molecular
evolutionary analyses. The popular Gblocks applies a set of heuristic
criteria (including thresholds on the amount of conservation and
gappiness) to identify conserved blocks deemed suitable for
phylogenetic or evolutionary analysis \citep{Castresana2000Selection},
while aligners such as T-Coffee \citep{Notredame2000TCoffee} and PRANK
\citep{Loytynoja2005From} produce estimates of alignment confidence or
reliability. Most recently, alignment confidence scores based on the
robustness of alignments to guide tree perturbations
\citep{Penn2010Alignment} or the sampling of suboptimal alignments
\citep{Kim2011PSAR} have been proposed. Unfortunately, despite their
widespread use, the impact of the many available filtering methods on
phylogenetic and evolutionary analyses has not been well-studied. Even
for a single filtering program, Gblocks, results have been
contradictory: one simulation-based study found that it improved the
phylogenetic signal \citep{Talavera2007Improvement} while an empirical
study across a wide range of taxa found that Gblocks-filtered
alignments produced worse phylogenetic trees
\citep{Dessimoz2010Phylogenetic}. The effect of filtering alignments
on the detection of positive selection is largely unexplored, although
it has become accepted practice to apply published filtering methods
before testing for positive selection \citep{Studer2008Pervasive,Aguileta2009Rapidly}.

This paper aims to use a simulation framework to incorporate
misalignment and alignment filtering into estimates of the power and
accuracy of sitewise evolutionary estimates. Our approach is similar
to that of Anisimova and Yang \citeyearpar{Anisimova2002Accuracy} in
that we use simulated protein alignments to evaluate methods for
detecting sitewise positive selection, considering each site to be an
independent hypothesis test producing a positive or negative
result. Our experimental design also shares some features with
Fletcher and Yang's \citeyearpar{Fletcher2010Effect} investigation of
the performance of branch-site methods for detecting genewise positive
selection under alignment error. Like Fletcher and Yang
\citeyearpar{Fletcher2010Effect}, we simulate codon alignments with
indels and calculate inferred alignments using a small but diverse
sample of aligners. Unlike their study, however, we focus on sitewise
detection of positive selection occurring throughout a phylogeny (as
opposed to genewise detection of selection acting at specific
branches), we explore a wider range of plausible tree sizes, indel
rates and divergence levels, and we evaluate the effects of a number
of alignment filtering methods on the error rate and power of sitewise
analysis. Given the recent proliferation of completed and planned
vertebrate genomes, we pay special attention to choosing simulation
parameters and analysis methods similar to those commonly encountered
in the sitewise analysis of vertebrate gene families.

\section*{Methods}

\subsection*{Alignment Simulations}

Three rooted trees were used to guide the simulation of protein-coding
DNA alignments: the artificial 6-taxon tree used by Anisimova and Yang
\citeyearpar{Anisimova2001Accuracy} and Massingham and Goldman
\citeyearpar{Massingham2005Detecting} rooted at its midpoint, the
17-taxon vertebrate B-globin tree from Yang et
al. \citeyearpar{Yang2000CodonSubstitution}, and the 44-taxon
vertebrate tree used by the ENCODE project
\citep{2007Identification,Nikolaev2007Early}. Trees, shown with their
original branch lengths in Figure 1, were scaled to comparable
divergence levels by normalizing their root-to-tip mean path length
(MPL), defined as the root-to-tip branch length averaged across all
lineages in the tree. We simulated alignments with MPL divergence
between 0.2 and 2.0 synonymous substitutions per synonymous site,
spanning the range of evolutionary divergences observed in several
clades of organisms with fully-sequenced genomes (Table ST1).

\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.5]{fig1.pdf}
\end{center}
\caption{Phylogenetic trees used for simulation and analysis in this
  article. (a) An artificial tree used in previous simulations
  \citep{Anisimova2001Accuracy,Massingham2005Detecting}. (b) A tree
  estimated from b-globin genes of 17 vertebrates and used in previous
  empirical analyses and simulation studies
  \citep{Anisimova2001Accuracy,Anisimova2002Accuracy}. (c) The
  44-taxon tree used by the ENCODE project
  \citep{2007Identification,Nikolaev2007Early}. The nodes labeled in
  red were used as the reference species (see Methods).}
\label{fig_1}
\end{figure}

The Indelible program \citep{Fletcher2009INDELible} was used to
simulate codon sequences with indels along each phylogenetic tree. The
length of the root sequence was set to 500 codons, $\bm{\kappa}$ (the
ratio of transition to transversion substitutions) was fixed at 4, and
100 replicate alignments were simulated for each set of
conditions. Indel lengths were drawn from a discretized power-law
distribution with an exponential decay parameter of 1.8 and a maximum
value of 40, yielding a mean indel length of 3.33 codons and standard
deviation of 5.51 codons. This power-law model of indel lengths is
well-supported by empirical studies
\citep{Benner1993Empirical,Cartwright2009Problems} and it was decided
after manual inspection of alignments from a range of parameter values
that the chosen model parameters resulted in alignments most closely
resembling those encountered in vertebrate alignments. The ratio of
insertion to deletion events was set to 1, and the rate of indel
formation was scaled from between 0 to 0.2 indel events per
substitution per site.

The distribution of sitewise selective pressures was modeled with a
log-normal distribution derived from a maximum-likelihood fit to a
large dataset of sitewise selective pressures estimated from mammalian
gene trees (log-normal parameters shown in Table 1). This
distribution, with mean w of 0.28 and 6\% of sites having w>1, is
consistent with the structure-based expectation of many protein sites
under purifying selection and few under neutral selection or positive
selection
\citep{Smith1970Natural,Kimura1974SomePrinciples}. Indelible's general
discrete model of sitewise w variation was used to approximate this
distribution by splitting the lognormal probability density into 50
equally-spaced bins between w values of 0 and 3, with the highest bin
containing the probability density for all values w>=3.

Branch lengths for each of the simulation trees were multiplied by a
correction factor before simulation to correct for the difference
between our definition of branch lengths as the number of synonymous
substitutions per synonymous site ($dS$) and Indelible's
interpretation of branch length as the average number of substitutions
per codon ($t$) \citep{Fletcher2010Effect}. They are related
approximately by $t = 3(NdN + SdS) = 3dS(\omg{}N + S)$, where $N$ and
$S$ are the proportion of nonsynonymous and synonymous sites and
$\omg{}$ is the mean $\omg{}$ across all sites. S is approximately 0.3
when $\bm{\kappa}=4$ and the mean $\omg{}$ ratio for our chosen
distribution is 0.277, yielding a $dS$-to-$t$ conversion factor of
1.48 for all simulations performed.

\subsection*{Sequence Alignment and Filtering}

Alignments were inferred using four alignment algorithms chosen for
their widespread use or demonstrated accuracy: ClustalW v1.82
\citep{Thompson1994ClustalW}, MAFFT \citep{Katoh2005MAFFT} and two
variants of PRANK \citep{Loytynoja2005From} based on an amino acid
model (subsequently referred to as \pranka{}) or an empirical codon
model (subsequently referred to as \prankc{}). Unaligned amino acid
sequences were given as input to all alignment programs (except
\prankc{}, which was provided the unaligned DNA sequences) and all
software was run using default parameters with the true phylogenetic
tree given as input where possible.

Alignments were filtered by masking out residues based on the output
of four alignment scoring methods: Gblocks conserved blocks
\citep{Castresana2000Selection}, T-Coffee consistency scores
\citep{Notredame2000TCoffee,Notedame2003Using}, and GUIDANCE alignment
confidence scores \citep{Penn2010Alignment}. Gblocks, which identifies
entire alignment columns as conserved or not conserved, was run using
a relaxed gap allowance parameter in order to avoid removing too much
of the alignment (command-line parameter 'b5=a') and all residues from
any columns not within an identified conserved block were masked with
'N's. GUIDANCE and T-Coffee, which produce individual scores for each
residue, were handled differently. (TODO!!!!). GUIDANCE was run with 30
bootstrap replicates and MAFFT as the bootstrap aligner (except when
\pranka{} or \prankc{} was the original aligner, in which case \pranka{}
was used to align the bootstrap replicates), and T-Coffee was run
using default settings and the 'evaluate\_mode -output=score\_ascii'
command-line parameter to output alignment scores.

Two unrealistic datasets were produced to serve as controls. First,
the true simulated alignment was used in order to evaluate the
sitewise performance without any alignment error. Second, an
additional filtering method was designed to represent an unattainable
best-case scenario for sequence filtering, using knowledge of the true
alignment to assign a score to each residue that summarizes how
correctly is has been placed in the inferred alignment. The approach
taken was to calculate, for each residue, the branch length of the
correct sub-tree (defined as the minimum sub-tree connecting all
sequences to which the current residue is correctly aligned) divided
by the branch length of the aligned sub-tree (defined as the minimum
sub-tree connecting all sequences with non-gap residues at the current
alignment column). This residue-wise score ranges from 0 to 1 and
encapsulates the expectation that correctly-aligned evolutionary
branch length is the main source of information from which the
sitewise inference methods derive their power. Scores from this
method, referred to as the Optimal method, were used to filter
inferred alignments in a manner similar to the T-Coffee and GUIDANCE
scores.

\subsection*{Sitewise Evolutionary Analysis}

Sitewise estimates of selective pressures were calculated using
maximum-likelihood methods implemented in the Phylogenetic Analysis by
Maximum Likelihood (PAML, \citealt{Yang2007PAML}) and Sitewise Likelihood Ratio
(SLR, \citealt{Massingham2005Detecting}) software packages.

The codeml program from PAML implements a number of likelihood ratio
tests (LRTs) for detecting the presence of positive selection in a
gene while allowing the $\omg{}$ ratio to vary among sites
\citep{Yang2000CodonSubstitution}. These models, known as the sites or
random sites models, use a variety of predefined statistical
distributions to account for heterogeneous w ratios among sites. After
the likelihood optimization is performed, Bayesian methods can be used
to estimate the posterior probability of each site being drawn from a
given site class, where a high posterior probability of a site
belonging to a class with $\omg{}>1$ can be considered strong
evidence that a site has evolved under positive selection
\citep{Yang2005Bayes}. We used the two models for which the
recommended Bayes Empirical Bayes method are implemented, M2a and M8.

SLR implements a method specifically designed for sitewise estimates
which has been shown in simulations to perform as well or better than
PAML's sitewise models \citep{Massingham2005Detecting}. SLR models
codon evolution as a continuous-time Markov process where
substitutions at one site are independent of substitutions at all
other sites. Importantly, no assumptions are made regarding the
distribution of $\omg{}$ ratios within the alignment. The value
of $\omg{}$ is considered to be an independent parameter at each
site: after first optimizing shared parameters using the whole
alignment, SLR uses the shared parameters and the data at each
alignment site to calculate a sitewise statistic for non-neutral
evolution. This statistic is based on a likelihood-ratio test where
the null model is neutral evolution ($\omg{}=1$) and the
alternative model is either purifying or positive selection
($\omg{}<1$ or $\omg{}>1$, respectively). The raw statistic
measures the strength of evidence for non-neutral evolution at each
site; following Massingham \citeyearpar{Massingham2005Detecting} we
use a signed version of the SLR statistic (created by negating the
statistic for sites with $\omg{}<1$) as the test statistic for
positive selection.

\subsection*{Measuring Performance}

In order to compare sitewise estimates from different alignments, a
single sequence from each tree was chosen as the reference (red
labels, Figure 1) and all sitewise statistics were mapped from
alignment columns to sequence positions in the reference
sequence. This approach bears some resemblance to the process of
mapping alignment-based evolutionary inferences onto a single member
of the alignment for further analysis and integration with other
genome-referenced data (as is often done, for example, using mammalian
alignments and a human reference). As a result of this reference
sequence based mapping, sites which were deleted in the reference
sequence or inserted in a lineage not ancestral to the reference were
dropped from the analysis.

To evaluate the power and error rates that might be achieved in
real-world data analysis, the recommended cutoff thresholds for PAML’s
Bayesian posterior probabilities and the SLR statistic were used to
identify positively selected sites. A posterior probability threshold
of 0.95 was used for PAML \citep{Yang2005Bayes} and a threshold of 3.84,
the 95\% critical value of the chi-square distribution with 1 degree
of freedom, was used for SLR \citep{Massingham2005Detecting}. Sites were
compared to their true simulated state (e.g. positively-selected or
non-positively selected) in order to identify correct and incorrect
inferences, and from these classifications we calculated the false
positive rate (FPR, defined as the proportion of all sites with true
$\omg{}<1$ falsely identified as positively selected) and true positive rate
(TPR, defined as the proportion of all sites with true $\omg{}>1$ correctly
identified as positively selected).

As the addition of alignment error is expected to affect the power and
error rates differently for each combination of simulation condition
and aligner, we also identified the score thresholds for each dataset
that resulted in an actual FPR of 1\% and calculated the TPR achieved
at this actual error rate (referred to as \tpr{} to distinguish it
from the TPR described above). Although this estimate of
error-controlled power would be impossible to calculate in an
empirical analysis where the error rate is unknown, it is useful in a
simulation context for allowing a controlled comparison of the
performance of sitewise analysis between different
conditions. Specifically, it should be sensitive both to increased
error rates from alignment error and to reduced power from alignment
filtering, in both cases showing a lowered error-controlled power as
the method performs worse.

\section*{Results and Discussion}

\subsection*{Performance of Three Methods for the Sitewise Detection of Positive Selection}

\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.75]{fig2.pdf}
\end{center}
\caption{The behavior of three methods of sitewise inference in the
  absence of indels. One hundred replicate alignments were simulated
  without indels for each tree and divergence level and were analyzed
  with either PAML M2 (red), PAML M8 (blue), or SLR. The performance
  of each method, as measured by four summary statistics, is plotted
  as a function of divergence level. From top to bottom: TPR at the
  recommended cutoff threshold (0.95 for PAML and 2.71 for SLR), FPR
  at the recommended cutoff threshold, TPR at a 1\% FPR threshold, and
  Pearson's correlation coefficient between the true and estimated
  $\omg{}$.}
\label{fig_2}
\end{figure}

We first evaluated the ability of three sitewise methods, PAML M2,
PAML M8 and SLR, to accurately estimate sitewise $\omg{}$ values and to
detect positive selection under a range of tree lengths in the absence
of alignment error. Figure 2 shows the TPR, FPR, \tpr{} and sitewise $\omg{}$
correlation over a range of MPLs for each of the three simulation
trees.

The detection power and ω correlation were weakest at low divergence
levels for all methods and all trees due to the low amount of
evolutionary information, as observed in previous simulations
\citep{Anisimova2002Accuracy}. Also confirming previous results, we
found a positive correlation between tree size and detection power,
with the highest values in the 44-taxon tree. Power generally
increased monotonically with divergence, except for the 6-taxon tree
which saw its maximum performance at moderate divergence levels (MPL
0.5-1.0) and began decreasing at higher values. The downward trend in
the 6-taxon tree was likely due to the impact of saturation of
synonymous sites in the very long branches present in such a sparse
tree at high divergence levels. The two larger trees, with lower
average branch lengths at equivalent divergence levels, showed less
susceptibility to saturation of substitutions along a single
branch. Interestingly, the two larger trees showed no signs of
decreased performance even at a MPL of 2 substitutions per site, which
is greater than any of the divergence levels found in groups of
commonly analyzed vertebrate, insect and fungal species (Table ST1).

Comparing between the three sitewise methods, we found that at the
recommended cutoff threshold (Figure 2, top row) SLR showed the
highest power to detect positive selection in all trees, followed by
PAML M8 and PAML M2. In the smallest tree, the power of the two PAML
methods was virtually zero while SLR reached a maximum of ~18\% TPR at
MPL=0.5. SLR showed a large increase in power in the larger trees
compared to the smallest tree, yielding ~25\% and ~45\% TPR at MPL=0.5
in the 17-taxon and 44-taxon trees, respectively. The two PAML methods
also showed increased power in the larger trees, with PAML M2 ranging
between 50-75\% of SLR’s power and the performance of PAML M8 falling
in between the two other methods.

The TPR measurements represent the power that might be achieved in
real-world analysis using recommended cutoff thresholds, but SLR’s
higher power may reflect merely a shifted balance between power and
accuracy at the recommended cutoff threshold as opposed to an
increased ability to discriminate positive from neutral or purifying
selection. The FPR and error-controlled \tpr{} results showed the
former to be the case. The FPR from SLR was higher than that from
either of the PAML methods for all trees and divergence levels (Figure
2b), suggesting that its higher power was the result of a
less-conservative cutoff rather than significantly improved
statistical performance. This hypothesis was confirmed by evaluating
the TPR at a cutoff threshold for each method that controlled for an
actual FPR of 1\% (\tpr{}, shown in Figure 2c), yielding virtually
identical \tpr{} values for all three methods. The highly similar
error-controlled power levels indicate that the three methods’
sitewise statistics were nearly equally sensitive to positive
selection under these simulation conditions.

The conservative nature of the default thresholds for PAML and SLR
have been previously noted
\citep{Anisimova2002Accuracy,Yang2005Bayes,Massingham2005Detecting}
and the extremely low false positive rates in our simulations
confirmed that all three methods would yield very few false positives
with a realistic distribution of $\omg{}$ values across a range of
tree sizes and divergence levels. For the purposes of our indel
experiments, the observed similarity in error-controlled power levels
indicated that the behavior of PAML M2, PAML M8, and SLR was similar
enough not to warrant separately evaluating all three methods in the
subsequent indel simulation experiments. As the runtime for SLR was
significantly lower than that of either PAML model, all subsequent
results are presented only based on the SLR test.

\subsection*{Effect of Alignment Error on Sitewise Power}

When the indel rate was greater than zero, performance levels varied
significantly for different tree sizes, alignment algorithms, and
evolutionary divergences. Figure 3 shows the same sitewise
measurements as Figure 2 for simulations without indels and with
indels, analyzed using three different aligners and the true
alignment. The indel rate was held constant at 0.1 indel event per
substitution, and the divergence level was varied between 0.2 and 2.0
substitutions per site.

\begin{figure}[t]
\begin{center}
\includegraphics[scale=0.75]{fig3.pdf}
\end{center}
\caption{The behavior of sitewise inference in the presence and absence of indels using three different aligners and the true alignment.}
\label{fig_3}
\end{figure}

Comparing the results without indels to those with indels under the
true alignment (shown as solid black and solid red lines in Figure 3)
we found a slight decrease in power and ω correlation and no
noticeable increase in FPR. The decreased power was expected, since
even in the absence of alignment error, alignment columns containing
gaps harbor less evolutionary information than columns flush with
sequence data. The lack of increased FPR showed that SLR retained its
conservative statistical performance even when analyzing gapped
alignments with indels. Surprisingly, at higher divergences (MPL $>$
1.0) under the six-taxon tree, the FPR with indels was lower than the
FPR without indels. This unexpected result may be attributed to the
large number of alignment columns under such conditions that contain
only a single non-gap sequence, as these columns are never inferred as
positively-selected by SLR due to the complete lack of
information. The two larger trees did not show a similar trend at high
divergence levels, suggesting that the effect was due mainly to the
highly sparse nature of the alignment in the 6-taxon tree. The rest of
our results from the 6-taxon tree at high divergences were similarly
anomalous in this respect; we surmised that the sparseness of the true
alignment combined with the extreme difficulty of accurately aligning
sequences along very long branches made sitewise analysis with indels
very unreliable at high divergences in the smallest tree.

When alignments were inferred using one of the three aligners tested,
the TPR, \tpr{} and $\omg{}$ correlation were all reduced relative to
the true alignment (dashed and dotted red lines, Figure 3A, B, D). The
degree of reduction varied depending on the aligner, simulation
conditions, and performance measurement being analyzed. At low
divergences (e.g. MPL $<$ 0.2), the inferred alignments generally
showed only a small decrease in performance, but the \tpr{} in the
6-taxon tree was a notable exception with large differences down to a
MPL of 0.1. As divergence levels increased, so did the difference
between the performance of the true alignment and the inferred
alignments. The three aligners tested could be consistently and
unambiguously ranked by each of the measured performance
characteristics, with \prankc{} always performing best and ClustalW
performing worst. The same ranking of aligners with respect to
detecting positive selection was found by Fletcher and Yang
\citeyearpar{Fletcher2010Effect}; our results corroborate their
finding and provide evidence that this ranking may be applicable to a
wider array of evolutionary analyses than the branch-site tests
evaluated in their analysis.

When alignments were inferred using one of the three aligners tested,
the TPR, \tpr{} and $\omg{}$ correlation were all reduced relative to
the true alignment (dashed and dotted red lines, Figure 3A, B, D). The
degree of reduction varied depending on the aligner, simulation
conditions, and performance measurement being analyzed. At low
divergences (e.g. MPL < 0.2), the inferred alignments generally showed
only a small decrease in performance, but the \tpr{} in the 6-taxon
tree was a notable exception with large differences down to a MPL of
0.1. As divergence levels increased, so did the difference between the
performance of the true alignment and the inferred alignments. The
three aligners tested could be consistently and unambiguously ranked
by each of the measured performance characteristics, with \prankc{}
always performing best and ClustalW performing worst. The same ranking
of aligners with respect to detecting positive selection was found by
Fletcher and Yang \citeyearpar{Fletcher2010Effect}; our results
corroborate their finding and provide evidence that this ranking may
be applicable to a wider array of evolutionary analyses than the
branch-site tests evaluated in their analysis.

We found that FPR levels under inferred alignments were generally
higher than under the true alignment. This effect was especially
strong in the 6-taxon tree at medium divergence levels (e.g. MPL
0.2-0.6), where ClustalW showed up to a fourfold increase, and \prankc{}
a nearly twofold increase, in FPR over the true alignment. As
previously noted, the 6-taxon tree showed an unexpected and presumably
anomalous FPR pattern at higher divergences, likely due to the highly
sparse true alignment. In the two larger trees, FPRs from inferred
alignments were less elevated compared to the true alignment and
relatively constant across the range of divergences. ClustalW's FPR
ranged between 0.003 to 0.005, while \prankc{}'s FPR was virtually
identical to the true alignment in both the 17-taxon and 44-taxon
trees.

A comparison of the TPR and FPR results between true and inferred
alignments allows the relative impact of false positive and false
negative errors resulting from misalignment to be quantified. A false
positive resulting from over-alignment of non-synonymous codons would
contribute to an elevated FPR, while a false negative resulting from
non-alignment of homologous codons would contribute to a reduced
TPR. The \tpr{} should be sensitive to both false positives and false
negatives since it measures the TPR when controlling for an actual FPR
of 1\%. Analyzing the TPR results in the 6-taxon tree, the three
aligners formed a cluster of lines well below the true alignment
value, indicating similar tendencies among the different aligners to
produce false negatives in the smaller tree. In larger trees the
different aligners showed a wider spread of TPR values, but even
\prankc{} showed a 5-10\% reduction compared to the true alignment at
MPL$=$1.0. These results show that the introduction of false negatives
is a significant and seemingly unavoidable result of alignment error
at medium to high divergence levels (MPL $>$ 0.5), with even the most
powerful aligner producing a marked reduction in TPR compared to the
true alignment. The FPR levels exhibited a very different trend, with
the widest range of values and highest FPRs in the 6-taxon tree and
the smallest range and lowest values in the 44-taxon
tree. Furthermore, the aligner with the lowest FPR, \prankc{}, had FPR
levels almost identical to those of the true alignment in the 17-taxon
and 44-taxon trees. The FPR results suggest that although false
positives resulting from misalignment are unavoidable in a
moderately-diverged 6-taxon tree, the addition of more taxa to the
tree provided enough information for \prankc{} to yield very few false
positives not in the true alignment. Other aligners, however,
maintained a constant elevated FPR in the two larger trees, indicating
a failure to effectively utilize the additional evolutionary
information present in larger trees to avoid introducing false
positive results.

The results shown in Figure 3 allow us to characterize the combined
effects of alignment error at different commonly analyzed levels of
divergence. For example, at a human-mouse divergence level (MPL $=$
0.2) misalignment had little impact on the TPR regardless of what
aligner was used. However, ClustalW yielded a notably higher FPR than
MAFFT or \prankc{}, and the error-controlled \tpr{} was
correspondingly lower for ClustalW in all three trees. Thus, at small
divergences we found that false positives were thus the main source of
error from misalignment, and different aligners had highly variable
tendencies to produce false positive results. At the higher vertebrate
and Drosophila divergence levels (MPL 0.8-1.0) false negatives became
much more prevalent. The TPR for all inferred alignments was virtually
zero in the 6-taxon tree, underscoring the necessity of including many
species in the analysis of highly diverged sequences. In the larger
trees, \prankc{} resulted in very few additional false positives, but
it suffered a 5-10\% reduction in TPR relative to the true alignment
due to false negatives. Meanwhile, ClustalW showed a 50\% TPR
reduction and maintained a strongly elevated FPR. At higher
divergences and in larger trees, false negatives appear to be the most
persistent alignment error, causing a marked reduction in sitewise
power even with the best-performing aligner.

\begin{figure*}[t]
\begin{center}
\includegraphics[scale=0.8]{fig4.pdf}
\end{center}
\caption{}
\label{fig_4}
\end{figure*}

\subsection*{Sitewise Power Under a Range of Indel Rates and Divergences}

To explore the effects of alignment error across a wider range of
simulation conditions, we extended the simulations from Figure 3
across a range of indel rates, from 0 to 0.2 indel events per
substitution. Figure 4 shows heatmaps of the TPR and FPR for ClustalW,
\prankc{} and the true alignment (Figure 4A,B) and a heatmap of the
error-controlled \tpr{} for all aligners tested (Figure 4C). The
results from Figure 3, which were simulated with an indel rate of 0.1,
correspond to the middle row of each panel in Figure 4; rows above and
below the middle row represent higher and lower indel rates,
respectively.

The TPR values (Figure 4A) showed a consistent pattern across the
range of indel rates, with power decreasing as either the indel rate
or the divergence level increased. \prankc{} showed a greater ability
than ClustalW to maintain a high TPR at higher indel rates, especially
in the 17-taxon and 44-taxon trees. At lower indel rates, the TPR
performance of both aligners and the true alignment were qualitatively
similar.

\prankc{} and ClustalW both showed a similar tendency towards elevated
FPRs in the 6-taxon tree (Figure 4B), but their behavior diverged
significantly in the 17-taxon and 44-taxon trees. In the 17-taxon
tree, \prankc{} only showed an elevated FPR at very high indel rates
and divergence levels, but the ClustalW FPR consistently increased
with indel rate, quadrupling in value from the lowest to highest indel
rate. Interestingly, for any given indel rate, the ClustalW FPR showed
little variation across the range of divergence levels. This result
was counter-intuitive, as we expected alignment errors to become more
common as divergence increases and the number of observed indel events
grows. Furthermore, \prankc{} behaved according to our expectations,
with a trend towards increased FPRs at higher divergences in the
17-taxon tree and the highest indel rate. The FPR results in the
44-taxon tree confirmed the strange effect of ClustalW’s alignments on
the sitewise FPR: at the highest indel rates, ClustalW showed a
negative correlation between FPR and divergence -- exactly opposite to
the trend we expected.

Finally, the error-controlled \tpr{} results (Figure 4C) provided a
comprehensive map of the effect of alignment error on detecting
sitewise positive selection. The two aligners not shown in the two
other panels, MAFFT and \pranka{}, exhibited \tpr{} values
intermediate to those from ClustalW and \prankc{} across the range of
parameters tested. At indel rates above 0.1, all aligners yielded low
\tpr{} values in the 6-taxon tree. In the larger two trees, excessive
false positives introduced by ClustalW and MAFFT alignments precluded
them from achieving high \tpr{} at high indel rates, while \pranka{}
and \prankc{} showed good \tpr{} levels at all but the highest
divergences and indel rates.

It is worth noting the exceptional ability of PRANK to maintain a low
level of false positive sites even under extremely difficult alignment
conditions. Although \prankc{} showed slightly elevated FPRs at high
indel rates in the 17-taxon tree, FPRs were nearly identical to the
true alignment across all simulated conditions in the 44-taxon
tree. This impressive performance suggests that, given a large enough
number of taxa, PRANK alignments would yield very few erroneous false
positives in scans for positive selection in sequences with divergence
levels well beyond those found in commonly analyzed vertebrate, insect
and fungal clades (Table ST1). Furthermore, these results showed that
false negatives contributed more than false positives to PRANK’s
reduction in sitewise performance --- a novel observation which
provides insight into the nature of PRANK alignments and their
application to sitewise evolutionary analysis.

\subsection*{Effect of Alignment Filtering on Sitewise Error Rates}

\begin{figure*}[t]
\begin{center}
\includegraphics[scale=0.8]{fig5.pdf}
\end{center}
\caption{}
\label{fig_5}
\end{figure*}

Having established that alignment error can lead to reduced sitewise
performance through the introduction of false negatives and false
positives we tested whether a number of alignment filtering methods
could reduce error rates and improve the power of sitewise detection
of positive selection. Using sequences simulated from the 17-taxon
tree and the same range of indel rates and divergence levels as in the
previous experiments, we calculated inferred alignments using ClustalW
and \prankc{} and applied four filtering methods to the alignments
before performing the sitewise analysis. Since we wished to determine
whether alignment filters either improved or worsened the error rate
and power of sitewise analysis, we measured the ratio of each
performance measure to the value obtained in the equivalent unfiltered
dataset. These relative scores are presented in Figure 5.

We first looked at the two controls included in this set of
experiments, the unfiltered true alignment and the inferred alignments
filtered with the optimal filter (Figure 5, top two rows). The true
alignment always showed reduced FPR and increased TPR, and \tpr{}
compared to the inferred alignments, generally with a with greater
magnitude of change compared to the ClustalW alignments than to the
\prankc{} alignments. These relative scores represented the direction
and magnitude of change that might be achieved by an effective
alignment filter. Our hypothesis was that the optimal filtering method
would show the same direction of change in FPR and TPR as the true
alignment. When applied to ClustalW alignments (Figure 5a), the
optimal filter indeed showed a clear pattern of improved sitewise
performance across the majority of simulation conditions, with reduced
FPR and elevated TPR and \tpr{}. At the highest indel rates and
divergence levels, however, little change was observed. This may have
been due to the combination of our filtering limit of 20\% masked
residues and the highly inaccurate nature of ClustalW alignments under
those conditions, as evidenced by the low TPRs and high FPRs in the
previous set of simulations (Figure 4A,B). When applied to \prankc{}
alignments, the optimal filter showed a similar reduction of FPR
levels across the parameter space, but the TPR and \tpr{} results were
mixed. At lower indel rates and divergence levels the TPR and \tpr{}
were reduced, while at higher indel rates and divergences both the TPR
and \tpr{} were slightly increased. The beneficial impact on power
levels under the more difficult alignment conditions sugested that the
optimal filter was able to successfully remove misaligned residues
that were contributing to false negatives in the \prankc{}
alignments. We suspect that the worse performance under easier
alignment conditions was due to the small number of false positives
and false negatives in those \prankc{} alignments, causing any amount
of filtering to result in reduced performance.

When interpreting the results from filtered alignments, we must keep
in mind that alignment filters act through removing alignment residues
or columns, resulting in a certain amount of reduction in the FPR and
TPR purely from the removal of information from the alignment. For
example, a filter that randomly removed a fraction of residues of each
alignment would be expected to produce equal reductions in FPR and TPR
levels. An effective filter that removes false positives may also
yield a reduced TPR, but the FPR reduction would be larger in
magnitude, making the detection of positive selection more powerful at
a given error rate. Thus, a reduced FPR is not necessarily indicative
of good filtering performance, nor is a reduced TPR necessarily
indicative of poor filtering performance. Furthermore, the prevalence
of false negatives resulting from misalignment suggests that alignment
filters may also improve power by removing false negatives, perhaps by
masking out misaligned residues that were preventing positive sites
from being identified. The removal of false negatives would result in
an increased TPR, further complicating the assessment of filtering
results based on FPR or TPR alone. As a result, we preferred the use
of the change in error-controlled \tpr{} when evaluating whether a
filter had successfully improved the sitewise power of a dataset,
since this value is sensitive to changes in both the FPR and TPR.

Using the change in \tpr{} as a measure of filtering performance, we
found T-Coffee and GUIDANCE to be highly effective at improving
ClustalW alignments, with magnitudes of improvement near to those from
the optimal filter. Interestingly, GUIDANCE showed increased
performance at the highest indel and divergence levels, where the
optimal filter had no effect. As with the optimal filter, results were
mixed when applying these two filters to the \prankc{} alignments: at
lower indel rates and divergences both filters resulted in reduced
performance, while T-Coffee achieved modest improvements while
GUIDANCE showed no change under more difficult conditions.

Gblocks behaved very differently from the other filters tested,
resulting in reduced FPR, TPR, and \tpr{} values under nearly all
simulation conditions. Only at high indel rates and low divergence
levels in the ClustalW alignments did Gblocks show an increased \tpr{}
relative to the unfiltered alignments. The poor performance of Gblocks
was likely due to overly-aggressive removal of alignment
columns. Since the program only identifies conserved blocks as opposed
to scoring residues or alignment columns, we could not cap the amount
of masked sequence as was done for the residue-wise filters. Even
though a relaxed setting was used which allowed conserved blocks to
contain gapped regions (see Methods), the dramatically reduced TPR
levels indicate that much of the power to detect positive selection
was eliminated by Gblocks filtering. Dessimoz et
al. \citeyearpar{Dessimoz2010Phylogenetic} found a similar result for
the effect of Gblocks on the accuracy of phylogenetic inference,
suggesting that Gblocks filtering tends to reduce, rather than
increase, the power and accuracy of a number of evolutionary analyses.

Overall, our alignment filtering simulations found that filtering
methods based on GUIDANCE and T-Coffee scores showed a good ability to
remove false positives and false negatives from ClustalW alignments,
with levels of improvement on par with those obtained with the optimal
filter. This was the case across a wide range of divergence levels and
indel rates, suggesting that filtering ClustalW alignments with
T-Coffee or GUIDANCE can usually be expected to result in improved
sitewise performance. Alignment filtering proved less helpful when
applied to \prankc{} alignments, with all filters reducing the
sitewise performance except under the most divergent and indel-rich
conditions. Importantly, alignment filtering only showed improved
performance at divergence levels well above those found in
commonly-analyzed groups of species (e.g., MPL $>$ 1.6). This leads us
to the conclusion that, based on the sample of aligners and filters
tested here, the unfiltered \prankc{} alignments would yield the best
performance for detecting sitewise positive selection when analyzing
protein-coding sequences across a wide range of indel rates and
commonly encountered sequence divergence levels.

\section*{Conclusions}

Sequence alignment is important yet under-studied]

Our simulations showed that alignment error has significant impact on sitewise detection of pos-sel. ClustalW worst, \prankc{} best.

Filtering improves low-quality alignments and high-quality alignments in extreme circumstances, but overall unfiltered \prankc{} is still the best.

Despite high FPRs from ClustalW, false positive rates are still very low... due mainly to conservative cutoff and prevalence of purifying selection in our simulation conditions

(fig.~\ref{sample_fig_2}).

\section*{Acknowledgments}

A.B.H. and P.V. acknowledge ``\'Ecole th\'ematique
g\'enomique environnementale'' for helpful discussions. M. Bormans is
acknowledged for comments on the manuscript. P.V. acknowledges GIS
``G\'enomique marine'' and ``Fondation Total'' and CNRS--ECCO for funding.

\bibliographystyle{mbe3}
\fontsize{9}{10}\selectfont%
\itemsep 0pt

\bibliography{slrsim}
\end{document}

### 2011-06-29 projects/slrsim/slrsim.functions.R
Slrsim updates from sanger.

@@ -19,6 +19,9 @@ paper.table <- function(df, adjust.pvals=F) {
   print("  stats")
   stats.05 <- df.stats(df, slr_thresh=3.84, paml_thresh=0.95, adjust.pvals=adjust.pvals)
 
+  print("  filtered fraction")
+  df.ff <- df.filter.fraction(df)
+
   rrow <- roc[1,]
   row <- df[1,]
 
@@ -49,11 +52,9 @@ paper.table <- function(df, adjust.pvals=F) {
     `cor`   = cor,
     `cor_pearson` = cor_pearson,
     `n_sites` = nrow(roc),
-    `n_genes` = n.genes
+    `n_genes` = n.genes,
+    `filtered_fraction` = df.ff$filtered_fraction
   )
-
-  #aln.acc.df <- df.alignment.accuracy(df)
-  #ret.df <- cbind(ret.df, aln.acc.df)
 
   print(ret.df[, c('tpr_at_fpr', 'tpr_at_thresh', 'fpr_at_thresh', 'cor')])
   return(ret.df)
@@ -324,6 +325,23 @@ df.gene.detection.rate = function(df) {
 
 df.cor = function(df) {
   return(list(cor=cor(df$true_dnds,df$aln_dnds,method='spearman',use='complete.obs')))
+}
+
+df.filter.fraction = function(df) {
+  first.r <- function(x,field) {
+    return(x[1, field])
+  }                      
+  mean.f <- function(field) {
+    return(mean(by(df, df$node_id, first.r, field)))
+  }
+  sum.f <- function(field) {
+    return(sum(by(df, df$node_id, first.r, field)))
+  }
+
+  out.df <- data.frame(
+    filtered_fraction=1 - (sum.f('residue_count') / sum.f('unmasked_residue_count'))
+  )
+  return(out.df)
 }
 
 df.alignment.accuracy = function(df) {

### 2011-06-29 projects/slrsim/slrsim_create_analyses.pl

@@ -65,16 +65,13 @@ my $h = new Bio::Greg::Hive::ComparaHiveLoaderUtils;
 $h->init($url);
 
 print "  cleaning tables...\n";
-$h->init_compara_tables;
 
 $h->clean_hive_tables;
-$h->clean_compara_analysis_tables;
-$h->clean_compara_tree_tables;
-my @truncate_tables = qw^
-      aln aln_scores omega
-      sites genes merged slrsim_results
+my @drop_tables = qw^
+      genes sites
+      slrsim_results slrsim_results_alt
   ^;
-$h->truncate_tables(\@truncate_tables);
+$h->drop_tables(\@drop_tables);
 
 load_trees();
 load_tree();
@@ -94,13 +91,17 @@ sub load_trees {
   my $params      = { experiment_name => $experiment_name };
   $h->create_analysis( $logic_name, $module, $params, 1, 1 );
 
-  $h->add_job_to_analysis( "LoadTrees", {} );    # Add a dummy job to run and load the trees.
+  $h->add_job_to_analysis( "LoadTrees", {
+    scratch_dir => 'scratch101'
+                           } );    # Add a dummy job to run and load the trees.
 }
 
 sub load_tree {
   my $logic_name  = "LoadTree";
   my $module      = "Bio::Greg::Slrsim::LoadTree";
-  my $params      = {};
+  my $params      = {
+    scratch_dir => 'scratch101'
+  };
   $h->create_analysis( $logic_name, $module, $params, 50, 1 );
 }
 

### 2011-07-02 projects/slrsim/paper/slrsim.tex
Paper draft.

@@
 \documentclass{mbe}
-\usepackage{natbib}
+\usepackage{natbib,graphicx,dblfloatfix,multirow,array,threeparttable}
+\usepackage{xspace}
 \setlength{\bibsep}{0.0pt}
@@
 
-\newcommand{\pranka}{PRANK\_A}
-\newcommand{\prankc}{PRANK\_C}
-\newcommand{\omg}{\bm{\omega}}
-\newcommand{\tpr}{TPR$_{1\%}$}
+\newcommand{\pranka}{PRANK$_{\textrm{AA}}$\xspace}
+\newcommand{\prankc}{PRANK$_{\textrm{C}}$\xspace}
+\newcommand{\omg}{\bm{\omega}\xspace}
+\newcommand{\tpr}{TPR$_{1\%}$\xspace}
+\newcommand{\sw}{sitewise\xspace}
+\newcommand{\Sw}{Sitewise\xspace}
+\newcommand{\Dr}{{\em Drosophila}\xspace}
 
-\title{The effects of alignment error and alignment filtering on the sitewise detection of positive selection}
+\title{The effects of alignment error and alignment filtering on the \sw detection of positive selection}
 
@@
 \authorlist{Jordan and Goldman}
-\shorttitle{Alignment eror and detecting sitewise positive selection}
+\shorttitle{Effect of Alignment Error on Detecting Sitewise Selection}
 
@@
 are not well understood. Alignment filters are commonly applied to try
-to reduce error rates resulting from misalignment, but neither the
-quantitative impact of misalignment nor the ability of filters to
-minimize such error has been systematically studied. We performed
-simulation experiments to quantify the level of false positive and
-false negative for sitewise detection of positive selection due to
-misalignment and the ability of alignment filters to minimize
-alignment-induced error. We find that while misalignment causes few
-false positives when analyzing mammalian-like datasets with an
-effective aligner, analyses at vertebrate-like or greater divergences
-with either a poor aligner or a small number of sequences results in
-high false positive rates.
+to reduce errors resulting from misalignment, but neither the
+prevalence of such errors nor the ability of filters to reduce them
+has been systematically studied. Focusing on the sitewise detection of
+positive selection, we performed simulation experiments to quantify
+the false positives and false negatives introduced by alignment error
+and the ability of alignment filters to improve performance. We found
+that some aligners produced many false positive results across a wide
+range of divergence levels, while others produced very few. False
+negatives were a problem for all aligners, increasing with
+divergence. Of the aligners tested, PRANK consistently performed the
+best and ClustalW performed the worst. The best alignment filters
+showed good ability to reduce the error rates from ClustalW
+alignments, but they failed to improve the performance of PRANK
+alignments under most circumstances. Of the filters tested, GUIDANCE
+performed the best and Gblocks performed the worst, failing to improve
+even the error-prone ClustalW alignments. The performance of the true
+alignment and an optimal filtering method suggested that performance
+improvements could be gained by improving aligners or filters to
+reduce the prevalence of false negatives, especially at higher
+divergence levels and indel rates. We conclude that for the sitewise
+evolutionary analysis of most vertebrate, insect and fungal
+phylogenies, the use of unfiltered PRANK alignments is recommended.
 
@@
 
-\keywords{Positive selection, multiple alignment, alignment filtering, likelihood.}
+\keywords{Alignment, positive selection, alignment filtering, indels}
 
@@
 sampling has been shown to be an effective means of boosting the
-accuracy and power of sitewise detection of positive selection and
+accuracy and power of \sw detection of positive selection and
 evolutionary constraint
@@
 
-A major concern in the sitewise detection of protein positive
+A major concern in the \sw detection of protein positive
 selection is that the effect of alignment error is not well
@@
 
+\input{table1.txt}
+
 The protein structure and the evolutionary divergence of a dataset may
@@
 \citep{delaChaux2007DNA} . This suggests that well-folded protein
-regions will experience fewer biological indels -- and will therefore
-be less susceptible to alignment error -- than less structured
+regions will experience fewer biological indels---and will therefore
+be less susceptible to alignment error---than less structured
 regions. Furthermore, the evolutionary divergence of a dataset affects
-the power of sitewise inference and the prevalence of indels in
-different ways. Because maximum-likelihood methods for detecting
-positive selection require data in the form of observed substitution
-events, they show little power at low divergence and their highest
-power at intermediate to high divergence levels
+the power of \sw inference and the prevalence of indels in different
+ways. Because maximum-likelihood methods for detecting positive
+selection require data in the form of observed substitution events,
+they show little power at low divergence and their highest power at
+intermediate to high divergence levels
 \citep{Anisimova2001Accuracy}. On the other hand, the number of indels
-in a dataset should increase monotonically with divergence. These two
-trends suggest that the overall power will be low at both extremes of
-divergence, with little inference power at low divergences (due to the
-scarcity of data in the form of observed substitutions) and an
-overwhelming amount of alignment error at high divergences (due to the
-large number of indel events). It is important to note, however, that
-the majority of genes in many biological clades of interest (such as
-mammals, vertebrates, fruit flies, and yeast) fall within the middle
-range of divergences where sitewise methods are at their most powerful
-and where multiple alignment is a difficult---but not
-hopeless---problem. As such, it is important to seek an understanding
-of the impact of alignment error on overall error rates within this
-important range of divergence levels.
+in a dataset should increase with divergence. These two trends suggest
+that the overall power will be low at both extremes of divergence,
+with little inference power at low divergence (due to the scarcity of
+data in the form of observed substitutions) and an overwhelming amount
+of alignment error at high divergence (due to the large number of
+indel events). It is important to note, however, that the majority of
+genes in many biological clades of interest (such as mammals,
+vertebrates, fruit flies, and yeast) fall within the middle range of
+divergences where \sw methods are at their most powerful and where
+multiple alignment is a difficult---but not hopeless---problem. As
+such, it is important to seek an understanding of the impact of
+alignment error on overall error rates within this important range of
+divergence levels.
 
@@
 \citep{Schneider2009Estimates,Markova-Raina2011High} . Most recently,
-Markova-Raina et al. \citeyearpar{Markova-Raina2011High} showed that
-the detection of positively-selected sites and genes in Drosophila
+Markova-Raina and Petrov \citeyearpar{Markova-Raina2011High} showed
+that the detection of positively-selected sites and genes in \Dr
 genomes is highly sensitive to aligner choice, with PRANK's codon
@@
 lack of a benchmark set of true alignments and positively-selected
-sites. Markova-Raina et al. \citeyearpar{Markova-Raina2011High} used
-the expected general effect of alignment error (an increase in false
-positives due to misalignment of non-homologous codons) as a proxy by
-which to compare different methods, allowing for the conclusion that
-PRANK was the least error-prone aligner in their analysis. However,
-the absolute number of false positives remained uncertain due to the
-possible conflation of multiple sources of error: in addition to
-alignment error, the authors noted that gene mis-annotation was
-responsible for many apparent false positives, and there was also an
-expected error rate from the likelihood inference method itself. This
-limitation leaves important and interesting questions regarding the
-nature of misalignment error and its quantitative impact on the
-detection of positive selection left unanswered.
+sites. Markova-Raina and Petrov \citeyearpar{Markova-Raina2011High}
+used the expected general effect of alignment error (an increase in
+false positives due to misalignment of non-homologous codons) as a
+proxy by which to compare different methods, allowing for the
+conclusion that PRANK was the least error-prone aligner in their
+analysis. However, the absolute number of false positives remained
+uncertain due to the possible conflation of multiple sources of error:
+in addition to alignment error, the authors noted that gene
+mis-annotation was responsible for many apparent false positives, and
+there is also an expected error rate from the likelihood inference
+method itself. This limitation leaves important and interesting
+questions regarding the nature of alignment error and its quantitative
+impact on the detection of positive selection left unanswered.
 
@@
 \citep{Dwivedi2009Phylogenetic,Ogden2006Multiple,Loytynoja2008PhylogenyAware}
-but those results cannot be easily extrapolated to the analysis of
-sitewise selective pressures. More recently, Fletcher and Yang
+but those results cannot be easily extrapolated to the analysis of \sw
+selective pressures. More recently, Fletcher and Yang
 \citeyearpar{Fletcher2010Effect} performed a series of simulation
@@
 models to detect positive selection in genes. Their results showed
-that most aligners caused false positives by over-aligning codons, and
+that most aligners caused false positives by over-aligning codons and
 that datasets from mammalian and vertebrate gene families contain
-enough evolutionary divergence to make errors resulting from
-misalignment a legitimate concern.
+enough evolutionary divergence to make false positive errors resulting
+from misalignment a legitimate concern.
 
@@
 evolutionary analyses. The popular Gblocks applies a set of heuristic
-criteria (including thresholds on the amount of conservation and
-gappiness) to identify conserved blocks deemed suitable for
-phylogenetic or evolutionary analysis \citep{Castresana2000Selection},
-while aligners such as T-Coffee \citep{Notredame2000TCoffee} and PRANK
-\citep{Loytynoja2005From} produce estimates of alignment confidence or
-reliability. Most recently, alignment confidence scores based on the
+criteria to identify conserved blocks deemed suitable for phylogenetic
+or evolutionary analysis \citep{Castresana2000Selection} while a
+number of aligners such as T-Coffee \citep{Notredame2000TCoffee} and
+PRANK \citep{Loytynoja2005From} produce estimates of alignment
+confidence or reliability. Alignment confidence scores based on the
 robustness of alignments to guide tree perturbations
 \citep{Penn2010Alignment} or the sampling of suboptimal alignments
-\citep{Kim2011PSAR} have been proposed. Unfortunately, despite their
-widespread use, the impact of the many available filtering methods on
-phylogenetic and evolutionary analyses has not been well-studied. Even
-for a single filtering program, Gblocks, results have been
-contradictory: one simulation-based study found that it improved the
-phylogenetic signal \citep{Talavera2007Improvement} while an empirical
-study across a wide range of taxa found that Gblocks-filtered
-alignments produced worse phylogenetic trees
-\citep{Dessimoz2010Phylogenetic}. The effect of filtering alignments
-on the detection of positive selection is largely unexplored, although
-it has become accepted practice to apply published filtering methods
-before testing for positive selection \citep{Studer2008Pervasive,Aguileta2009Rapidly}.
+\citep{Kim2011PSAR} have also been proposed. Unfortunately, despite
+their widespread use, the impact of the many available filtering
+methods on phylogenetic and evolutionary analyses has not been well
+studied. Even for a single filtering program, Gblocks, results have
+been contradictory: one simulation-based study found that it improved
+the phylogenetic signal \citep{Talavera2007Improvement} while an
+empirical study across a wide range of taxa found that
+Gblocks-filtered alignments produced worse phylogenetic trees than
+unfiltered alignments \citep{Dessimoz2010Phylogenetic}. With the
+application of published filtering methods to alignments before
+testing for positive selection becoming standard practice
+\citep{Studer2008Pervasive,Aguileta2009Rapidly}, an analysis of the
+benefit---if any---of alignment filtering to the detection of positive
+selection would seem well-warranted.
 
-This paper aims to use a simulation framework to incorporate
-misalignment and alignment filtering into estimates of the power and
-accuracy of sitewise evolutionary estimates. Our approach is similar
-to that of Anisimova and Yang \citeyearpar{Anisimova2002Accuracy} in
-that we use simulated protein alignments to evaluate methods for
-detecting sitewise positive selection, considering each site to be an
-independent hypothesis test producing a positive or negative
-result. Our experimental design also shares some features with
-Fletcher and Yang's \citeyearpar{Fletcher2010Effect} investigation of
-the performance of branch-site methods for detecting genewise positive
-selection under alignment error. Like Fletcher and Yang
+This paper aims to use a simulation framework to incorporate alignment
+error and alignment filtering into estimates of the error rate and
+power of \sw evolutionary estimates. Our approach is similar to that
+of Anisimova and Yang \citeyearpar{Anisimova2002Accuracy} in that we
+use simulated protein alignments to evaluate methods for detecting \sw
+positive selection, considering each site to be an independent
+hypothesis test producing a positive or negative result. Our
+experimental design also shares some features with Fletcher and Yang's
+\citeyearpar{Fletcher2010Effect} investigation of the performance of
+branch-site method for detecting genewise positive selection under
+alignment error. Like Fletcher and Yang
 \citeyearpar{Fletcher2010Effect}, we simulate codon alignments with
 indels and calculate inferred alignments using a small but diverse
-sample of aligners. Unlike their study, however, we focus on sitewise
-detection of positive selection occurring throughout a phylogeny (as
-opposed to genewise detection of selection acting at specific
-branches), we explore a wider range of plausible tree sizes, indel
-rates and divergence levels, and we evaluate the effects of a number
-of alignment filtering methods on the error rate and power of sitewise
-analysis. Given the recent proliferation of completed and planned
-vertebrate genomes, we pay special attention to choosing simulation
-parameters and analysis methods similar to those commonly encountered
-in the sitewise analysis of vertebrate gene families.
+sample of aligners. Unlike their study, here we focus on \sw detection
+of positive selection occurring throughout a phylogeny (as opposed to
+genewise detection of selection acting at specific branches), we
+explore a wider range of plausible tree sizes, indel rates and
+divergence levels, and we evaluate the impact of a number of alignment
+filtering methods on the \sw analysis. Given the recent proliferation
+of completed and planned vertebrate genomes, we pay special attention
+to choosing simulation parameters and analysis methods similar to
+those commonly encountered in the \sw analysis of vertebrate gene
+families.
 
@@
 
+\input{table2.txt}
+
 \subsection*{Alignment Simulations}
 
-Three rooted trees were used to guide the simulation of protein-coding
-DNA alignments: the artificial 6-taxon tree used by Anisimova and Yang
+An overview of the simulation parameters used in this study can be
+found in Table \ref{table_1}. Three rooted trees were used to guide
+the simulation of protein-coding DNA alignments: the artificial
+6-taxon tree used by Anisimova and Yang
 \citeyearpar{Anisimova2001Accuracy} and Massingham and Goldman
@@
 \citep{2007Identification,Nikolaev2007Early}. Trees, shown with their
-original branch lengths in Figure 1, were scaled to comparable
-divergence levels by normalizing their root-to-tip mean path length
+original branch lengths in Figure \ref{fig_1}, were scaled to
+comparable divergence levels by normalizing their mean path length
 (MPL), defined as the root-to-tip branch length averaged across all
 lineages in the tree. We simulated alignments with MPL divergence
-between 0.2 and 2.0 synonymous substitutions per synonymous site,
+between 0.05 and 2.0 synonymous substitutions per synonymous site,
 spanning the range of evolutionary divergences observed in several
-clades of organisms with fully-sequenced genomes (Table ST1).
-
-\begin{figure}[h]
+clades of organisms with fully-sequenced genomes (Table
+\ref{table_2}).
+\begin{figure}[t]
 \begin{center}
@@
 \end{center}
-\caption{Phylogenetic trees used for simulation and analysis in this
-  article. (a) An artificial tree used in previous simulations
+\caption{Phylogenetic trees used for simulation and analysis. The
+  original scale for each tree is indicated by a scale bar, but trees
+  were scaled to equal mean path length (MPL) divergence levels for
+  simulation. (a) A 6-taxon artificial tree used in previous
+  simulations
   \citep{Anisimova2001Accuracy,Massingham2005Detecting}. (b) A tree
-  estimated from b-globin genes of 17 vertebrates and used in previous
-  empirical analyses and simulation studies
+  estimated from $\beta$-globin genes of 17 vertebrates and used in
+  previous empirical analyses and simulation studies
   \citep{Anisimova2001Accuracy,Anisimova2002Accuracy}. (c) The
-  44-taxon tree used by the ENCODE project
-  \citep{2007Identification,Nikolaev2007Early}. The nodes labeled in
-  red were used as the reference species (see Methods).}
+  44-species tree used by the ENCODE project
+  \citep{2007Identification,Nikolaev2007Early}. The nodes indicated by
+  arrows were used as the reference species when comparing the true
+  and inferred alignment (see Methods).}
 \label{fig_1}
@@
 ratio of transition to transversion substitutions) was fixed at 4, and
-100 replicate alignments were simulated for each set of
-conditions. Indel lengths were drawn from a discretized power-law
-distribution with an exponential decay parameter of 1.8 and a maximum
-value of 40, yielding a mean indel length of 3.33 codons and standard
-deviation of 5.51 codons. This power-law model of indel lengths is
-well-supported by empirical studies
-\citep{Benner1993Empirical,Cartwright2009Problems} and it was decided
-after manual inspection of alignments from a range of parameter values
-that the chosen model parameters resulted in alignments most closely
+100 replicate alignments were simulated for each set of conditions
+unless otherwise noted. Indel lengths were drawn from a discretized
+power-law distribution with an exponential decay parameter of 1.8 and
+a maximum value of 40, yielding a mean indel length of 3.33 codons and
+standard deviation of 5.51 codons. This power-law model of indel
+lengths is well-supported by empirical studies
+\citep{Benner1993Empirical,Cartwright2009Problems} and manual
+inspection of alignments from a range of parameter values identified
+the chosen model parameters as resulting in alignments most closely
 resembling those encountered in vertebrate alignments. The ratio of
 insertion to deletion events was set to 1, and the rate of indel
-formation was scaled from between 0 to 0.2 indel events per
-substitution per site.
+formation was scaled between 0 to 0.2 indel events per substitution
+per site.
 
-The distribution of sitewise selective pressures was modeled with a
-log-normal distribution derived from a maximum-likelihood fit to a
-large dataset of sitewise selective pressures estimated from mammalian
-gene trees (log-normal parameters shown in Table 1). This
-distribution, with mean w of 0.28 and 6\% of sites having w>1, is
-consistent with the structure-based expectation of many protein sites
-under purifying selection and few under neutral selection or positive
-selection
+The distribution of \sw selective pressures (embodied by the parameter
+$\omg$, the ratio of the rate of non-synonymous substitution to the
+rate of synonymous substitution) was modeled with a log-normal
+distribution derived from a maximum-likelihood fit to a large dataset
+of \sw selective pressures estimated from mammalian gene trees
+(unpublished data; log-normal parameters shown in Table
+\ref{table_1}). This distribution, with mean $\omg$ of 0.28 and 6\% of
+sites having $\omg>1$, is consistent with the structure-based
+expectation of many protein sites under purifying selection and few
+under neutral selection or positive selection
 \citep{Smith1970Natural,Kimura1974SomePrinciples}. Indelible's general
-discrete model of sitewise w variation was used to approximate this
-distribution by splitting the lognormal probability density into 50
-equally-spaced bins between w values of 0 and 3, with the highest bin
-containing the probability density for all values w>=3.
+discrete model of \sw $\omg$ variation was used to approximate the
+lognormal distribution by splitting the probability density into 50
+equally-spaced bins between $\omg$ values of 0 and 3, with the highest
+bin containing the probability density for all values $\omg>3$.
 
@@
 per codon ($t$) \citep{Fletcher2010Effect}. They are related
-approximately by $t = 3(NdN + SdS) = 3dS(\omg{}N + S)$, where $N$ and
+approximately by $t=3(NdN+SdS) = 3dS(\bar{\omg{}}N+S)$, where $N$ and
 $S$ are the proportion of nonsynonymous and synonymous sites and
-$\omg{}$ is the mean $\omg{}$ across all sites. S is approximately 0.3
-when $\bm{\kappa}=4$ and the mean $\omg{}$ ratio for our chosen
+$\bar{\omg}$ is the mean $\omg$ across all sites. $S$ is approximately 0.3
+when $\bm{\kappa}=4$ \citep{Yang1998Synonymous} and the mean $\omg$ ratio for our chosen
 distribution is 0.277, yielding a $dS$-to-$t$ conversion factor of
@@
 Alignments were filtered by masking out residues based on the output
-of four alignment scoring methods: Gblocks conserved blocks
+of three alignment scoring methods: Gblocks conserved blocks
 \citep{Castresana2000Selection}, T-Coffee consistency scores
-\citep{Notredame2000TCoffee,Notedame2003Using}, and GUIDANCE alignment
-confidence scores \citep{Penn2010Alignment}. Gblocks, which identifies
-entire alignment columns as conserved or not conserved, was run using
-a relaxed gap allowance parameter in order to avoid removing too much
-of the alignment (command-line parameter 'b5=a') and all residues from
-any columns not within an identified conserved block were masked with
-'N's. GUIDANCE and T-Coffee, which produce individual scores for each
-residue, were handled differently. (TODO!!!!). GUIDANCE was run with 30
+\citep{Notredame2000TCoffee,Notredame2003Using}, and GUIDANCE
+alignment confidence scores \citep{Penn2010Alignment}. Gblocks, which
+identifies entire alignment columns as conserved or not conserved, was
+run using an increased gap tolerance and a reduced minimum block
+length in order to reduce the amount of each alignment removed
+(command-line parameters {\em b5=a} and {\em b4=3}), and all residues
+from any columns not within an identified conserved block were masked
+with $N$s. GUIDANCE and T-Coffee, which produce individual scores for
+each residue, were handled differently. A cutoff threshold was chosen
+for each method (0.5 for GUIDANCE and 5 for T-Coffee) and residues
+equal to or below that threshold were masked. If the cutoff threshold
+caused greater than 50\% of an alignment's residues to be masked, the
+threshold was relaxed to the highest value for which at least 50\% of
+residues remained unmasked. This cap on the severity of each filter
+was necessary because the scores from both methods were strongly
+affected by the simulation conditions, with much lower average scores
+at higher indel rates and divergences. GUIDANCE was run with 100
 bootstrap replicates and MAFFT as the bootstrap aligner (except when
-\pranka{} or \prankc{} was the original aligner, in which case \pranka{}
-was used to align the bootstrap replicates), and T-Coffee was run
-using default settings and the 'evaluate\_mode -output=score\_ascii'
-command-line parameter to output alignment scores.
+\pranka{} or \prankc{} was the original aligner, in which case 30
+bootstrap replicates were generated with \pranka{}) and T-Coffee was
+run using default settings and the {\em evaluate\_mode
+  -output=score\_ascii} command-line parameters to output alignment
+scores.
 
 Two unrealistic datasets were produced to serve as controls. First,
-the true simulated alignment was used in order to evaluate the
-sitewise performance without any alignment error. Second, an
-additional filtering method was designed to represent an unattainable
+the true simulated alignment was included in order to evaluate the \sw
+performance without any alignment error. Second, an additional
+filtering method was constructed to represent an unattainable
 best-case scenario for sequence filtering, using knowledge of the true
-alignment to assign a score to each residue that summarizes how
-correctly is has been placed in the inferred alignment. The approach
-taken was to calculate, for each residue, the branch length of the
-correct sub-tree (defined as the minimum sub-tree connecting all
-sequences to which the current residue is correctly aligned) divided
-by the branch length of the aligned sub-tree (defined as the minimum
-sub-tree connecting all sequences with non-gap residues at the current
+alignment to assign a score to each residue reflecting how correctly
+is has been placed in the inferred alignment. The approach taken was
+to calculate, for each residue, the branch length of the correct
+sub-tree (defined as the sub-tree connecting all sequences to which
+the current residue was correctly aligned) divided by the branch
+length of the total aligned sub-tree (defined as the sub-tree
+connecting all sequences with non-gap residues at the current
 alignment column). This residue-wise score ranges from 0 to 1 and
-encapsulates the expectation that correctly-aligned evolutionary
-branch length is the main source of information from which the
-sitewise inference methods derive their power. Scores from this
-method, referred to as the Optimal method, were used to filter
-inferred alignments in a manner similar to the T-Coffee and GUIDANCE
-scores.
+reflects the expectation that correctly-aligned evolutionary branch
+length is the main source of information from which \sw inference
+methods derive their power. We refer to this method as the optimal
+method. Scores were handled in a manner similar to GUIDANCE and
+T-Coffee, using a score threshold of 0.5.
 
-\subsection*{Sitewise Evolutionary Analysis}
+\subsection*{\Sw Evolutionary Analysis}
 
-Sitewise estimates of selective pressures were calculated using
+\Sw estimates of selective pressures were calculated using
 maximum-likelihood methods implemented in the Phylogenetic Analysis by
@@
 
-The codeml program from PAML implements a number of likelihood ratio
+The $codeml$ program from PAML implements a number of likelihood ratio
 tests (LRTs) for detecting the presence of positive selection in a
-gene while allowing the $\omg{}$ ratio to vary among sites
+gene while allowing the $\omg$ ratio to vary among sites
 \citep{Yang2000CodonSubstitution}. These models, known as the sites or
 random sites models, use a variety of predefined statistical
-distributions to account for heterogeneous w ratios among sites. After
+distributions to account for heterogeneous $\omg$ ratios among sites. After
 the likelihood optimization is performed, Bayesian methods can be used
@@
 given site class, where a high posterior probability of a site
-belonging to a class with $\omg{}>1$ can be considered strong
+belonging to a class with $\omg>1$ can be considered strong
 evidence that a site has evolved under positive selection
@@
 
-SLR implements a method specifically designed for sitewise estimates
-which has been shown in simulations to perform as well or better than
-PAML's sitewise models \citep{Massingham2005Detecting}. SLR models
-codon evolution as a continuous-time Markov process where
+SLR implements a method specifically designed for \sw estimates which
+has been shown in simulations to perform as well as or better than
+PAML's \sw random sites models \citep{Massingham2005Detecting}. SLR
+models codon evolution as a continuous-time Markov process where
 substitutions at one site are independent of substitutions at all
 other sites. Importantly, no assumptions are made regarding the
-distribution of $\omg{}$ ratios within the alignment. The value
-of $\omg{}$ is considered to be an independent parameter at each
-site: after first optimizing shared parameters using the whole
-alignment, SLR uses the shared parameters and the data at each
-alignment site to calculate a sitewise statistic for non-neutral
-evolution. This statistic is based on a likelihood-ratio test where
-the null model is neutral evolution ($\omg{}=1$) and the
-alternative model is either purifying or positive selection
-($\omg{}<1$ or $\omg{}>1$, respectively). The raw statistic
-measures the strength of evidence for non-neutral evolution at each
-site; following Massingham \citeyearpar{Massingham2005Detecting} we
-use a signed version of the SLR statistic (created by negating the
-statistic for sites with $\omg{}<1$) as the test statistic for
-positive selection.
+distribution of $\omg$ ratios within the alignment. The value of
+$\omg$ is considered to be an independent parameter at each site:
+after first optimizing shared parameters using the whole alignment,
+SLR uses the shared parameters and the data at each alignment site to
+calculate a \sw statistic for non-neutral evolution. This statistic is
+based on a likelihood-ratio test where the null model is neutral
+evolution ($\omg=1$) and the alternative model is either purifying or
+positive selection ($\omg<1$ or $\omg>1$, respectively). The raw
+statistic measures the strength of evidence for non-neutral evolution
+at each site; following Massingham
+\citeyearpar{Massingham2005Detecting} we use a signed version of the
+SLR statistic (created by negating the statistic for sites with
+$\omg<1$) as the test statistic for positive selection.
 
@@
 
-In order to compare sitewise estimates from different alignments, a
-single sequence from each tree was chosen as the reference (red
-labels, Figure 1) and all sitewise statistics were mapped from
-alignment columns to sequence positions in the reference
-sequence. This approach bears some resemblance to the process of
-mapping alignment-based evolutionary inferences onto a single member
-of the alignment for further analysis and integration with other
-genome-referenced data (as is often done, for example, using mammalian
-alignments and a human reference). As a result of this reference
-sequence based mapping, sites which were deleted in the reference
-sequence or inserted in a lineage not ancestral to the reference were
-dropped from the analysis.
+In order to compare \sw estimates from different alignments, a single
+sequence from each tree was chosen as the reference (arrows, Figure
+\ref{fig_1}) and all \sw statistics were mapped from alignment columns
+to sequence positions in the reference sequence. This approach bears
+some resemblance to the process of mapping alignment-based
+evolutionary estimates onto a single member of the alignment for
+further analysis and integration with other genome-referenced data (as
+is often done, for example, using mammalian alignments and a human
+reference). As a result of this reference sequence based mapping,
+sites which were deleted in the reference sequence or inserted in a
+lineage not ancestral to the reference were dropped from the analysis.
 
@@
 of 0.95 was used for PAML \citep{Yang2005Bayes} and a threshold of 3.84,
-the 95\% critical value of the chi-square distribution with 1 degree
+the 95\% critical value of the $\chi^2$ distribution with 1 degree
 of freedom, was used for SLR \citep{Massingham2005Detecting}. Sites were
@@
 positive rate (FPR, defined as the proportion of all sites with true
-$\omg{}<1$ falsely identified as positively selected) and true positive rate
-(TPR, defined as the proportion of all sites with true $\omg{}>1$ correctly
+$\omg<1$ falsely identified as positively selected) and true positive rate
+(TPR, defined as the proportion of all sites with true $\omg>1$ correctly
 identified as positively selected).
@@
 error rates differently for each combination of simulation condition
-and aligner, we also identified the score thresholds for each dataset
-that resulted in an actual FPR of 1\% and calculated the TPR achieved
-at this actual error rate (referred to as \tpr{} to distinguish it
-from the TPR described above). Although this estimate of
+and aligner, we identified the score thresholds for each dataset that
+resulted in an actual FPR of 1\% and calculated the TPR achieved at
+this actual error rate (hereafter referred to as \tpr{} to distinguish
+it from the TPR described above). Although this estimate of
 error-controlled power would be impossible to calculate in an
@@
 simulation context for allowing a controlled comparison of the
-performance of sitewise analysis between different
-conditions. Specifically, it should be sensitive both to increased
-error rates from alignment error and to reduced power from alignment
-filtering, in both cases showing a lowered error-controlled power as
-the method performs worse.
+performance of \sw analysis between different
+conditions. Specifically, it should be sensitive to changes in the
+numbers of both false positives and false negatives resulting from
+alignment error or alignment filtering; in both cases a lowered
+error-controlled power would result, as fewer true positives are
+identified at the constant 1\% FPR.
 
-\section*{Results and Discussion}
+We also evaluated the ability of each method to accurately infer the
+$\omg$ value at each site by collecting sitewise $\omg$ estimates from
+the output of each method and calculating the Pearson's correlation
+coefficient between the true and inferred $\omg$ values for each set
+of simulation conditions.
 
-\subsection*{Performance of Three Methods for the Sitewise Detection of Positive Selection}
-
-\begin{figure}[h]
+\begin{figure}[t]
 \begin{center}
@@
 \end{center}
-\caption{The behavior of three methods of sitewise inference in the
-  absence of indels. One hundred replicate alignments were simulated
-  without indels for each tree and divergence level and were analyzed
-  with either PAML M2 (red), PAML M8 (blue), or SLR. The performance
-  of each method, as measured by four summary statistics, is plotted
-  as a function of divergence level. From top to bottom: TPR at the
-  recommended cutoff threshold (0.95 for PAML and 2.71 for SLR), FPR
-  at the recommended cutoff threshold, TPR at a 1\% FPR threshold, and
-  Pearson's correlation coefficient between the true and estimated
-  $\omg{}$.}
+\caption{Alignments were simulated without indels for three tree
+  shapes and analyzed with SLR, PAML M8, or PAML M2. Fifty replicate
+  alignments were simulated for each data point. The performance of
+  each analysis method, as measured by four summary statistics, is
+  plotted as a function of the mean path length (MPL) divergence. From
+  top to bottom: true positive rate (TPR) at the recommended cutoff
+  threshold (0.95 for PAML and 3.84 for SLR); false positive rate
+  (FPR) at the recommended cutoff threshold; TPR at a 1\% FPR
+  threshold; Pearson's correlation coefficient between the true and
+  inferred sitewise $\omg$.}
 \label{fig_2}
@@
 
-We first evaluated the ability of three sitewise methods, PAML M2,
-PAML M8 and SLR, to accurately estimate sitewise $\omg{}$ values and to
+\section*{Results and Discussion}
+\subsection*{The Performance of Three Methods for Detecting \Sw Positive Selection}
+We first evaluated the ability of three \sw methods, PAML M2,
+PAML M8 and SLR, to accurately estimate \sw $\omg$ values and to
 detect positive selection under a range of tree lengths in the absence
-of alignment error. Figure 2 shows the TPR, FPR, \tpr{} and sitewise $\omg{}$
+of alignment error. Figure \ref{fig_2} shows the TPR, FPR, \tpr{} and \sw $\omg$
 correlation over a range of MPLs for each of the three simulation
@@
 
-The detection power and ω correlation were weakest at low divergence
-levels for all methods and all trees due to the low amount of
-evolutionary information, as observed in previous simulations
-\citep{Anisimova2002Accuracy}. Also confirming previous results, we
-found a positive correlation between tree size and detection power,
-with the highest values in the 44-taxon tree. Power generally
-increased monotonically with divergence, except for the 6-taxon tree
-which saw its maximum performance at moderate divergence levels (MPL
-0.5-1.0) and began decreasing at higher values. The downward trend in
-the 6-taxon tree was likely due to the impact of saturation of
-synonymous sites in the very long branches present in such a sparse
-tree at high divergence levels. The two larger trees, with lower
-average branch lengths at equivalent divergence levels, showed less
-susceptibility to saturation of substitutions along a single
-branch. Interestingly, the two larger trees showed no signs of
-decreased performance even at a MPL of 2 substitutions per site, which
-is greater than any of the divergence levels found in groups of
-commonly analyzed vertebrate, insect and fungal species (Table ST1).
+The detection power and $\omg$ correlation were weakest at low
+divergence levels for all methods and all trees due to the low amount
+of evolutionary information, as observed in previous simulations
+\citep{Anisimova2002Accuracy}. We found a positive correlation between
+tree size and detection power, with the highest performance in the
+44-taxon tree. Power generally increased monotonically with
+divergence, except for the 6-taxon tree which saw its maximum
+performance at moderate divergence levels (MPL 0.5--1.0) and began
+decreasing at higher values. The downward trend in the 6-taxon tree
+was likely due to the impact of saturation of synonymous sites in the
+very long branches present in such a sparse tree at high divergence
+levels. With lower average branch lengths at equivalent divergence
+levels, the two larger trees showed no signs of decreased performance
+even at a MPL of 2 substitutions per site, which is greater than any
+of the divergence levels found in groups of commonly analyzed
+vertebrate, insect and fungal species (Table \ref{table_2}).
 
-Comparing between the three sitewise methods, we found that at the
-recommended cutoff threshold (Figure 2, top row) SLR showed the
-highest power to detect positive selection in all trees, followed by
-PAML M8 and PAML M2. In the smallest tree, the power of the two PAML
-methods was virtually zero while SLR reached a maximum of ~18\% TPR at
-MPL=0.5. SLR showed a large increase in power in the larger trees
-compared to the smallest tree, yielding ~25\% and ~45\% TPR at MPL=0.5
-in the 17-taxon and 44-taxon trees, respectively. The two PAML methods
-also showed increased power in the larger trees, with PAML M2 ranging
-between 50-75\% of SLR’s power and the performance of PAML M8 falling
-in between the two other methods.
+Comparing between the three \sw methods, we found that at the
+recommended cutoff threshold (Figure \ref{fig_2}, top row) SLR showed
+the highest power to detect positive selection in all trees, followed
+by PAML M8 and PAML M2. In the smallest tree, the power of the two
+PAML methods was virtually zero while SLR reached a maximum TPR of ~18\%,
+at MPL=0.5. At the same divergence, SLR yielded TPRs of 25\% and 45\%
+in the 17-taxon and 44-taxon trees, respectively, with PAML M2 ranging
+between 50--75\% of SLR’s power and PAML M8 falling in between the two
+other methods.
 
 The TPR measurements represent the power that might be achieved in
-real-world analysis using recommended cutoff thresholds, but SLR’s
-higher power may reflect merely a shifted balance between power and
-accuracy at the recommended cutoff threshold as opposed to an
-increased ability to discriminate positive from neutral or purifying
-selection. The FPR and error-controlled \tpr{} results showed the
-former to be the case. The FPR from SLR was higher than that from
-either of the PAML methods for all trees and divergence levels (Figure
-2b), suggesting that its higher power was the result of a
-less-conservative cutoff rather than significantly improved
-statistical performance. This hypothesis was confirmed by evaluating
-the TPR at a cutoff threshold for each method that controlled for an
-actual FPR of 1\% (\tpr{}, shown in Figure 2c), yielding virtually
-identical \tpr{} values for all three methods. The highly similar
-error-controlled power levels indicate that the three methods’
-sitewise statistics were nearly equally sensitive to positive
-selection under these simulation conditions.
+real-world analysis using recommended cutoff thresholds, but the
+higher power from SLR may merely reflect a shifted balance between
+power and accuracy at the recommended cutoff threshold as opposed to
+an increased absolute ability to discriminate positive from neutral or
+purifying selection. The FPR and error-controlled \tpr results
+revealed that this was indeed the case: the FPR from SLR was higher
+than that from either of the PAML methods for all trees and divergence
+levels, suggesting that its higher power was the result of a
+less-conservative cutoff value. This was further verified by
+evaluating the TPR at a cutoff threshold that controlled for an actual
+FPR of 1\% for each method (\tpr{}, third row in Figure
+\ref{fig_2}). The error-controlled \tpr values were virtually
+identical for all three methods, providing strong evidence that the
+three methods' \sw statistics were nearly equally sensitive to
+positive selection under our chosen simulation conditions.
 
@@
 have been previously noted
-\citep{Anisimova2002Accuracy,Yang2005Bayes,Massingham2005Detecting}
-and the extremely low false positive rates in our simulations
-confirmed that all three methods would yield very few false positives
-with a realistic distribution of $\omg{}$ values across a range of
-tree sizes and divergence levels. For the purposes of our indel
-experiments, the observed similarity in error-controlled power levels
-indicated that the behavior of PAML M2, PAML M8, and SLR was similar
-enough not to warrant separately evaluating all three methods in the
-subsequent indel simulation experiments. As the runtime for SLR was
-significantly lower than that of either PAML model, all subsequent
-results are presented only based on the SLR test.
+\citep{Anisimova2002Accuracy,Yang2005Bayes,Massingham2005Detecting},
+but the extremely low false positive rates in our simulations showed
+that in the absence of alignment error all three methods would yield
+very few false positives when analyzing genes with a typical
+mammalian-like distribution of $\omg$ values. The low FPRs were likely
+due to the large proportion of sites under moderately strong purifying
+selection in our $\omg$ distribution used for simulation. Such sites
+are less likely to yield false positives than sites under neutral
+evolution, which the null model against which tests for positive
+selection are traditionally controlled.
 
-\subsection*{Effect of Alignment Error on Sitewise Power}
+For the purposes of our indel experiments, the observed similarity in
+error-controlled power levels indicated that the behavior of PAML M2,
+PAML M8, and SLR was similar enough not to warrant separately
+evaluating all three methods in the subsequent indel simulation
+experiments. As the runtime for SLR was significantly lower than that
+of either PAML model, all subsequent results are presented only based
+on the SLR test.
+
+\subsection*{The Effect of Alignment Error on \Sw Power}
 
@@
 significantly for different tree sizes, alignment algorithms, and
-evolutionary divergences. Figure 3 shows the same sitewise
-measurements as Figure 2 for simulations without indels and with
-indels, analyzed using three different aligners and the true
-alignment. The indel rate was held constant at 0.1 indel event per
-substitution, and the divergence level was varied between 0.2 and 2.0
-substitutions per site.
-
+evolutionary divergences. Figure \ref{fig_3} shows the same
+performance measurements as Figure \ref{fig_2} for simulations without
+indels (gray lines, Figure \ref{fig_2}) and with indels (black lines,
+Figure \ref{fig_2}) analyzed using three different aligners and the
+true alignment. For the indel simulations, the indel rate was held
+constant at 0.1 indel event per substitution.
 \begin{figure}[t]
@@
 \end{center}
-\caption{The behavior of sitewise inference in the presence and absence of indels using three different aligners and the true alignment.}
+\caption{Sequences were simulated without indels (gray lines) or with
+  indels (black lines) using one of three tree shapes, aligned with
+  one of three aligners, and analyzed with SLR; true alignments were
+  separately analyzed with SLR (solid black lines). One hundred
+  replicate alignments were simulated for each data point. The
+  performance of each dataset, as measured by four summary statistics,
+  is plotted as a function of the mean path length (MPL)
+  divergence. From top to bottom: true positive rate (TPR) at the
+  recommended cutoff threshold; false positive rate (FPR) at the
+  recommended cutoff threshold; TPR at a 1\% FPR threshold; Pearson's
+  correlation coefficient between the true and inferred $\omg$.}
 \label{fig_3}
@@
 Comparing the results without indels to those with indels under the
-true alignment (shown as solid black and solid red lines in Figure 3)
-we found a slight decrease in power and ω correlation and no
-noticeable increase in FPR. The decreased power was expected, since
-even in the absence of alignment error, alignment columns containing
-gaps harbor less evolutionary information than columns flush with
-sequence data. The lack of increased FPR showed that SLR retained its
-conservative statistical performance even when analyzing gapped
-alignments with indels. Surprisingly, at higher divergences (MPL $>$
-1.0) under the six-taxon tree, the FPR with indels was lower than the
-FPR without indels. This unexpected result may be attributed to the
-large number of alignment columns under such conditions that contain
-only a single non-gap sequence, as these columns are never inferred as
-positively-selected by SLR due to the complete lack of
+true alignment we found a slight decrease in power and $\omg$
+correlation and no noticeable increase in FPR. The decreased power was
+expected, since even in the absence of alignment error alignment
+columns containing gaps harbor less evolutionary information than
+columns flush with sequence data. The lack of increased FPR showed
+that SLR retained its conservative statistical performance even when
+analyzing gapped alignments. Surprisingly, at higher divergences (MPL
+$>$ 1.0) under the six-taxon tree, the FPR with indels was lower than
+the FPR without indels. This unexpected result may be attributed to
+the large number of alignment columns under such conditions that
+contained only a single non-gap sequence, as those columns were never
+inferred as positively-selected by SLR due to the complete lack of
 information. The two larger trees did not show a similar trend at high
-divergence levels, suggesting that the effect was due mainly to the
-highly sparse nature of the alignment in the 6-taxon tree. The rest of
-our results from the 6-taxon tree at high divergences were similarly
+divergence levels, suggesting that this effect was indeed due to the
+highly sparse nature of the alignments in the 6-taxon tree. All other
+results from the 6-taxon tree at high divergences were similarly
 anomalous in this respect; we surmised that the sparseness of the true
-alignment combined with the extreme difficulty of accurately aligning
-sequences along very long branches made sitewise analysis with indels
-very unreliable at high divergences in the smallest tree.
+alignment, combined with the extreme difficulty of accurately aligning
+sequences along very long branches, made \sw analysis with indels very
+unreliable at high divergences in the smallest tree.
 
 When alignments were inferred using one of the three aligners tested,
-the TPR, \tpr{} and $\omg{}$ correlation were all reduced relative to
-the true alignment (dashed and dotted red lines, Figure 3A, B, D). The
+the TPR, \tpr{} and $\omg$ correlation were all reduced relative to
+the true alignment (dashed and dotted lines, Figure \ref{fig_3}). The
 degree of reduction varied depending on the aligner, simulation
 conditions, and performance measurement being analyzed. At low
-divergences (e.g. MPL $<$ 0.2), the inferred alignments generally
-showed only a small decrease in performance, but the \tpr{} in the
-6-taxon tree was a notable exception with large differences down to a
-MPL of 0.1. As divergence levels increased, so did the difference
-between the performance of the true alignment and the inferred
-alignments. The three aligners tested could be consistently and
-unambiguously ranked by each of the measured performance
-characteristics, with \prankc{} always performing best and ClustalW
-performing worst. The same ranking of aligners with respect to
-detecting positive selection was found by Fletcher and Yang
+divergences (e.g. MPL $<$ 0.2) the inferred alignments generally
+showed only a small decrease in performance. As divergence levels
+increased, so did the difference between the performance of the true
+alignment and the inferred alignments. The three aligners tested could
+be consistently and unambiguously ranked by all of the measured
+performance characteristics, with \prankc always performing best and
+ClustalW performing worst. The same ranking of aligners with respect
+to detecting positive selection was found by Fletcher and Yang
 \citeyearpar{Fletcher2010Effect}; our results corroborate their
@@
 
-When alignments were inferred using one of the three aligners tested,
-the TPR, \tpr{} and $\omg{}$ correlation were all reduced relative to
-the true alignment (dashed and dotted red lines, Figure 3A, B, D). The
-degree of reduction varied depending on the aligner, simulation
-conditions, and performance measurement being analyzed. At low
-divergences (e.g. MPL < 0.2), the inferred alignments generally showed
-only a small decrease in performance, but the \tpr{} in the 6-taxon
-tree was a notable exception with large differences down to a MPL of
-0.1. As divergence levels increased, so did the difference between the
-performance of the true alignment and the inferred alignments. The
-three aligners tested could be consistently and unambiguously ranked
-by each of the measured performance characteristics, with \prankc{}
-always performing best and ClustalW performing worst. The same ranking
-of aligners with respect to detecting positive selection was found by
-Fletcher and Yang \citeyearpar{Fletcher2010Effect}; our results
-corroborate their finding and provide evidence that this ranking may
-be applicable to a wider array of evolutionary analyses than the
-branch-site tests evaluated in their analysis.
+Looking at the TPR results for inferred alignments, we observed that
+in the 6-taxon tree the three aligners formed a cluster of lines well
+below the true alignment value, indicating similar tendencies among
+the different aligners to produce false negatives in the smaller
+tree. In larger trees the different aligners showed a wider spread of
+TPR values, but even \prankc{} showed a 5--10\% reduction compared to
+the true alignment at MPL$=$1.0. These results show that the
+introduction of false negatives is a significant and seemingly
+unavoidable result of alignment error at medium to high divergence
+levels (MPL $>$ 0.5), with even the most powerful aligner producing a
+marked reduction in TPR compared to the true alignment. The \tpr and
+$\omg$ results were qualitatively similar to the TPR results, with the
+\tpr values slightly more spread out between the different aligners
+than the TPRs, and the $\omg$ correlations slightly less so.
 
-We found that FPR levels under inferred alignments were generally
-higher than under the true alignment. This effect was especially
-strong in the 6-taxon tree at medium divergence levels (e.g. MPL
-0.2-0.6), where ClustalW showed up to a fourfold increase, and \prankc{}
-a nearly twofold increase, in FPR over the true alignment. As
-previously noted, the 6-taxon tree showed an unexpected and presumably
-anomalous FPR pattern at higher divergences, likely due to the highly
-sparse true alignment. In the two larger trees, FPRs from inferred
-alignments were less elevated compared to the true alignment and
-relatively constant across the range of divergences. ClustalW's FPR
-ranged between 0.003 to 0.005, while \prankc{}'s FPR was virtually
-identical to the true alignment in both the 17-taxon and 44-taxon
-trees.
+The FPRs for infered alignments exhibited a very different trend from
+the other performance measures, with generally higher FPRs than the
+true alignment and the widest range of values occurring in the 6-taxon
+tree. In this tree at medium divergence levels (e.g. MPL 0.2--0.6)
+ClustalW showed up to a fourfold increase, and \prankc{} a nearly
+twofold increase, in FPR over the true alignment. As previously noted,
+the 6-taxon tree showed an anomalous FPR pattern at higher
+divergences, with lower FPRs for inferred alignments than the true
+alignment, likely due to the highly sparse true alignment under those
+conditions. In the two larger trees, FPRs from inferred alignments
+were less elevated compared to the true alignment, less variable
+between aligners, and relatively constant across the range of
+divergences. ClustalW's FPR ranged between 0.001 to 0.005, while
+\prankc{}'s FPR was virtually identical to the true alignment in the
+17-taxon and 44-taxon trees.
 
-A comparison of the TPR and FPR results between true and inferred
-alignments allows the relative impact of false positive and false
-negative errors resulting from misalignment to be quantified. A false
-positive resulting from over-alignment of non-synonymous codons would
-contribute to an elevated FPR, while a false negative resulting from
-non-alignment of homologous codons would contribute to a reduced
-TPR. The \tpr{} should be sensitive to both false positives and false
-negatives since it measures the TPR when controlling for an actual FPR
-of 1\%. Analyzing the TPR results in the 6-taxon tree, the three
-aligners formed a cluster of lines well below the true alignment
-value, indicating similar tendencies among the different aligners to
-produce false negatives in the smaller tree. In larger trees the
-different aligners showed a wider spread of TPR values, but even
-\prankc{} showed a 5-10\% reduction compared to the true alignment at
-MPL$=$1.0. These results show that the introduction of false negatives
-is a significant and seemingly unavoidable result of alignment error
-at medium to high divergence levels (MPL $>$ 0.5), with even the most
-powerful aligner producing a marked reduction in TPR compared to the
-true alignment. The FPR levels exhibited a very different trend, with
-the widest range of values and highest FPRs in the 6-taxon tree and
-the smallest range and lowest values in the 44-taxon
-tree. Furthermore, the aligner with the lowest FPR, \prankc{}, had FPR
-levels almost identical to those of the true alignment in the 17-taxon
-and 44-taxon trees. The FPR results suggest that although false
-positives resulting from misalignment are unavoidable in a
-moderately-diverged 6-taxon tree, the addition of more taxa to the
-tree provided enough information for \prankc{} to yield very few false
-positives not in the true alignment. Other aligners, however,
-maintained a constant elevated FPR in the two larger trees, indicating
-a failure to effectively utilize the additional evolutionary
-information present in larger trees to avoid introducing false
-positive results.
-
-The results shown in Figure 3 allow us to characterize the combined
-effects of alignment error at different commonly analyzed levels of
-divergence. For example, at a human-mouse divergence level (MPL $=$
-0.2) misalignment had little impact on the TPR regardless of what
-aligner was used. However, ClustalW yielded a notably higher FPR than
-MAFFT or \prankc{}, and the error-controlled \tpr{} was
-correspondingly lower for ClustalW in all three trees. Thus, at small
-divergences we found that false positives were thus the main source of
+We found it useful to combine divergence estimates from Table
+\ref{table_2} with the results from Figure \ref{fig_3} to characterize
+the combined effects of alignment error at different commonly analyzed
+levels of divergence. For example, at a human-mouse divergence level
+(MPL$=$0.2) misalignment had little impact on the TPR regardless of
+what aligner was used. However, ClustalW yielded a notably higher FPR
+than MAFFT or \prankc, and the error-controlled \tpr{} was
+correspondingly lower for ClustalW in all three trees. Thus, at low
+divergences we found that false positives were the main source of
 error from misalignment, and different aligners had highly variable
-tendencies to produce false positive results. At the higher vertebrate
-and Drosophila divergence levels (MPL 0.8-1.0) false negatives became
+tendencies to produce false positive results. At higher vertebrate and
+\Dr divergence levels (MPL$=$0.8--1.0) false negatives became
 much more prevalent. The TPR for all inferred alignments was virtually
 zero in the 6-taxon tree, underscoring the necessity of including many
-species in the analysis of highly diverged sequences. In the larger
-trees, \prankc{} resulted in very few additional false positives, but
-it suffered a 5-10\% reduction in TPR relative to the true alignment
-due to false negatives. Meanwhile, ClustalW showed a 50\% TPR
-reduction and maintained a strongly elevated FPR. At higher
-divergences and in larger trees, false negatives appear to be the most
-persistent alignment error, causing a marked reduction in sitewise
-power even with the best-performing aligner.
+species in the analysis of highly diverged sequences. In the two
+larger trees, \prankc resulted in very few additional false
+positives, but it suffered a 5--10\% reduction in TPR relative to the
+true alignment. Meanwhile, ClustalW showed a 50\% TPR reduction and
+maintained a strongly elevated FPR. At higher divergences and in
+larger trees, false negatives were thus the most persistent effect of
+alignment error, causing a marked reduction in \sw power even with the
+best-performing aligner.
 
@@
 \end{center}
-\caption{}
+\caption{Sequences were simulated with indels using one of three tree
+  shapes and a range of indel rates and mean path length (MPL)
+  divergence levels. Alignments were inferred with one of four
+  aligners and analyzed with SLR; true alignments were separately
+  analyzed with SLR. One hundred replicates were simulated for each
+  set of conditions. Each square is colored according to the
+  performance at a given (indel rate, MPL) pair as measured by one of
+  three summary statistics: (a) the true positive rate (TPR) at the
+  recommended cutoff threshold, (b) the false positive rate (FPR) at
+  the recommended cutoff threshold, or (c) the TPR at a 1\% FPR
+  threshold.}
 \label{fig_4}
@@
 
-\subsection*{Sitewise Power Under a Range of Indel Rates and Divergences}
+\subsection*{\Sw Power Under a Range of Indel Rates and Divergences}
 
 To explore the effects of alignment error across a wider range of
-simulation conditions, we extended the simulations from Figure 3
-across a range of indel rates, from 0 to 0.2 indel events per
-substitution. Figure 4 shows heatmaps of the TPR and FPR for ClustalW,
-\prankc{} and the true alignment (Figure 4A,B) and a heatmap of the
-error-controlled \tpr{} for all aligners tested (Figure 4C). The
-results from Figure 3, which were simulated with an indel rate of 0.1,
-correspond to the middle row of each panel in Figure 4; rows above and
-below the middle row represent higher and lower indel rates,
-respectively.
+simulation conditions, we extended the simulations from Figure
+\ref{fig_3} across multiple indel rates. Figure \ref{fig_4} shows
+heatmaps of the TPR and FPR for ClustalW, \prankc and the true
+alignment (Figure \ref{fig_4}A,B) and a heatmap of the
+error-controlled \tpr for all aligners tested (Figure
+\ref{fig_4}C). The results from Figure \ref{fig_3}, which were
+simulated with an indel rate of 0.1, correspond to the middle row of
+each panel in Figure \ref{fig_4}; rows above and below the middle row
+represent higher and lower indel rates, respectively.
 
-The TPR values (Figure 4A) showed a consistent pattern across the
-range of indel rates, with power decreasing as either the indel rate
-or the divergence level increased. \prankc{} showed a greater ability
-than ClustalW to maintain a high TPR at higher indel rates, especially
-in the 17-taxon and 44-taxon trees. At lower indel rates, the TPR
-performance of both aligners and the true alignment were qualitatively
-similar.
+The TPR values (Figure \ref{fig_4}A) showed a consistent pattern
+across the range of indel rates, with power decreasing as either the
+indel rate or the divergence level increased (except at the lowest
+divergence levels, where the lack of evolutionary information yielded
+slightly lower TPRs in the larger two trees). \prankc{} showed a
+greater ability than ClustalW to maintain a high TPR at higher indel
+rates, especially in the 17-taxon and 44-taxon trees. At lower indel
+rates, the TPR performance of both aligners and the true alignment
+were qualitatively similar.
 
-\prankc{} and ClustalW both showed a similar tendency towards elevated
-FPRs in the 6-taxon tree (Figure 4B), but their behavior diverged
-significantly in the 17-taxon and 44-taxon trees. In the 17-taxon
-tree, \prankc{} only showed an elevated FPR at very high indel rates
-and divergence levels, but the ClustalW FPR consistently increased
-with indel rate, quadrupling in value from the lowest to highest indel
-rate. Interestingly, for any given indel rate, the ClustalW FPR showed
-little variation across the range of divergence levels. This result
-was counter-intuitive, as we expected alignment errors to become more
-common as divergence increases and the number of observed indel events
-grows. Furthermore, \prankc{} behaved according to our expectations,
-with a trend towards increased FPRs at higher divergences in the
-17-taxon tree and the highest indel rate. The FPR results in the
-44-taxon tree confirmed the strange effect of ClustalW’s alignments on
-the sitewise FPR: at the highest indel rates, ClustalW showed a
-negative correlation between FPR and divergence -- exactly opposite to
-the trend we expected.
+\prankc{} and ClustalW both showed a qualitatively similar pattern of
+elevated FPRs in the 6-taxon tree (Figure \ref{fig_4}B), but their
+behavior diverged significantly in the 17-taxon and 44-taxon trees. In
+the 17-taxon tree, \prankc{} only showed an elevated FPR compared to
+the true alignment at very high indel rates and divergence levels, but
+the ClustalW FPR increased steadily with the indel rate, quadrupling
+in value from the lowest to highest indel rate. Interestingly, for any
+given indel rate, the ClustalW FPR showed little variation across the
+range of divergence levels. This result was counter-intuitive, as we
+expected alignment errors to become more common as divergence
+increased and the number of observed indel events grew. Furthermore,
+\prankc{} behaved according to our expectations, showing increased
+FPRs only at the highest divergences and indel rates in the 17-taxon
+tree. The FPR results in the 44-taxon tree confirmed the strange
+effect of ClustalW’s alignments on the \sw FPR: at the highest indel
+rates, ClustalW showed a negative relationship between FPR and
+divergence---exactly opposite to the trend we expected. \prankc{}'s
+FPR in the 44-taxon tree was equal to or below that of the true
+alignment under almost all conditions.
 
-Finally, the error-controlled \tpr{} results (Figure 4C) provided a
-comprehensive map of the effect of alignment error on detecting
-sitewise positive selection. The two aligners not shown in the two
-other panels, MAFFT and \pranka{}, exhibited \tpr{} values
-intermediate to those from ClustalW and \prankc{} across the range of
-parameters tested. At indel rates above 0.1, all aligners yielded low
-\tpr{} values in the 6-taxon tree. In the larger two trees, excessive
-false positives introduced by ClustalW and MAFFT alignments precluded
-them from achieving high \tpr{} at high indel rates, while \pranka{}
-and \prankc{} showed good \tpr{} levels at all but the highest
-divergences and indel rates.
+The error-controlled \tpr results (Figure \ref{fig_4}C) provided a
+comprehensive map of the effect of alignment error on the detection of
+\sw positive selection. The two aligners not shown in the two other
+panels, MAFFT and \pranka, exhibited \tpr values intermediate to those
+from ClustalW and \prankc across the range of parameters tested. As
+expected, performance was very similar between aligners at very low
+indel rates. At higher indel rates, most aligners yielded similar
+patterns of low \tpr in the 6-taxon tree, but in the larger two trees
+only ClustalW and MAFFT alignments were unable to achieve high \tpr
+values, presumably due largely to their elevated FPR in those trees.
 
-It is worth noting the exceptional ability of PRANK to maintain a low
-level of false positive sites even under extremely difficult alignment
-conditions. Although \prankc{} showed slightly elevated FPRs at high
-indel rates in the 17-taxon tree, FPRs were nearly identical to the
-true alignment across all simulated conditions in the 44-taxon
+It is worth noting the exceptional ability of \prankc to maintain a
+very low level of false positive sites even under extremely difficult
+alignment conditions. Although \prankc showed slightly elevated FPRs
+at high indel rates in the 17-taxon tree, FPRs were nearly identical
+to the true alignment across all simulated conditions in the 44-taxon
 tree. This impressive performance suggests that, given a large enough
-number of taxa, PRANK alignments would yield very few erroneous false
-positives in scans for positive selection in sequences with divergence
-levels well beyond those found in commonly analyzed vertebrate, insect
-and fungal clades (Table ST1). Furthermore, these results showed that
-false negatives contributed more than false positives to PRANK’s
-reduction in sitewise performance --- a novel observation which
-provides insight into the nature of PRANK alignments and their
-application to sitewise evolutionary analysis.
+number of taxa, \prankc alignments would yield very few erroneous
+false positives in scans for positive selection in sequences with very
+high divergence levels. Furthermore, these results showed that false
+negatives contributed more than false positives to \prankc{}'s
+reduction in \sw performance---a novel observation which provides
+insight into the nature of \prankc alignments and their application to
+\sw evolutionary analysis.
 
-\subsection*{Effect of Alignment Filtering on Sitewise Error Rates}
+\subsection*{Effect of Alignment Filtering on \Sw Error Rates}
 
@@
 \begin{center}
-\includegraphics[scale=0.8]{fig5.pdf}
+\includegraphics[scale=0.65]{fig5.pdf}
 \end{center}
-\caption{}
+\caption{Sequences were simulated using the 17-taxon tree and a range
+  of indel rates and mean path length (MPL) divergence
+  levels. Alignments were inferred using (a) ClustalW or (b) \prankc,
+  either left unfiltered or filtered with one of four alignment
+  filters, and analyzed with SLR; true alignments were left unfiltered
+  and separately analyzed with SLR. One hundred fifty replicates were
+  simulated for each set of conditions. Squares are colored according
+  to the ratio of the performance of the indicated filter to the
+  performance of the unfiltered ClustalW or \prankc alignment as
+  measured by one of four summary statistics. In columns from left to
+  right: the number of unfiltered (i.e., non-N) residues remaining
+  in the alignment; the false positive rate (FPR) at the recommended
+  cutoff threshold; the true positive rate (TPR) at the recommended
+  cutoff threshold; the TPR at a 1\% FPR threshold (\tpr). Note that the
+  maximum percentage of filtered residues was capped at 50\% for all
+  methods except Gblocks.}
 \label{fig_5}
@@
 
-Having established that alignment error can lead to reduced sitewise
+Having established that alignment error can lead to reduced \sw
 performance through the introduction of false negatives and false
-positives we tested whether a number of alignment filtering methods
-could reduce error rates and improve the power of sitewise detection
-of positive selection. Using sequences simulated from the 17-taxon
-tree and the same range of indel rates and divergence levels as in the
-previous experiments, we calculated inferred alignments using ClustalW
-and \prankc{} and applied four filtering methods to the alignments
-before performing the sitewise analysis. Since we wished to determine
-whether alignment filters either improved or worsened the error rate
-and power of sitewise analysis, we measured the ratio of each
-performance measure to the value obtained in the equivalent unfiltered
-dataset. These relative scores are presented in Figure 5.
+positives, we tested whether a number of alignment filtering methods
+could reduce error rates and improve the power of \sw detection of
+positive selection. Using sequences simulated from the 17-taxon tree
+and a range of indel rates and divergence levels, we calculated
+inferred alignments using ClustalW and \prankc and applied four
+filtering methods before performing the \sw analysis. Since we wished
+to determine whether alignment filters either improved or worsened the
+error rates and power of \sw analysis, we measured the ratio of each
+performance measure to the value obtained from the equivalent
+unfiltered alignments. These relative values are presented in Figure
+\ref{fig_5}.
 
-We first looked at the two controls included in this set of
-experiments, the unfiltered true alignment and the inferred alignments
-filtered with the optimal filter (Figure 5, top two rows). The true
-alignment always showed reduced FPR and increased TPR, and \tpr{}
-compared to the inferred alignments, generally with a with greater
-magnitude of change compared to the ClustalW alignments than to the
-\prankc{} alignments. These relative scores represented the direction
-and magnitude of change that might be achieved by an effective
-alignment filter. Our hypothesis was that the optimal filtering method
-would show the same direction of change in FPR and TPR as the true
-alignment. When applied to ClustalW alignments (Figure 5a), the
-optimal filter indeed showed a clear pattern of improved sitewise
-performance across the majority of simulation conditions, with reduced
-FPR and elevated TPR and \tpr{}. At the highest indel rates and
-divergence levels, however, little change was observed. This may have
-been due to the combination of our filtering limit of 20\% masked
-residues and the highly inaccurate nature of ClustalW alignments under
-those conditions, as evidenced by the low TPRs and high FPRs in the
-previous set of simulations (Figure 4A,B). When applied to \prankc{}
-alignments, the optimal filter showed a similar reduction of FPR
-levels across the parameter space, but the TPR and \tpr{} results were
-mixed. At lower indel rates and divergence levels the TPR and \tpr{}
-were reduced, while at higher indel rates and divergences both the TPR
-and \tpr{} were slightly increased. The beneficial impact on power
-levels under the more difficult alignment conditions sugested that the
-optimal filter was able to successfully remove misaligned residues
-that were contributing to false negatives in the \prankc{}
-alignments. We suspect that the worse performance under easier
-alignment conditions was due to the small number of false positives
-and false negatives in those \prankc{} alignments, causing any amount
-of filtering to result in reduced performance.
-
-When interpreting the results from filtered alignments, we must keep
-in mind that alignment filters act through removing alignment residues
-or columns, resulting in a certain amount of reduction in the FPR and
-TPR purely from the removal of information from the alignment. For
-example, a filter that randomly removed a fraction of residues of each
-alignment would be expected to produce equal reductions in FPR and TPR
-levels. An effective filter that removes false positives may also
-yield a reduced TPR, but the FPR reduction would be larger in
-magnitude, making the detection of positive selection more powerful at
-a given error rate. Thus, a reduced FPR is not necessarily indicative
-of good filtering performance, nor is a reduced TPR necessarily
-indicative of poor filtering performance. Furthermore, the prevalence
-of false negatives resulting from misalignment suggests that alignment
-filters may also improve power by removing false negatives, perhaps by
-masking out misaligned residues that were preventing positive sites
-from being identified. The removal of false negatives would result in
-an increased TPR, further complicating the assessment of filtering
-results based on FPR or TPR alone. As a result, we preferred the use
-of the change in error-controlled \tpr{} when evaluating whether a
-filter had successfully improved the sitewise power of a dataset,
+As alignment filters act through the removal of alignment residues or
+columns, a certain amount of reduction in the FPR and TPR was expected
+purely from the decreased amount of information available. For
+example, a filter that randomly removes a fraction of residues of each
+alignment would be expected to produce equal reductions in FPR and
+TPR. A more effective filter may also yield a reduced TPR, but the FPR
+reduction would be larger in magnitude, making the detection of
+positive selection more powerful for a given error rate. Thus, a
+reduced FPR is not necessarily indicative of good filtering
+performance, nor is a reduced TPR necessarily indicative of poor
+filtering performance. Additionally, the prevalence of false negatives
+resulting from misalignment suggested that alignment filters may also
+improve power by removing false negatives, perhaps by masking out
+residues that were preventing positive sites from being
+identified. The removal of false negatives would result in an
+increased TPR, further complicating the assessment of filtering
+results based on FPR or TPR alone. As a result, we focused on the
+change in error-controlled \tpr{} as the best single measure of
+whether a filter had successfully improved the \sw power of a dataset
 since this value is sensitive to changes in both the FPR and TPR.
 
-Using the change in \tpr{} as a measure of filtering performance, we
-found T-Coffee and GUIDANCE to be highly effective at improving
-ClustalW alignments, with magnitudes of improvement near to those from
-the optimal filter. Interestingly, GUIDANCE showed increased
-performance at the highest indel and divergence levels, where the
-optimal filter had no effect. As with the optimal filter, results were
-mixed when applying these two filters to the \prankc{} alignments: at
-lower indel rates and divergences both filters resulted in reduced
-performance, while T-Coffee achieved modest improvements while
-GUIDANCE showed no change under more difficult conditions.
+We first examined the two controls, the unfiltered true alignment and
+the inferred alignments filtered with the optimal filter (top two
+rows, Figure \ref{fig_5}). The true alignment nearly always showed
+reduced FPR and increased TPR and \tpr compared to the inferred
+alignments, with a greater magnitude of change relative to the
+ClustalW alignments than to the \prankc alignments. These scores
+represented the direction and an upper limit on the magnitude of
+change that might be achieved by a perfect alignment filter. One
+exception to the general trend was the observation of two data points
+with slightly elevated FPR compared to \prankc alignments at the
+lowest indel rate. This small incosistency may be explained by
+stochastic variation in false positive counts, as the absolute value
+of the FPR was very low in both datasets under those conditions (on
+the order of $5\times10^{-4}$, Figure \ref{fig_4}B). A similar slight FPR
+elevation was also observed at the same indel rate for the optimal and
+GUIDANCE filters.
+
+Our hypothesis was that the optimal filter would show the same
+direction of change in FPR and TPR as the true alignment, but with
+slightly lower magnitudes. Indeed, improved \sw performance was
+achieved in nearly all simulation conditions by the optimal filter,
+with the magnitude of \tpr change slightly lower than for true
+alignment. For ClustalW alignments the amount of improvement was quite
+large, with $>$70\% increase in \tpr for nearly all conditions with an
+indel rate above 0.1. The improvement was more modest for \prankc
+alignments with a maximum of 15--35\% \tpr increase.
+
+Looking at the reduction in the number of non-masked residues
+remaining after filtering, we found that the optimal filter reached
+the maximum of 50\% filtered residues for all ClustalW alignments with
+MPL$>$1 and an indel rate$>$0.1. This meant that more than 50\% of
+residues were correctly aligned across less than 50\% of the tree in
+those alignments. By contrast, the optimal filter applied to \prankc
+alignments only reached the maximum of 50\% filtered residues at the
+highest tested divergence level and indel rate combination.
+
+The TPR improvements achieved by the optimal filter provided some
+insight into the nature of sitewise false negatives resulting from
+alignment error. Two different types of alignment error might cause a
+false negative at a positively-selected site: either misalignment of
+one or more nonhomologous codons causing the positive signal to be
+masked, or non-alignment of homologous codons causing the amount of
+evolutionary information to be reduced. The former type of error would
+be recoverable by alignment filtering (through removal of the codon(s)
+masking the positive signal), but the latter would not. Thus, the
+ability of the optimal filter to improve TPR levels across the board
+provided evidence that a sizeable portion of false negatives from both
+ClustalW and \prankc alignments were due to misaligned codons and thus
+amenable to recovery by filtering. Although the optimal filter was
+unrealistic in that it was based on perfect knowledge of which codons
+were misaligned, this result provided hope that one of the other
+filters might show a similar ability to recover false negative errors
+from \prankc alignments.
+
+Turning to the three filters under investigation, we found T-Coffee
+and GUIDANCE both to be highly effective at improving ClustalW
+alignments, with magnitudes of improvement near those of the optimal
+filter. When applied to \prankc alignments, however, the two filters'
+behavior diverged: T-Coffee only showed unchanged or reduced \tpr, but
+GUIDANCE yielded slightly improved \tpr at high divergence levels and
+indel rates, with values 5--15\% greater than the unfiltered \prankc
+alignments. Both filters removed similar amounts of sequence
+information and resulted in similarly reduced FPR levels, but GUIDANCE
+showed a unique ability to recover false negatives from \prankc
+alignments at the highest divergence levels and indel rates, and the
+resulting TPR elevation appears to have been responsible for the
+increased \tpr performance.
 
 Gblocks behaved very differently from the other filters tested,
-resulting in reduced FPR, TPR, and \tpr{} values under nearly all
-simulation conditions. Only at high indel rates and low divergence
-levels in the ClustalW alignments did Gblocks show an increased \tpr{}
-relative to the unfiltered alignments. The poor performance of Gblocks
-was likely due to overly-aggressive removal of alignment
-columns. Since the program only identifies conserved blocks as opposed
-to scoring residues or alignment columns, we could not cap the amount
-of masked sequence as was done for the residue-wise filters. Even
-though a relaxed setting was used which allowed conserved blocks to
-contain gapped regions (see Methods), the dramatically reduced TPR
-levels indicate that much of the power to detect positive selection
-was eliminated by Gblocks filtering. Dessimoz et
-al. \citeyearpar{Dessimoz2010Phylogenetic} found a similar result for
-the effect of Gblocks on the accuracy of phylogenetic inference,
+resulting in reduced FPR, TPR, and \tpr under nearly all simulation
+conditions. Only at high indel rates and low divergence levels in the
+ClustalW alignments did Gblocks show increased \tpr relative to the
+unfiltered alignments. This poor performance was likely due to
+overly-aggressive removal of alignment columns. We could not limit the
+amount of sequence masked by Gblocks, so many alignments saw more than
+70\% of residues removed, resulting in the loss of a large number of
+correctly-aligned true positive sites. Dessimoz et
+al. \citeyearpar{Dessimoz2010Phylogenetic} found Gblocks filtering to
+have a negative effetc on the accuracy of phylogenetic inference; our
+results provide additional evidence in support of their finding,
 suggesting that Gblocks filtering tends to reduce, rather than
-increase, the power and accuracy of a number of evolutionary analyses.
+increase, the power and accuracy of alignments when applied to a
+number of evolutionary analyses.
+
+There was some evidence that the column-wise nature of Gblocks
+filtering was partly responsible for its poor performance in our
+experiments. Since false negative errors cannot be recovered through
+the removal of entire alignment columns, it made sense that the
+\prankc alignments---which resulted in varying numbers of false
+negatives but always very few false positives---would not see improved
+performance after column-wise filtering. On the other hand, ClustalW
+alignments showed a relatively constant level of false positives and
+an increasing number of false negatives as divergence levels
+increased. The application of a column-wise filter like Gblocks would
+thus be expected to show good improvement at low divergences where
+false positives dominated, but less improvement at higher divergences
+where false negatives became more prominent. Indeed, this was the
+pattern observed when applying Gblocks to ClustalW alignments.
 
 Overall, our alignment filtering simulations found that filtering
-methods based on GUIDANCE and T-Coffee scores showed a good ability to
-remove false positives and false negatives from ClustalW alignments,
-with levels of improvement on par with those obtained with the optimal
-filter. This was the case across a wide range of divergence levels and
-indel rates, suggesting that filtering ClustalW alignments with
-T-Coffee or GUIDANCE can usually be expected to result in improved
-sitewise performance. Alignment filtering proved less helpful when
-applied to \prankc{} alignments, with all filters reducing the
-sitewise performance except under the most divergent and indel-rich
-conditions. Importantly, alignment filtering only showed improved
-performance at divergence levels well above those found in
-commonly-analyzed groups of species (e.g., MPL $>$ 1.6). This leads us
-to the conclusion that, based on the sample of aligners and filters
-tested here, the unfiltered \prankc{} alignments would yield the best
-performance for detecting sitewise positive selection when analyzing
-protein-coding sequences across a wide range of indel rates and
-commonly encountered sequence divergence levels.
+methods based on GUIDANCE and T-Coffee scores had the ability to mask
+away many residues causing false positives and false negatives from
+ClustalW alignments, with high levels of improvement across a wide
+range of divergence levels and indel rates. Filtering was less
+beneficial when applied to \prankc alignments, with T-Coffee reducing
+performance and GUIDANCE yielding only mild \tpr
+improvements. Importantly, GUIDANCE only showed improved performance
+at high divergence levels (e.g., MPL$>$1.6) well above those found in
+commonly-analyzed groups of species. Thus, the use of unfiltered
+\prankc alignments would yield the best performance for detecting \sw
+positive selection when analyzing protein-coding sequences across a
+wide range of indel rates and commonly encountered sequence divergence
+levels.
 
@@
 
-Sequence alignment is important yet under-studied]
+In this paper, we investigated the performance of sitewise detection
+of positive selection under a range of tree sizes, indel rates, and
+divergence levels, using simulation parameters designed to approximate
+mammalian protein-coding genes. We evaluated the ability of four
+alignment methods and three alignment filtering methods to accurately
+reconstruct the codon-level homology for sitewise inference, using the
+FPR, TPR, and error-controlled \tpr of the sitewise detection of
+positive selection as our main measures of performance.
 
-Our simulations showed that alignment error has significant impact on sitewise detection of pos-sel. ClustalW worst, \prankc{} best.
+The simulation results showed that alignment error can have a
+measurable impact on the error rates and power of the sitewise
+detection of positive selection under all but the least difficult
+alignment conditions. We confirmed and extended the findings of
+Fletcher and Yang \citeyearpar{Fletcher2010Effect} regarding the
+relative accuracy of different aligners, showing that \prankc had the
+best performance and ClustalW had the worst performance for subsequent
+sitewise analysis. Notably, our simulations found that ClustalW
+produced more sitewise false positives than MAFFT or \prankc even at
+low divergence levels, suggesting that its use should be avoided even
+when analyzing closely-related sequences. \prankc, on the other hand,
+resulted in very low FPRs even at very high divergences. In
+particular, when the number of sequences was large, its sitewise FPRs
+were virtually indistinguishable from the true alignment.
 
-Filtering improves low-quality alignments and high-quality alignments in extreme circumstances, but overall unfiltered \prankc{} is still the best.
+The very low FPRs observed for \prankc alignments conflicted somewhat
+with the results of Fletcher and Yang
+\citeyearpar{Fletcher2010Effect}, who found that the FPRs for the
+branch-site test were not under control even with \prankc
+alignments. This apparent discrepancy can be explained by different
+sensitivities to alignment error: the branch-site test would yield
+false positives when misalignment causes apparent positive selection
+along only the foreground branch, while the SLR test would produce
+false positives only when misalignment causes a signal of positive
+selection strong enough to overpower the non-positive signal
+throughout the tree. This effect stems from the different biological
+hypotheses tested by the two methods; their differential sensitivity
+to misalignment underscores the necessity of considering the
+biological sensitivity and robustness to alignment error when applying
+either of these tests to detect positive selection within an
+alignment.
 
-Despite high FPRs from ClustalW, false positive rates are still very low... due mainly to conservative cutoff and prevalence of purifying selection in our simulation conditions
+Despite producing very low FPRs, \prankc alignments still resulted in
+an increased number of false negatives compared to the true
+alignment. We showed that some of these false negatives were possible
+to recover as true positives through alignment filtering, and we found
+that both the optimal filter and GUIDANCE were able to successfully
+recover these false negatives at high divergence levels, resulting in
+small but measurable performance improvements over the unfiltered
+\prankc alignments.
 
-(fig.~\ref{sample_fig_2}).
+Among the aligners and filters tested here, we can recommend the use
+of unfiltered \prankc alignments for the detection of sitewise
+positive selection in all but the most divergent phylogenies. GUIDANCE
+showed some ability to improve sitewise power under difficult
+alignment conditions at high divergence levels, but this improvement
+was achieved largely through the recovery of false negatives as
+opposed to the elimination of false positives. For an analysis where
+the control of false positives is paramount, the added computational
+expense of running many bootstrap alignment replicates (as is done
+GUIDANCE) may not be worth the possibility of a slight increase in
+power. The development of efficient and effective alignment filters,
+particularly with an eye towards recovering false negatives as
+identified in this study, will be an interesting area for future
+research.
+
+Our simulations did not include biologically realistic models of
+spatial or temporal variability in the rate of indel formation or in
+the distribution of selective pressures \citep{Whelan2008Spatial}, but
+we do not expect that such heterogeneity would affect our main
+conclusions regarding the relative performance of different aligners
+and filters. The trends we observed were consistent across a wide
+range of parameter values, suggesting that they reflect fundamental
+differences in each method's ability to align or filter sequences as
+opposed to artifacts due to the relative simplicity of our
+simulations.
+
+However, the apropriateness of our simulation scheme should be
+critically considered when evaluating our specific power and error
+rate estimates in the context of real-world data analysis. Functional
+differences between genes may influence the $\omg$ distribution and
+indel rate---with some genes or domains showing lower or higher
+tolerance to indels and fewer or more neutrally-evolving sites than
+modeled in our simulations---making false positive results more or
+less likely. Second, species-level differences may also be important,
+as the efficacy of natural selection is highly dependent on effective
+population size \citep{Ellegren2009Selection}. For example, proteins
+evolving in \Dr species with a high effective population size should
+experience stronger positive and purifying selection than in mammals,
+potentially leading to increased power and reduced error rates when
+compared to our simulations based on a mammalian-like $\omg$
+distribution.
+
+We showed here that even relatively simple evolutionary simulation
+experiments could sensitively assess the performance characteristics
+of different aligners, provide quantitative insight into the practical
+effects of alignment error, and suggest areas for future development
+of alignment and filtering methods. In the future, we expect the
+development of more realistic simulations for protein
+evolution---perhaps incorporating structurally-motivated and
+empirically validated models of mutation, indel formation and
+constraint---to further increase the applicability and accuracy of
+such experiments. As genomes rapidly accumulate in the databases and
+large-scale analyses become the norm, we hope that the development and
+application of alignment methods, which are arguably the most
+important step in any evolutionary analysis, will be based on a
+rigorous understanding of their behavior and performance when applied
+to a wide variety of evolutionary analyses.
 
@@
 
-A.B.H. and P.V. acknowledge ``\'Ecole th\'ematique
-g\'enomique environnementale'' for helpful discussions. M. Bormans is
-acknowledged for comments on the manuscript. P.V. acknowledges GIS
-``G\'enomique marine'' and ``Fondation Total'' and CNRS--ECCO for funding.
-
-\bibliographystyle{mbe3}
-\fontsize{9}{10}\selectfont%
-\itemsep 0pt
+\bibliographystyle{mbe3} \fontsize{9}{10}\selectfont% \itemsep 0pt
 

### 2011-07-02 projects/slrsim/combined.plot.R
Combined plots update.

@@ -47,7 +47,7 @@ multi.methods <- function() {
   pdf(file="fig_zero.pdf", width=6, height=5)
   p <- ggplot(comb.df, aes(x=length, y=y.val, colour=analysis, linetype=analysis))
   p <- p + theme_bw()
-  p  <- p + geom_line(alpha=0.7)
+  p  <- p + geom_line(alpha=1, size=0.5)
   clrs <- c('gray10', 'gray10', 'gray10')
   line.types <- c('dashed', 'dotted', 'solid')
   p <- p + scale_colour_manual(name="Analysis Method", values=clrs)
@@ -56,7 +56,8 @@ multi.methods <- function() {
   p <- p + facet_grid(stat ~ tree)
   p <- p + coord_equal(ratio=2)
   p <- p + opts(
-    axis.text.x = theme_text(angle=90, hjust=1),
+    axis.text.x = theme_text(size=5, angle=90, hjust=1.2),
+    axis.text.y = theme_text(size=5, angle=0, hjust=1.2),
     axis.title.x = theme_blank(),
     axis.title.y = theme_blank(),
     strip.text.x = theme_blank(),
@@ -70,17 +71,21 @@ multi.methods <- function() {
 }
 
 multi.aligners <- function() {
-  tbl.a <- read.csv('~/scratch/gj1_fig_one_a/current/table.csv')
-  tbl.b <- read.csv('~/scratch/gj1_fig_one_b/current/table.csv')
-  tbl.c <- read.csv('~/scratch/gj1_fig_one_c/current/table.csv')
+  tbl.a <- read.csv('~/scratch/gj1_fig_one_a/current/table.csv', stringsAsFactors=F)
+  tbl.b <- read.csv('~/scratch/gj1_fig_one_b/current/table.csv', stringsAsFactors=F)
+  tbl.c <- read.csv('~/scratch/gj1_fig_one_c/current/table.csv', stringsAsFactors=F)
   
   tbl.a$tree <- 'artificial'
   tbl.b$tree <- 'bglobin'
   tbl.c$tree <- 'encode'
+
+  tbl.a <- subset(tbl.a, ins_rate == 0.05 | (ins_rate == 0 & aligner == 'none'))
+  tbl.b <- subset(tbl.b, ins_rate == 0.05 | (ins_rate == 0 & aligner == 'none'))
[…]

### 2011-07-18 projects/slrsim/paper/slrsim.tex
paper formatting: remove MBE style in favor of double-spaced ms format.

@@
-\documentclass{mbe}
+\documentclass{article}
 \usepackage{natbib,graphicx,dblfloatfix,multirow,array,threeparttable}
-\usepackage{xspace}
+\usepackage{xspace,bm,amsmath,amssymb}
+\usepackage[nolists,tablesfirst,nomarkers]{endfloat}
+
+\usepackage[a4paper]{geometry}
+
+\linespread{1.5}
+
 \setlength{\bibsep}{0.0pt}
@@
 \affiliation{European Bioinformatics Institute, Wellcome Trust Genome Campus, Hinxton, Cambridgeshire, CB10 1SD, United Kingdom}}
-\authorlist{Jordan and Goldman}
-\shorttitle{Effect of Alignment Error on Detecting Sitewise Selection}
-
-\mbeabstract{%
+%\authorlist{Jordan and Goldman}
+%\shorttitle{Effect of Alignment Error on Detecting Sitewise Selection}
+
+%\mbeabstract{% 
+\section*{Abstract}
 
@@
 
-}
-
-\keywords{Alignment, positive selection, alignment filtering, indels}
-
-\email{greg@ebi.ac.uk}
-
-\mbestyle
+%}
+
+%\keywords{Alignment, positive selection, alignment filtering, indels}
+
+%\email{greg@ebi.ac.uk}
+
+%\mbestyle
 
@@
 
-Large-scale comparative genomics has been widely employed to identify
-shared and unique biological features between organisms of interest to
-the biomedical sciences, with applications ranging from evolutionary
-annotation transfer \citep{Bork1998Predicting} to phylogenetics
-\citep{Murphy2007Using,Prasad2008Confirming} and population genetic
-inference \citep{Hurst2009Genetics}. The decreasing cost of DNA
-sequencing has triggered a striking increase in the number of model
-and non-model organisms with planned genome sequencing projects,
-suggesting that the range and scale of comparative genomics
-applications will continue to expand
-\citep{Green20072x,2007Identification}.
+The decreasing cost of DNA sequencing has triggered a striking
+increase in the number of model and non-model organisms with planned
+genome sequencing projects, suggesting that the range and scale of
+comparative genomics applications will continue to expand
+\citep{Green20072x,2007Identification}. These clusters of
+closely-related genome sequences across a wide taxonomic range have
+led to a better understanding of which aspects of molecular evolution
+are variable and which are constant \citep{Wolf2009Nonlinear} and an
+increased sampling of species should continue to boost the power and
+accuracy of individual analyses within a given clade.
 
@@
 particular has flourished as a result of the growth in comparative
-genomics datasets. Clusters of closely-related genome sequences across
-a wide taxonomic range have led to a better understanding of which
-aspects of molecular evolution are variable and which are constant
-\citep{Wolf2009Nonlinear}, and an increased sampling of species should
-continue to boost the power and accuracy of individual analyses within
-a given clade. This is especially beneficial for the calculation of
-spatially precise evolutionary estimates, as additional species
+genomics datasets. This is especially beneficial for the calculation
+of spatially precise evolutionary estimates, as additional species
 sampling has been shown to be an effective means of boosting the
@@
 structural features \citep{Lin2007Proportion,Ramsey2011Relationship},
-human population diversity \citep{2010Map}, and human disease
+human population diversity \citep{2010Map} and human disease
 mutations \citep{Arbiza2006Selective}.
 
-A major concern in the \sw detection of protein positive
-selection is that the effect of alignment error is not well
-characterized. Intuitively, one might expect alignment error to result
-mainly in an increased number of false positives, as the spurious
-alignment of nonhomologous codons on average would result in a high
-number of apparent non-synonymous substitutions and a low number of
-synonymous substitutions (since a randomly chosen pair of codons are
-more likely to be nonsynonymous than synonymous). However, false
-negatives may also be introduced, either through over-alignment of
-synonymous codons (increasing the synonymous substitution rate) or
-non-alignment of homologous codons (causing reduced power due to less
-evolutionary information). Since different aligners employ a variety
-of algorithms, evolutionary models, and heuristic optimizations
-\citep{Notredame2007Recent}, each program may be more or less prone to different
-types of alignment error, causing potentially large variations in the
-nature and magnitude of its impact on the detection of positive
-selection.
+A major concern in the detection of protein positive selection is that
+the effect of alignment error is not well characterized. Intuitively,
+one might expect alignment error to result mainly in an increased
+number of false positives, as the spurious alignment of nonhomologous
+codons on average would result in a high number of apparent
+non-synonymous substitutions and a low number of synonymous
+substitutions (since a randomly chosen pair of codons are more likely
+to be nonsynonymous than synonymous). However, false negatives may
+also be introduced, either through over-alignment of synonymous codons
+(increasing the synonymous substitution rate) or non-alignment of
+homologous codons (causing reduced power due to less evolutionary
+information). Since different aligners employ a variety of algorithms,
+evolutionary models, and heuristic optimizations
+\citep{Notredame2007Recent}, each program may be more or less prone to
+different types of alignment error, causing potentially large
+variations in the nature and magnitude of its impact on the detection
+of positive selection.
 
@@
 be less susceptible to alignment error---than less structured
-regions. Furthermore, the evolutionary divergence of a dataset affects
-the power of \sw inference and the prevalence of indels in different
-ways. Because maximum-likelihood methods for detecting positive
-selection require data in the form of observed substitution events,
-they show little power at low divergence and their highest power at
-intermediate to high divergence levels
-\citep{Anisimova2001Accuracy}. On the other hand, the number of indels
-in a dataset should increase with divergence. These two trends suggest
-that the overall power will be low at both extremes of divergence,
-with little inference power at low divergence (due to the scarcity of
-data in the form of observed substitutions) and an overwhelming amount
-of alignment error at high divergence (due to the large number of
-indel events). It is important to note, however, that the majority of
-genes in many biological clades of interest (such as mammals,
-vertebrates, fruit flies, and yeast) fall within the middle range of
-divergences where \sw methods are at their most powerful and where
-multiple alignment is a difficult---but not hopeless---problem. As
-such, it is important to seek an understanding of the impact of
-alignment error on overall error rates within this important range of
-divergence levels.
+regions.
+
+Furthermore, the evolutionary divergence of a dataset affects the
+power of \sw inference and the prevalence of indels in different
+ways. As maximum-likelihood methods for detecting positive selection
+require data in the form of observed substitution events, they show
+little power at low divergence and their highest power at intermediate
+to high divergence levels \citep{Anisimova2001Accuracy}. However,
+alignment error should be greatest at high divergences, which may have
+the effect of reducing power. These two trends suggest that the
+overall power will be low at both extremes of divergence, with little
+inference power at low divergence (due to the scarcity of data in the
+form of observed substitutions) and an overwhelming amount of
+alignment error at high divergence (due to the large number of indel
+events).
+
+Luckily the majority of genes in many biological clades of interest
+(such as mammals, vertebrates, fruit flies, and yeast) fall within the
+middle range of divergences where \sw methods are at their most
+powerful and where multiple alignment is a difficult---but not
+hopeless---problem. As such, it is important to seek an understanding
+of the impact of alignment error on overall error rates within this
+important range of divergence levels.
 
@@
 sites. Markova-Raina and Petrov \citeyearpar{Markova-Raina2011High}
-used the expected general effect of alignment error (an increase in
+used their expected general effect of alignment error (an increase in
 false positives due to misalignment of non-homologous codons) as a
@@
 analysis. However, the absolute number of false positives remained
-uncertain due to the possible conflation of multiple sources of error:
-in addition to alignment error, the authors noted that gene
+uncertain and there was the possibility of conflating multiple sources
+of error: in addition to alignment error, the authors noted that gene
 mis-annotation was responsible for many apparent false positives, and
@@
 questions regarding the nature of alignment error and its quantitative
-impact on the detection of positive selection left unanswered.
+impact on the detection of positive selection left unanswered by
+empirical studies.
 
@@
 investigating error rates in detail, allowing one to pinpoint the
-source of errors in multi-step analyses such as alignment followed by
-evolutionary inference. This approach has widely been employed in
-assessing the robustness of phylogenetic inference methods to
-misalignment
+sources of error in multi-step analyses such as alignment followed by
+evolutionary inference. This approach has been employed in assessing
+the robustness of phylogenetic inference methods to misalignment
 \citep{Dwivedi2009Phylogenetic,Ogden2006Multiple,Loytynoja2008PhylogenyAware}
@@
 \citeyearpar{Fletcher2010Effect} performed a series of simulation
-experiments investigating alignment error in the use of branch-site
-models to detect positive selection in genes. Their results showed
-that most aligners caused false positives by over-aligning codons and
-that datasets from mammalian and vertebrate gene families contain
-enough evolutionary divergence to make false positive errors resulting
-from misalignment a legitimate concern.
+experiments investigating alignment error in the use of the
+branch-site test to detect positive selection in genes. Their results
+showed that most aligners caused false positives by over-aligning
+codons and that datasets from mammalian and vertebrate gene families
+contain enough evolutionary divergence to make false positive errors
+resulting from misalignment a legitimate concern.
 
@@
 regions have been commonly used in phylogenetic and molecular
-evolutionary analyses. The popular Gblocks applies a set of heuristic
-criteria to identify conserved blocks deemed suitable for phylogenetic
-or evolutionary analysis \citep{Castresana2000Selection} while a
-number of aligners such as T-Coffee \citep{Notredame2000TCoffee} and
-PRANK \citep{Loytynoja2005From} produce estimates of alignment
-confidence or reliability. Alignment confidence scores based on the
-robustness of alignments to guide tree perturbations
-\citep{Penn2010Alignment} or the sampling of suboptimal alignments
-\citep{Kim2011PSAR} have also been proposed. Unfortunately, despite
-their widespread use, the impact of the many available filtering
-methods on phylogenetic and evolutionary analyses has not been well
-studied. Even for a single filtering program, Gblocks, results have
-been contradictory: one simulation-based study found that it improved
-the phylogenetic signal \citep{Talavera2007Improvement} while an
-empirical study across a wide range of taxa found that
+evolutionary analyses. The popular Gblocks program applies a set of
+heuristic criteria to identify conserved blocks deemed suitable for
+phylogenetic or evolutionary analysis \citep{Castresana2000Selection}
+while a number of aligners such as T-Coffee
+\citep{Notredame2000TCoffee} and PRANK \citep{Loytynoja2005From}
+produce estimates of alignment confidence or reliability. Alignment
+confidence scores based on the robustness of alignments to guide tree
+perturbations \citep{Penn2010Alignment} or the sampling of suboptimal
+alignments \citep{Kim2011PSAR} have also been proposed. Unfortunately,
+despite their widespread use, the impact of the many available
+filtering methods on phylogenetic and evolutionary analyses has not
+been well studied. Even for a single filtering program, Gblocks,
+results have been contradictory: one simulation-based study found that
+it improved the phylogenetic signal \citep{Talavera2007Improvement}
+while an empirical study across a wide range of taxa found that
 Gblocks-filtered alignments produced worse phylogenetic trees than
@@
 \citep{Studer2008Pervasive,Aguileta2009Rapidly}, an analysis of the
-benefit---if any---of alignment filtering to the detection of positive
-selection would seem well-warranted.
+benefits of alignment filtering to the detection of positive selection
+would seem well-warranted.
 
@@
 \begin{center}
-\includegraphics[scale=0.5]{fig1.pdf}
+%\includegraphics[scale=0.5]{fig1.pdf}
+\includegraphics[scale=0.8]{fig1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[scale=0.75]{fig2.pdf}
+%\includegraphics[scale=0.75]{fig2.pdf}
+\includegraphics[scale=1.25]{fig2.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[scale=0.75]{fig3.pdf}
+%\includegraphics[scale=0.75]{fig3.pdf}
+\includegraphics[scale=1.25]{fig3.pdf}
 \end{center}
@@
 \begin{center}
+%\includegraphics[scale=0.8]{fig4.pdf}
 \includegraphics[scale=0.8]{fig4.pdf}
@@
 \begin{center}
+%\includegraphics[scale=0.65]{fig5.pdf}
 \includegraphics[scale=0.65]{fig5.pdf}

### 2011-07-19 projects/slrsim/paper/slrsim.tex
Manuscript revs

@@
 \usepackage[nolists,tablesfirst,nomarkers]{endfloat}
-
 \usepackage[a4paper]{geometry}
+\usepackage{color}
+
+\definecolor{red}{rgb}{1,0,0}
 
@@
 
+\newcommand{\mtwo}{PAML M2a\xspace}
+\newcommand{\meight}{PAML M8\xspace}
 \newcommand{\pranka}{PRANK$_{\textrm{AA}}$\xspace}
@@
 and the ability of alignment filters to improve performance. We found
-that some aligners produced many false positive results across a wide
-range of divergence levels, while others produced very few. False
+that some aligners led to many false positive results across a wide
+range of divergence levels, while others resulted in very few. False
 negatives were a problem for all aligners, increasing with
@@
 reduce the prevalence of false negatives, especially at higher
-divergence levels and indel rates. We conclude that for the sitewise
-evolutionary analysis of most vertebrate, insect and fungal
-phylogenies, the use of unfiltered PRANK alignments is recommended.
+divergence levels and indel rates. \textcolor{red}{As no combination
+  of aligner plus filter significantly outperformed the best aligner
+  alone,} we conclude that for the sitewise evolutionary analysis of
+most vertebrate, insect and fungal phylogenies, the use of unfiltered
+PRANK alignments is recommended.
 
@@
 Furthermore, the evolutionary divergence of a dataset affects the
-power of \sw inference and the prevalence of indels in different
+power of \sw inference and the prevalence of indels in multiple
 ways. As maximum-likelihood methods for detecting positive selection
@@
 downstream evolutionary analyses such as phylogeny inference
-\citep{Wong2008Alignment} et and estimates of positive selection
-\citep{Schneider2009Estimates,Markova-Raina2011High} . Most recently,
+\citep{Wong2008Alignment} and estimates of positive selection
+\citep{Schneider2009Estimates,Markova-Raina2011High}. Most recently,
 Markova-Raina and Petrov \citeyearpar{Markova-Raina2011High} showed
@@
 genomes is highly sensitive to aligner choice, with PRANK's codon
-model consistently producing alignments with the lowest amount of
-positive selection. Still, according to the authors' manual inspection
-of alignments, even positively-selected sites identified with PRANK
-alignments contained a sizable proportion of apparent false positives.
-
-A limitation to the analysis of error in empirical datasets is the
+model \citep{Loytynoja2008PhylogenyAware} consistently producing
+alignments with the lowest amount of positive selection. Still,
+according to the authors' manual inspection of alignments, even
+positively-selected sites identified with PRANK alignments contained a
+sizable proportion of apparent false positives.
+
+A limitation of the analysis of error in empirical datasets is the
 lack of a benchmark set of true alignments and positively-selected
@@
 method itself. This limitation leaves important and interesting
-questions regarding the nature of alignment error and its quantitative
-impact on the detection of positive selection left unanswered by
-empirical studies.
+questions, regarding the nature of alignment error and its
+quantitative impact on the detection of positive selection unanswered
+by empirical studies.
 
@@
 while a number of aligners such as T-Coffee
-\citep{Notredame2000TCoffee} and PRANK \citep{Loytynoja2005From}
-produce estimates of alignment confidence or reliability. Alignment
-confidence scores based on the robustness of alignments to guide tree
-perturbations \citep{Penn2010Alignment} or the sampling of suboptimal
-alignments \citep{Kim2011PSAR} have also been proposed. Unfortunately,
-despite their widespread use, the impact of the many available
-filtering methods on phylogenetic and evolutionary analyses has not
-been well studied. Even for a single filtering program, Gblocks,
-results have been contradictory: one simulation-based study found that
-it improved the phylogenetic signal \citep{Talavera2007Improvement}
-while an empirical study across a wide range of taxa found that
-Gblocks-filtered alignments produced worse phylogenetic trees than
-unfiltered alignments \citep{Dessimoz2010Phylogenetic}. With the
-application of published filtering methods to alignments before
-testing for positive selection becoming standard practice
+\citep{Notredame2000TCoffee} and PRANK
+\citep{Loytynoja2008PhylogenyAware} produce estimates of alignment
+confidence or reliability. \textcolor{red}{GUIDANCE, which measures
+  the robustness of alignment regions to perturbations in the guide
+  tree used for progressive alignment, has also been proposed as an
+  alignment confidence score
+  \citep{Penn2010Alignment}}. Unfortunately, despite their widespread
+use, the impact of the many available \textcolor{red}{alignment
+  scoring and filtering} methods on phylogenetic and evolutionary
+analyses has not been well studied. Even for a single filtering
+program, Gblocks, results have been contradictory: one
+simulation-based study found that it improved the phylogenetic signal
+\citep{Talavera2007Improvement} while an empirical study across a wide
+range of taxa found that Gblocks-filtered alignments produced worse
+phylogenetic trees than unfiltered alignments
+\citep{Dessimoz2010Phylogenetic}. With the application of published
+filtering methods to alignments before testing for positive selection
+becoming standard practice
 \citep{Studer2008Pervasive,Aguileta2009Rapidly}, an analysis of the
@@
 error and alignment filtering into estimates of the error rate and
-power of \sw evolutionary estimates. Our approach is similar to that
-of Anisimova and Yang \citeyearpar{Anisimova2002Accuracy} in that we
-use simulated protein alignments to evaluate methods for detecting \sw
-positive selection, considering each site to be an independent
-hypothesis test producing a positive or negative result. Our
-experimental design also shares some features with Fletcher and Yang's
-\citeyearpar{Fletcher2010Effect} investigation of the performance of
-branch-site method for detecting genewise positive selection under
-alignment error. Like Fletcher and Yang
-\citeyearpar{Fletcher2010Effect}, we simulate codon alignments with
-indels and calculate inferred alignments using a small but diverse
-sample of aligners. Unlike their study, here we focus on \sw detection
-of positive selection occurring throughout a phylogeny (as opposed to
-genewise detection of selection acting at specific branches), we
-explore a wider range of plausible tree sizes, indel rates and
-divergence levels, and we evaluate the impact of a number of alignment
-filtering methods on the \sw analysis. Given the recent proliferation
-of completed and planned vertebrate genomes, we pay special attention
-to choosing simulation parameters and analysis methods similar to
-those commonly encountered in the \sw analysis of vertebrate gene
-families.
+power of \sw evolutionary inference of positive selection. Our
+approach is similar to that of Anisimova et al.
+\citeyearpar{Anisimova2002Accuracy} in that we use simulated protein
+alignments to evaluate methods for detecting \sw positive selection,
+considering each site to be an independent hypothesis test producing a
+positive or negative result. Our experimental design also shares some
+features with Fletcher and Yang's \citeyearpar{Fletcher2010Effect}
+investigation of the performance of branch-site method for detecting
+genewise positive selection under alignment error. Like Fletcher and
+Yang \citeyearpar{Fletcher2010Effect}, we simulate codon alignments
+with indels and calculate inferred alignments using a small but
+diverse sample of aligners. Unlike their study, here we focus on \sw
+detection of positive selection occurring throughout a phylogeny (as
+opposed to genewise detection of selection acting at specific
+branches), we explore a wider range of plausible tree sizes, indel
+rates and divergence levels, and we evaluate the impact of a number of
+alignment filtering methods on the \sw analysis. Given the recent
+proliferation of completed and planned vertebrate genomes, we pay
+special attention to choosing simulation parameters and analysis
+methods similar to those commonly encountered in the \sw analysis of
+vertebrate gene families.
 
@@
 the simulation of protein-coding DNA alignments: the artificial
-6-taxon tree used by Anisimova and Yang
+6-taxon tree used by Anisimova et al.
 \citeyearpar{Anisimova2001Accuracy} and Massingham and Goldman
 \citeyearpar{Massingham2005Detecting} rooted at its midpoint, the
-17-taxon vertebrate B-globin tree from Yang et
-al. \citeyearpar{Yang2000CodonSubstitution}, and the 44-taxon
+17-taxon vertebrate $\beta$-globin tree from Yang et
+al. \citeyearpar{Yang2000CodonSubstitution} and the 44-taxon
 vertebrate tree used by the ENCODE project
@@
 simulate codon sequences with indels along each phylogenetic tree. The
-length of the root sequence was set to 500 codons, $\bm{\kappa}$ (the
-ratio of transition to transversion substitutions) was fixed at 4, and
-100 replicate alignments were simulated for each set of conditions
-unless otherwise noted. Indel lengths were drawn from a discretized
+length of the root sequence was set to 500 codons and $\bm{\kappa}$ (the
+ratio of transition to transversion substitutions) was fixed at 4.
+Indel lengths were drawn from a discretized
 power-law distribution with an exponential decay parameter of 1.8 and
@@
 insertion to deletion events was set to 1, and the rate of indel
-formation was scaled between 0 to 0.2 indel events per substitution
+formation was varied between 0 and 0.2 indel events per substitution
 per site.
@@
 of \sw selective pressures estimated from mammalian gene trees
-(unpublished data; log-normal parameters shown in Table
+(\citeyearpar{LindbladToh2011High}; log-normal parameters shown in Table
 \ref{table_1}). This distribution, with mean $\omg$ of 0.28 and 6\% of
@@
 discrete model of \sw $\omg$ variation was used to approximate the
-lognormal distribution by splitting the probability density into 50
+log-normal distribution by splitting the probability density into 50
 equally-spaced bins between $\omg$ values of 0 and 3, with the highest
@@
 
-Branch lengths for each of the simulation trees were multiplied by a
-correction factor before simulation to correct for the difference
-between our definition of branch lengths as the number of synonymous
-substitutions per synonymous site ($dS$) and Indelible's
-interpretation of branch length as the average number of substitutions
-per codon ($t$) \citep{Fletcher2010Effect}. They are related
-approximately by $t=3(NdN+SdS) = 3dS(\bar{\omg{}}N+S)$, where $N$ and
-$S$ are the proportion of nonsynonymous and synonymous sites and
-$\bar{\omg}$ is the mean $\omg$ across all sites. $S$ is approximately 0.3
-when $\bm{\kappa}=4$ \citep{Yang1998Synonymous} and the mean $\omg$ ratio for our chosen
-distribution is 0.277, yielding a $dS$-to-$t$ conversion factor of
-1.48 for all simulations performed.
+Branch lengths for each of the simulation trees were scaled before
+simulation to correct for the difference between our definition of
+branch lengths as the number of synonymous substitutions per
+synonymous site ($dS$) and Indelible's interpretation of branch length
+as the average number of substitutions per codon ($t$)
+\citep{Fletcher2010Effect}. They are related approximately by
+$t=3(NdN+SdS) = 3dS(\bar{\omg{}}N+S)$, where $N$ and $S$ are the
+proportion of nonsynonymous and synonymous sites and $\bar{\omg}$ is
+the mean $\omg$ across all sites. $S$ is approximately 0.3 when
+$\bm{\kappa}=4$ \citep{Yang1998Synonymous} and the mean $\omg$ ratio
+for our chosen distribution is 0.277, yielding a $dS$-to-$t$
+conversion factor of 1.48 for all simulations performed.
 
@@
 from any columns not within an identified conserved block were masked
-with $N$s. GUIDANCE and T-Coffee, which produce individual scores for
-each residue, were handled differently. A cutoff threshold was chosen
-for each method (0.5 for GUIDANCE and 5 for T-Coffee) and residues
-equal to or below that threshold were masked. If the cutoff threshold
-caused greater than 50\% of an alignment's residues to be masked, the
-threshold was relaxed to the highest value for which at least 50\% of
-residues remained unmasked. This cap on the severity of each filter
-was necessary because the scores from both methods were strongly
-affected by the simulation conditions, with much lower average scores
-at higher indel rates and divergences. GUIDANCE was run with 100
-bootstrap replicates and MAFFT as the bootstrap aligner (except when
-\pranka{} or \prankc{} was the original aligner, in which case 30
-bootstrap replicates were generated with \pranka{}) and T-Coffee was
-run using default settings and the {\em evaluate\_mode
-  -output=score\_ascii} command-line parameters to output alignment
-scores.
-
-Two unrealistic datasets were produced to serve as controls. First,
+with $N$s.
+
+\textcolor{red}{In contrast to Gblocks, GUIDANCE and T-Coffee produce
+  individual scores for each residue,} \textcolor{red}{ allowing for
+  individual residues to be masked instead of entire columns. GUIDANCE
+  generates many replicate alignments, each using a slightly perturbed
+  guide tree, and assigns to each residue a score from 0 to 1 based on
+  how consistently it was placed in the replicate alignments. We ran
+  GUIDANCE with 100 replicates using MAFFT as the aligner, except when
+  \prankc was the original aligner, in which case 30 replicate
+  alignments were produced using \pranka. T-Coffee calculates the
+  consistency between an input multiple alignment and independently
+  calculated pairwise alignments \citep{Notredame2003Using}, rounding
+  and normalizing residue scores into integers between 0 and
+  9. T-Coffee was run using its default settings and the {\em
+    evaluate\_mode --output=score\_ascii} command-line parameters to
+  output alignment scores.}
+
+\textcolor{red}{To filter alignments based on these residue-wise
+  scores, a cutoff threshold was chosen for each method (0.5 for
+  GUIDANCE and 5 for T-Coffee) and residues equal to or below that
+  threshold were masked. On a per-alignment basis, if the default
+  threshold caused greater than 50\% of residues to be masked, then
+  the threshold was relaxed to the highest value for which at least
+  50\% of residues remained unmasked. We found this adjustment
+  necessary because the scores from GUIDANCE and T-Coffee were
+  strongly affected by the simulation conditions, with much lower
+  average scores at higher indel rates and divergences. Requiring at
+  least 50\% of residues to remain unmasked ensured that enough data
+  were available for meaningful evolutionary analysis, mimicking
+  typical treatment of real data sets.}
+
+Two unrealistic but informative datasets were produced to serve as controls. First,
 the true simulated alignment was included in order to evaluate the \sw
@@
 length is the main source of information from which \sw inference
-methods derive their power. We refer to this method as the optimal
-method. Scores were handled in a manner similar to GUIDANCE and
+methods derive their power. We refer to this method as the 'optimal'
+filtering method. Scores were handled in a manner similar to GUIDANCE and
 T-Coffee, using a score threshold of 0.5.
@@
 maximum-likelihood methods implemented in the Phylogenetic Analysis by
-Maximum Likelihood (PAML, \citealt{Yang2007PAML}) and Sitewise Likelihood Ratio
-(SLR, \citealt{Massingham2005Detecting}) software packages.
+Maximum Likelihood (PAML; \citealt{Yang2007PAML}) and Sitewise Likelihood Ratio
+(SLR; \citealt{Massingham2005Detecting}) software packages.
 
@@
 substitutions at one site are independent of substitutions at all
-other sites. Importantly, no assumptions are made regarding the
+other sites. No assumptions are made regarding the
 distribution of $\omg$ ratios within the alignment. The value of
@@
 statistic measures the strength of evidence for non-neutral evolution
-at each site; following Massingham
+at each site; following Massingham and Goldman
 \citeyearpar{Massingham2005Detecting} we use a signed version of the
@@
 sites which were deleted in the reference sequence or inserted in a
-lineage not ancestral to the reference were dropped from the analysis.
+lineage not ancestral to the reference were \textcolor{red}{not
+  included in the final performance analysis.}
 
 To evaluate the power and error rates that might be achieved in
-real-world data analysis, the recommended cutoff thresholds for PAML’s
+real-world data analysis, the recommended cutoff thresholds for PAML's
 Bayesian posterior probabilities and the SLR statistic were used to
@@
 \caption{Alignments were simulated without indels for three tree
-  shapes and analyzed with SLR, PAML M8, or PAML M2. Fifty replicate
+  shapes and analyzed with SLR, \meight, or \mtwo. Fifty replicate
   alignments were simulated for each data point. The performance of
@@
 \subsection*{The Performance of Three Methods for Detecting \Sw Positive Selection}
-We first evaluated the ability of three \sw methods, PAML M2,
-PAML M8 and SLR, to accurately estimate \sw $\omg$ values and to
+We first evaluated the ability of three \sw methods, \mtwo,
+\meight and SLR, to accurately estimate \sw $\omg$ values and to
 detect positive selection under a range of tree lengths in the absence
@@
 very long branches present in such a sparse tree at high divergence
-levels. With lower average branch lengths at equivalent divergence
-levels, the two larger trees showed no signs of decreased performance
-even at a MPL of 2 substitutions per site, which is greater than any
-of the divergence levels found in groups of commonly analyzed
-vertebrate, insect and fungal species (Table \ref{table_2}).
-
-Comparing between the three \sw methods, we found that at the
+levels. With lower average branch lengths at equivalent MPLs, the two
+larger trees showed no signs of decreased performance even at a MPL of
+2 substitutions per site, which is greater than any of the divergence
+levels found in groups of commonly analyzed vertebrate, insect and
+fungal species (Table \ref{table_2}).
+
+Comparing the three methods for detecting positively selected sites, we found that at the
 recommended cutoff threshold (Figure \ref{fig_2}, top row) SLR showed
 the highest power to detect positive selection in all trees, followed
-by PAML M8 and PAML M2. In the smallest tree, the power of the two
-PAML methods was virtually zero while SLR reached a maximum TPR of ~18\%,
-at MPL=0.5. At the same divergence, SLR yielded TPRs of 25\% and 45\%
-in the 17-taxon and 44-taxon trees, respectively, with PAML M2 ranging
-between 50--75\% of SLR’s power and PAML M8 falling in between the two
+by \meight and \mtwo. In the smallest tree, the power of the two
+PAML methods was virtually zero while SLR reached a maximum TPR of ~18\%
+(at MPL=0.5). At the same divergence, SLR yielded TPRs of 25\% and 45\%
+in the 17-taxon and 44-taxon trees, respectively, with \mtwo ranging
+between 50--75\% of SLR's power and \meight falling between the two
 other methods.
@@
 The conservative nature of the default thresholds for PAML and SLR
-have been previously noted
+has been previously noted
 \citep{Anisimova2002Accuracy,Yang2005Bayes,Massingham2005Detecting},
@@
 are less likely to yield false positives than sites under neutral
-evolution, which the null model against which tests for positive
+evolution ($\omg$=1), the null model against which tests for positive
 selection are traditionally controlled.
@@
 For the purposes of our indel experiments, the observed similarity in
-error-controlled power levels indicated that the behavior of PAML M2,
-PAML M8, and SLR was similar enough not to warrant separately
+error-controlled power levels indicated that the behavior of \mtwo,
+\meight, and SLR was similar enough not to warrant separately
 evaluating all three methods in the subsequent indel simulation
@@
 performance measurements as Figure \ref{fig_2} for simulations without
-indels (gray lines, Figure \ref{fig_2}) and with indels (black lines,
-Figure \ref{fig_2}) analyzed using three different aligners and the
-true alignment. For the indel simulations, the indel rate was held
+indels (gray lines, Figure \ref{fig_3}) and with indels (black lines,
+Figure \ref{fig_3}) analyzed using three different aligners and the
+true alignment. For the indel simulations, the indel rate here was held
 constant at 0.1 indel event per substitution.
@@
 columns containing gaps harbor less evolutionary information than
-columns flush with sequence data. The lack of increased FPR showed
+columns with complete sequence data. The lack of increased FPR showed
 that SLR retained its conservative statistical performance even when
@@
 finding and provide evidence that this ranking may be applicable to a
-wider array of evolutionary analyses than the branch-site tests
-evaluated in their analysis.
+wider array of evolutionary analyses.
 
@@
 unavoidable result of alignment error at medium to high divergence
-levels (MPL $>$ 0.5), with even the most powerful aligner producing a
-marked reduction in TPR compared to the true alignment. The \tpr and
-$\omg$ results were qualitatively similar to the TPR results, with the
-\tpr values slightly more spread out between the different aligners
-than the TPRs, and the $\omg$ correlations slightly less so.
+levels (MPL $>$ 0.5), with even the most successful aligner producing
+a marked reduction in TPR compared to the true
+alignment. \textcolor{red}{The \tpr and $\omg$ results in the larger
+  two trees were qualitatively similar to the TPR results, showing
+  that the aligners tested led to different levels of sitewise
+  performance even when controlling for actual error rates or
+  assessing the sitewise $\omg$ correlation.}
 
@@
 divergences. ClustalW's FPR ranged between 0.001 to 0.005, while
-\prankc{}'s FPR was virtually identical to the true alignment in the
+\prankc{}'s FPR was virtually identical to that of the true alignment in the
 17-taxon and 44-taxon trees.
@@
 tendencies to produce false positive results. At higher vertebrate and
-\Dr divergence levels (MPL$=$0.8--1.0) false negatives became
-much more prevalent. The TPR for all inferred alignments was virtually
-zero in the 6-taxon tree, underscoring the necessity of including many
+\Dr divergence levels (MPL$=$0.8--1.0) false negatives became much
+more prevalent. The TPR for all inferred alignments was virtually zero
+in the 6-taxon tree, underscoring the necessity of including many
 species in the analysis of highly diverged sequences. In the two
-larger trees, \prankc resulted in very few additional false
-positives, but it suffered a 5--10\% reduction in TPR relative to the
-true alignment. Meanwhile, ClustalW showed a 50\% TPR reduction and
+larger trees, \prankc resulted in very few additional false positives,
+but it suffered a 5--10\% reduction in TPR relative to the true
+alignment. Meanwhile, ClustalW showed a 50\% TPR reduction and
 maintained a strongly elevated FPR. At higher divergences and in
@@
 alignment error, causing a marked reduction in \sw power even with the
-best-performing aligner.
+best-performing aligner. \textcolor{red}{Overall, the ranking of
+  aligners was clear, with \prankc performing better than MAFFT, and
+  MAFFT performing better than ClustalW, across all performance
+  measures and MPL divergence levels.}
 
@@
 To explore the effects of alignment error across a wider range of
-simulation conditions, we extended the simulations from Figure
+simulation conditions, we extended the simulations of Figure
 \ref{fig_3} across multiple indel rates. Figure \ref{fig_4} shows
@@
 each panel in Figure \ref{fig_4}; rows above and below the middle row
-represent higher and lower indel rates, respectively.
-
-The TPR values (Figure \ref{fig_4}A) showed a consistent pattern
+represent higher and lower indel rates, respectively. \textcolor{red}{Similarly, the
+bottom row of each panel in Figure \ref{fig_4} was simulated with an
+indel rate of 0 and corresponds to the 'No Indels' data in Figure
+\ref{fig_3}.}
+
+The TPR values (Figure \ref{fig_4}A) show a consistent pattern
 across the range of indel rates, with power decreasing as either the
-indel rate or the divergence level increased (except at the lowest
+indel rate or the divergence level increases (except at the lowest
 divergence levels, where the lack of evolutionary information yielded
@@
 
-The error-controlled \tpr results (Figure \ref{fig_4}C) provided a
-comprehensive map of the effect of alignment error on the detection of
+The error-controlled \tpr results (Figure \ref{fig_4}C) provide a
+comprehensive picture of the effect of alignment error on the detection of
 \sw positive selection. The two aligners not shown in the two other
 panels, MAFFT and \pranka, exhibited \tpr values intermediate to those
-from ClustalW and \prankc across the range of parameters tested. As
+from ClustalW and \prankc across the range of parameters tested, with \pranka performing
+better than MAFFT. As
 expected, performance was very similar between aligners at very low
@@
 patterns of low \tpr in the 6-taxon tree, but in the larger two trees
-only ClustalW and MAFFT alignments were unable to achieve high \tpr
+ClustalW and MAFFT alignments were unable to achieve high \tpr
 values, presumably due largely to their elevated FPR in those trees.
@@
 number of taxa, \prankc alignments would yield very few erroneous
-false positives in scans for positive selection in sequences with very
+false positives in scans for positive selection in sequences with even very
 high divergence levels. Furthermore, these results showed that false
@@
   either left unfiltered or filtered with one of four alignment
-  filters, and analyzed with SLR; true alignments were left unfiltered
-  and separately analyzed with SLR. One hundred fifty replicates were
-  simulated for each set of conditions. Squares are colored according
-  to the ratio of the performance of the indicated filter to the
-  performance of the unfiltered ClustalW or \prankc alignment as
-  measured by one of four summary statistics. In columns from left to
-  right: the number of unfiltered (i.e., non-N) residues remaining
-  in the alignment; the false positive rate (FPR) at the recommended
-  cutoff threshold; the true positive rate (TPR) at the recommended
-  cutoff threshold; the TPR at a 1\% FPR threshold (\tpr). Note that the
-  maximum percentage of filtered residues was capped at 50\% for all
-  methods except Gblocks.}
+  filters (Optimal, T-Coffee, GUIDANCE, Gblocks), and analyzed with
+  SLR; true alignments were left unfiltered and separately analyzed
+  with SLR. One hundred and fifty replicates were simulated for each
+  set of conditions. Squares are colored according to the ratio of the
+  performance of the indicated filter to the performance of the
+  unfiltered ClustalW or \prankc alignment as measured by one of four
+  summary statistics. In columns from left to right: the number of
+  unfiltered (i.e., non-N) residues remaining in the alignment; the
+  false positive rate (FPR) at the recommended cutoff threshold; the
+  true positive rate (TPR) at the recommended cutoff threshold; the
+  TPR at a 1\% FPR threshold (\tpr). Note that the maximum percentage
+  of residues removed by filtering was capped at 50\% for all methods
+  except Gblocks.}
 \label{fig_5}
@@
 performance through the introduction of false negatives and false
-positives, we tested whether a number of alignment filtering methods
+positives, we tested whether alignment filtering methods
 could reduce error rates and improve the power of \sw detection of
@@
 whether a filter had successfully improved the \sw power of a dataset
-since this value is sensitive to changes in both the FPR and TPR.
+since this value is sensitive to changes in both the FPR and TPR
+\textcolor{red}{(note that the \tpr{} controls the FPR post-filtering,
+  accounting for the tendency of filtering to reduce the FPR at a
+  given cutoff threshold).}
 
@@
 rows, Figure \ref{fig_5}). The true alignment nearly always showed
-reduced FPR and increased TPR and \tpr compared to the inferred
+smaller FPR and greater TPR and \tpr compared to the inferred
 alignments, with a greater magnitude of change relative to the
@@
 change that might be achieved by a perfect alignment filter. One
-exception to the general trend was the observation of two data points
-with slightly elevated FPR compared to \prankc alignments at the
-lowest indel rate. This small incosistency may be explained by
+exception to the general trend was the observation of two simulation conditions
+with slightly elevated FPRs in the true alignment compared to \prankc alignments at the
+lowest indel rate. This small inconsistency may be explained by
 stochastic variation in false positive counts, as the absolute value
@@
 70\% of residues removed, resulting in the loss of a large number of
-correctly-aligned true positive sites. Dessimoz et
-al. \citeyearpar{Dessimoz2010Phylogenetic} found Gblocks filtering to
-have a negative effetc on the accuracy of phylogenetic inference; our
+correctly-aligned true positive sites. Dessimoz and Gil
+\citeyearpar{Dessimoz2010Phylogenetic} found Gblocks filtering to have
+a negative effetc on the accuracy of phylogenetic inference; our
 results provide additional evidence in support of their finding,
@@
 improvements. Importantly, GUIDANCE only showed improved performance
-at high divergence levels (e.g., MPL$>$1.6) well above those found in
+at high divergence levels (e.g., MPL$>$1.6), well above those found in
 commonly-analyzed groups of species. Thus, the use of unfiltered
@@
 divergence levels, using simulation parameters designed to approximate
-mammalian protein-coding genes. We evaluated the ability of four
-alignment methods and three alignment filtering methods to accurately
-reconstruct the codon-level homology for sitewise inference, using the
-FPR, TPR, and error-controlled \tpr of the sitewise detection of
-positive selection as our main measures of performance.
+the analysis of mammalian and other typical protein-coding genes. We
+evaluated the ability of four alignment methods and three alignment
+filtering methods to produce alignments for detecting positively
+selected sites, using the FPR, TPR, and error-controlled \tpr of the
+sitewise detection of positive selection as our main measures of
+performance.
 
@@
 particular, when the number of sequences was large, its sitewise FPRs
-were virtually indistinguishable from the true alignment.
+were virtually indistinguishable from the true alignment's.
 
@@
 false positives when misalignment causes apparent positive selection
-along only the foreground branch, while the SLR test would produce
+along only the foreground branch, while the SLR sitewise test would produce
 false positives only when misalignment causes a signal of positive
@@
 
-Our simulations did not include biologically realistic models of
+Our simulations did not include fully biologically realistic models of
 spatial or temporal variability in the rate of indel formation or in
@@
 constraint---to further increase the applicability and accuracy of
-such experiments. As genomes rapidly accumulate in the databases and
+such experiments, \textcolor{red}{and we believe that flexible and
+  accessible simulation programs such as INDELible
+  \citep{Fletcher2009INDELible} and PhyloSim \citep{Sipos2011PhyloSim}
+  will play an important role in the quantitative assessment of
+  alignment algorithms and alignment-dependent comaprative analyses.}
+
+As genomes rapidly accumulate in the databases and
 large-scale analyses become the norm, we hope that the development and
@@
 
+GJ was funded by a Gates Cambridge Trust Scholarship and is a member
+of Darwin College, University of Cambridge.
+
 \bibliographystyle{mbe3} \fontsize{9}{10}\selectfont% \itemsep 0pt

### 2011-07-21 projects/slrsim/paper/slrsim.tex
Paper final draft.

@@
 \usepackage[nolists,tablesfirst,nomarkers]{endfloat}
-\usepackage[a4paper]{geometry}
+\usepackage[a4paper,hmargin=2.5cm,vmargin=2.5cm]{geometry}
 \usepackage{color}
@@
 performed the best and Gblocks performed the worst, failing to improve
-even the error-prone ClustalW alignments. The performance of the true
+even the error-prone ClustalW alignments. Controls using the true
 alignment and an optimal filtering method suggested that performance
@@
 reduce the prevalence of false negatives, especially at higher
-divergence levels and indel rates. \textcolor{red}{As no combination
-  of aligner plus filter significantly outperformed the best aligner
-  alone,} we conclude that for the sitewise evolutionary analysis of
-most vertebrate, insect and fungal phylogenies, the use of unfiltered
-PRANK alignments is recommended.
+divergence levels and indel rates. As no combination of aligner plus
+filter significantly outperformed the best aligner alone, we conclude
+that for the sitewise evolutionary analysis of most vertebrate, insect
+and fungal phylogenies, the use of unfiltered PRANK alignments is
+recommended.
 
@@
 proteins than in highly folded enzymes and housekeeping genes
-\citep{delaChaux2007DNA} . This suggests that well-folded protein
+\citep{delaChaux2007DNA}. This suggests that well-folded protein
 regions will experience fewer biological indels---and will therefore
@@
 questions, regarding the nature of alignment error and its
-quantitative impact on the detection of positive selection unanswered
+quantitative impact on the detection of positive selection, unanswered
 by empirical studies.
@@
 \citep{Loytynoja2008PhylogenyAware} produce estimates of alignment
-confidence or reliability. \textcolor{red}{GUIDANCE, which measures
-  the robustness of alignment regions to perturbations in the guide
-  tree used for progressive alignment, has also been proposed as an
-  alignment confidence score
-  \citep{Penn2010Alignment}}. Unfortunately, despite their widespread
-use, the impact of the many available \textcolor{red}{alignment
-  scoring and filtering} methods on phylogenetic and evolutionary
+confidence or reliability. GUIDANCE, which measures the robustness of
+alignment regions to perturbations in the guide tree used for
+progressive alignment, has also been proposed as an alignment
+confidence score \citep{Penn2010Alignment}. Unfortunately, despite
+their widespread use, the impact of the many available alignment
+scoring and filtering methods on phylogenetic and evolutionary
 analyses has not been well studied. Even for a single filtering
@@
 
-The Indelible program \citep{Fletcher2009INDELible} was used to
+The INDELible program \citep{Fletcher2009INDELible} was used to
 simulate codon sequences with indels along each phylogenetic tree. The
@@
 a maximum value of 40, yielding a mean indel length of 3.33 codons and
-standard deviation of 5.51 codons. This power-law model of indel
+standard deviation of 5.51 codons. The power-law model of indel
 lengths is well-supported by empirical studies
@@
 under neutral selection or positive selection
-\citep{Smith1970Natural,Kimura1974SomePrinciples}. Indelible's general
+\citep{Smith1970Natural,Kimura1974SomePrinciples}. INDELible's general
 discrete model of \sw $\omg$ variation was used to approximate the
@@
 branch lengths as the number of synonymous substitutions per
-synonymous site ($dS$) and Indelible's interpretation of branch length
+synonymous site ($dS$) and INDELible's interpretation of branch length
 as the average number of substitutions per codon ($t$)
@@
 
-\textcolor{red}{In contrast to Gblocks, GUIDANCE and T-Coffee produce
-  individual scores for each residue,} \textcolor{red}{ allowing for
-  individual residues to be masked instead of entire columns. GUIDANCE
-  generates many replicate alignments, each using a slightly perturbed
-  guide tree, and assigns to each residue a score from 0 to 1 based on
-  how consistently it was placed in the replicate alignments. We ran
-  GUIDANCE with 100 replicates using MAFFT as the aligner, except when
-  \prankc was the original aligner, in which case 30 replicate
-  alignments were produced using \pranka. T-Coffee calculates the
-  consistency between an input multiple alignment and independently
-  calculated pairwise alignments \citep{Notredame2003Using}, rounding
-  and normalizing residue scores into integers between 0 and
-  9. T-Coffee was run using its default settings and the {\em
-    evaluate\_mode --output=score\_ascii} command-line parameters to
-  output alignment scores.}
-
-\textcolor{red}{To filter alignments based on these residue-wise
-  scores, a cutoff threshold was chosen for each method (0.5 for
-  GUIDANCE and 5 for T-Coffee) and residues equal to or below that
-  threshold were masked. On a per-alignment basis, if the default
-  threshold caused greater than 50\% of residues to be masked, then
-  the threshold was relaxed to the highest value for which at least
-  50\% of residues remained unmasked. We found this adjustment
-  necessary because the scores from GUIDANCE and T-Coffee were
-  strongly affected by the simulation conditions, with much lower
-  average scores at higher indel rates and divergences. Requiring at
-  least 50\% of residues to remain unmasked ensured that enough data
-  were available for meaningful evolutionary analysis, mimicking
-  typical treatment of real data sets.}
+GUIDANCE and T-Coffee produce individual scores for each residue,
+allowing individual residues to be masked instead of entire
+columns. GUIDANCE generates many replicate alignments, each using a
+slightly perturbed guide tree, with either MAFFT or \pranka as the
+bootstrap aligner. The program then assigns to each residue from the
+input alignment a score from 0 to 1 based on how consistently it was
+placed in the replicate alignments. In order to maximize the
+similarity between the input aligner and the bootstrap aligner, we ran
+GUIDANCE with 100 MAFFT replicates when filtering ClustalW alignments
+and with 30 \pranka replicates when filtering \prankc
+alignments. T-Coffee calculates the residue-wise consistency between
+an input multiple alignment and independently calculated pairwise
+alignments \citep{Notredame2003Using}, rounding and normalizing
+residue scores into integers between 0 and 9. T-Coffee was run using
+its default settings and the {\em evaluate\_mode
+  --output=score\_ascii} command-line parameters to output alignment
+scores.
+
+To filter alignments based on these residue-wise scores, a cutoff
+threshold was chosen for each method (0.5 for GUIDANCE and 5 for
+T-Coffee) and residues equal to or below that threshold were
+masked. On a per-alignment basis, if the default threshold caused
+greater than 50\% of residues to be masked, then the threshold was
+relaxed to the highest value for which at least 50\% of residues
+remained. We found this adjustment necessary because the scores from
+GUIDANCE and T-Coffee were strongly affected by the simulation
+conditions, with much lower average scores at higher indel rates and
+divergences. Requiring at least 50\% of residues to remain unmasked
+ensured that enough data were available for meaningful evolutionary
+analysis, mimicking typical treatment of real data sets.
 
@@
 length is the main source of information from which \sw inference
-methods derive their power. We refer to this method as the 'optimal'
+methods derive their power. We refer to this method as the `optimal'
 filtering method. Scores were handled in a manner similar to GUIDANCE and
@@
 \ref{fig_1}) and all \sw statistics were mapped from alignment columns
-to sequence positions in the reference sequence. This approach bears
-some resemblance to the process of mapping alignment-based
-evolutionary estimates onto a single member of the alignment for
-further analysis and integration with other genome-referenced data (as
-is often done, for example, using mammalian alignments and a human
-reference). As a result of this reference sequence based mapping,
-sites which were deleted in the reference sequence or inserted in a
-lineage not ancestral to the reference were \textcolor{red}{not
-  included in the final performance analysis.}
+to sequence positions in the reference sequence. This approach
+corresponds to the process of mapping alignment-based evolutionary
+estimates onto a single member of the alignment for further analysis
+and integration with other genome-referenced data (as is often done,
+for example, using mammalian alignments and a human reference). As a
+result of this reference sequence based mapping, sites which were
+deleted in the reference sequence or inserted in a lineage not
+ancestral to the reference were not included in the final performance
+analysis.
 
@@
 
-The conservative nature of the default thresholds for PAML and SLR
-has been previously noted
+The conservative nature of the default thresholds for PAML and SLR has
+been previously noted
 \citep{Anisimova2002Accuracy,Yang2005Bayes,Massingham2005Detecting},
@@
 performance measurements as Figure \ref{fig_2} for simulations without
-indels (gray lines, Figure \ref{fig_3}) and with indels (black lines,
-Figure \ref{fig_3}) analyzed using three different aligners and the
-true alignment. For the indel simulations, the indel rate here was held
+indels (gray lines, Figure \ref{fig_3}) and with indels (black and
+textured lines, Figure \ref{fig_3}) analyzed using three different
+aligners (ClustalW, MAFFT, and \prankc) and the true alignment. (We
+omitted results for \pranka alignments from Figure \ref{fig_3} in
+order to reduce visual clutter; a full comparison of all aligners
+tested is shown in Figure \ref{fig_4}C and discussed in the next
+section.) For the indel simulations, the indel rate here was held
 constant at 0.1 indel event per substitution.
@@
 \end{center}
-\caption{Sequences were simulated without indels (gray lines) or with
-  indels (black lines) using one of three tree shapes, aligned with
+\caption{Sequences were simulated without indels (solid gray lines) or with
+  indels (solid black and textured lines) using one of three tree shapes, aligned with
   one of three aligners, and analyzed with SLR; true alignments were
@@
 levels (MPL $>$ 0.5), with even the most successful aligner producing
-a marked reduction in TPR compared to the true
-alignment. \textcolor{red}{The \tpr and $\omg$ results in the larger
-  two trees were qualitatively similar to the TPR results, showing
-  that the aligners tested led to different levels of sitewise
-  performance even when controlling for actual error rates or
-  assessing the sitewise $\omg$ correlation.}
+a marked reduction in TPR compared to the true alignment. The \tpr and
+$\omg$ results in the larger two trees were qualitatively similar to
+the TPR results, showing that the aligners tested led to different
+levels of sitewise performance even when controlling for actual error
+rates or assessing the sitewise $\omg$ correlation.
 
@@
 alignment error, causing a marked reduction in \sw power even with the
-best-performing aligner. \textcolor{red}{Overall, the ranking of
-  aligners was clear, with \prankc performing better than MAFFT, and
-  MAFFT performing better than ClustalW, across all performance
-  measures and MPL divergence levels.}
+best-performing aligner. Overall, the ranking of aligners was clear,
+with \prankc performing better than MAFFT and MAFFT performing better
+than ClustalW across all performance measures and MPL divergence
+levels.
 
@@
 \caption{Sequences were simulated with indels using one of three tree
-  shapes and a range of indel rates and mean path length (MPL)
-  divergence levels. Alignments were inferred with one of four
-  aligners and analyzed with SLR; true alignments were separately
-  analyzed with SLR. One hundred replicates were simulated for each
-  set of conditions. Each square is colored according to the
-  performance at a given (indel rate, MPL) pair as measured by one of
-  three summary statistics: (a) the true positive rate (TPR) at the
-  recommended cutoff threshold, (b) the false positive rate (FPR) at
-  the recommended cutoff threshold, or (c) the TPR at a 1\% FPR
-  threshold.}
+  shapes (t-, 17- or 44-taxon) and a range of indel rates and mean
+  path length (MPL) divergence levels. Alignments were inferred with
+  one of four aligners (ClustalW, MAFFT, \pranka{}, \prankc) and
+  analyzed with SLR; true alignments were separately analyzed with
+  SLR. One hundred replicates were simulated for each set of
+  conditions. Each cell is colored according to the performance at a
+  given (indel rate, MPL) pair as measured by one of three summary
+  statistics: (a) the true positive rate (TPR) at the recommended
+  cutoff threshold, (b) the false positive rate (FPR) at the
+  recommended cutoff threshold, or (c) the TPR at a 1\% FPR
+  threshold. Results for MAFFT and \pranka are omitted from (a) and
+  (b); as in (c) they show characteristics intermediate between
+  ClustalW and \prankc.}
 \label{fig_4}
@@
 error-controlled \tpr for all aligners tested (Figure
-\ref{fig_4}C). The results from Figure \ref{fig_3}, which were
-simulated with an indel rate of 0.1, correspond to the middle row of
-each panel in Figure \ref{fig_4}; rows above and below the middle row
-represent higher and lower indel rates, respectively. \textcolor{red}{Similarly, the
-bottom row of each panel in Figure \ref{fig_4} was simulated with an
-indel rate of 0 and corresponds to the 'No Indels' data in Figure
-\ref{fig_3}.}
+\ref{fig_4}C). (MAFFT and \pranka were omitted from Figure
+\ref{fig_4}A--B to save space, but their performance fell between that
+of ClustalW and \prankc for all performance measures, with \pranka
+slightly outperforming MAFFT; a comprehensive set of TPR, FPR, and
+\tpr results can be found in Supplementary Figure S1.)
+The results from Figure \ref{fig_3}, which were simulated with an
+indel rate of 0.1, correspond to the middle row of each panel in
+Figure \ref{fig_4}; rows above and below the middle row represent
+higher and lower indel rates, respectively. Similarly, the bottom row
+of each panel in Figure \ref{fig_4} was simulated with an indel rate
+of zero and corresponds to the `No Indels' data in Figure \ref{fig_3}.
 
@@
 tree. The FPR results in the 44-taxon tree confirmed the strange
-effect of ClustalW’s alignments on the \sw FPR: at the highest indel
+effect of ClustalW's alignments on the \sw FPR: at the highest indel
 rates, ClustalW showed a negative relationship between FPR and
@@
 \sw positive selection. The two aligners not shown in the two other
-panels, MAFFT and \pranka, exhibited \tpr values intermediate to those
+panels (MAFFT and \pranka) exhibited \tpr values intermediate to those
 from ClustalW and \prankc across the range of parameters tested, with \pranka performing
@@
   with SLR. One hundred and fifty replicates were simulated for each
-  set of conditions. Squares are colored according to the ratio of the
+  set of conditions. Cells are colored according to the ratio of the
   performance of the indicated filter to the performance of the
@@
   summary statistics. In columns from left to right: the number of
-  unfiltered (i.e., non-N) residues remaining in the alignment; the
+  unfiltered (i.e., non-$N$) residues remaining in the alignment; the
   false positive rate (FPR) at the recommended cutoff threshold; the
@@
 whether a filter had successfully improved the \sw power of a dataset
-since this value is sensitive to changes in both the FPR and TPR
-\textcolor{red}{(note that the \tpr{} controls the FPR post-filtering,
-  accounting for the tendency of filtering to reduce the FPR at a
-  given cutoff threshold).}
+since this value is sensitive to changes in both the FPR and TPR. Note
+that the \tpr controls the FPR post-filtering, accounting for the
+tendency of filtering to reduce the FPR at a given cutoff threshold.
 
@@
 rows, Figure \ref{fig_5}). The true alignment nearly always showed
-smaller FPR and greater TPR and \tpr compared to the inferred
-alignments, with a greater magnitude of change relative to the
-ClustalW alignments than to the \prankc alignments. These scores
-represented the direction and an upper limit on the magnitude of
-change that might be achieved by a perfect alignment filter. One
-exception to the general trend was the observation of two simulation conditions
-with slightly elevated FPRs in the true alignment compared to \prankc alignments at the
-lowest indel rate. This small inconsistency may be explained by
-stochastic variation in false positive counts, as the absolute value
-of the FPR was very low in both datasets under those conditions (on
-the order of $5\times10^{-4}$, Figure \ref{fig_4}B). A similar slight FPR
-elevation was also observed at the same indel rate for the optimal and
-GUIDANCE filters.
+smaller FPR (red cells) and greater TPR and \tpr (blue cells) compared
+to the inferred alignments, with a greater magnitude of change
+relative to the ClustalW alignments than to the \prankc alignments
+(darker shades cf. lighter shades). These scores represented the
+direction and an upper limit on the magnitude of change that might be
+achieved by a perfect alignment filter. One exception to the general
+trend was the observation of two simulation conditions with slightly
+elevated FPRs in the true alignment compared to \prankc alignments (at
+an indel rate of 0.04 and MPL of 0.8 and 2.0). This small
+inconsistency may be explained by stochastic variation in false
+positive counts, as the absolute value of the FPR was very low in both
+datasets under those conditions (on the order of $5\times10^{-4}$,
+Figure \ref{fig_4}B). A similar slight FPR elevation was also observed
+at the same indel rate for the optimal, GUIDANCE and T-Coffee filters.
 
@@
 
-Overall, our alignment filtering simulations found that filtering
-methods based on GUIDANCE and T-Coffee scores had the ability to mask
-away many residues causing false positives and false negatives from
-ClustalW alignments, with high levels of improvement across a wide
-range of divergence levels and indel rates. Filtering was less
-beneficial when applied to \prankc alignments, with T-Coffee reducing
-performance and GUIDANCE yielding only mild \tpr
-improvements. Importantly, GUIDANCE only showed improved performance
-at high divergence levels (e.g., MPL$>$1.6), well above those found in
-commonly-analyzed groups of species. Thus, the use of unfiltered
-\prankc alignments would yield the best performance for detecting \sw
-positive selection when analyzing protein-coding sequences across a
-wide range of indel rates and commonly encountered sequence divergence
-levels.
+Overall, our alignment filtering simulations found that Gblocks rarely
+improves alignments for sitewise detection of positive selection, but
+filtering methods based on GUIDANCE and T-Coffee scores have a good
+ability to mask out misaligned residues that cause false positives and
+false negatives in sitewise inference. This beneficial effect is
+highly dependent on simulation conditions and the input aligner. For
+ClustalW alignments (which, left unfiltered, led to many false
+positives and false negatives) both GUIDANCE and T-Coffee showed good
+ability to improve sitewise performance, behaving qualitatively
+similarly to the `optimal' filter. In order to compare the TPR, FPR
+and \tpr values obtained with filtered ClustalW alignments to those
+obtained with unfiltered alignments, we ran ClustalW with GUIDANCE
+filtering using all three trees and the same range of indel rates and
+MPLs as used for Figure \ref{fig_4}; these results are included in
+Supplementary Figure S1. GUIDANCE filtering generally
+improved the performance of ClustalW alignments, resulting in sitewise
+performance comparable to (but not better than) unfiltered MAFFT
+alignments.
+
+Filtering was less beneficial when applied to the more accurate
+\prankc alignments, with T-Coffee reducing performance and GUIDANCE
+yielding only mild \tpr improvements. Importantly, GUIDANCE only
+showed improved performance at high divergence levels (e.g.,
+MPL$>$1.6), well above those found in commonly-analyzed groups of
+species. Thus, the use of unfiltered \prankc alignments would yield
+the best performance for detecting \sw positive selection when
+analyzing protein-coding sequences across a wide range of indel rates
+and commonly encountered sequence divergence levels. The performance
+of the `optimal' filter on \prankc alignments suggests that mild
+further improvements to filtering strategies may be possible, but the
+potential for improvement is small and may be of little value.
 
@@
 produced more sitewise false positives than MAFFT or \prankc even at
-low divergence levels, suggesting that its use should be avoided even
-when analyzing closely-related sequences. \prankc, on the other hand,
-resulted in very low FPRs even at very high divergences. In
-particular, when the number of sequences was large, its sitewise FPRs
-were virtually indistinguishable from the true alignment's.
+low MPLs, suggesting that its use should be avoided even when
+analyzing closely-related sequences. \prankc, on the other hand,
+resulted in very low FPRs even at higher MPLs. In particular, when the
+number of sequences in the tree was large, \prankc{}'s sitewise FPRs
+were virtually indistinguishable from those of the true alignment.
+
+An important observation regarding the size of tree analyzed was that
+the 6-taxon tree caused qualitatively similar problems (e.g., elevated
+FPRs and reduced TPRs) for all aligners, suggesting that poor
+performance is inevitable when analyzing a small number of moderately
+divergent sequences. The small amount of evolutionary information
+combined with the longer branch lengths makes alignment difficult and
+increases the tendency of misalignment to cause sitewise false
+positives. Thus, we reiterate the well-established recommendation to
+use large numbers of sequences when inferring sitewise positive
+selection \citep{Anisimova2001Accuracy,Anisimova2002Accuracy}. We
+additionally point out that when analyzing sequences with indels, the
+shape of the tree may matter as well: trees with long internal
+branches may be especially prone to false positives, as longer
+branches are more difficult to align.
 
@@
 to recover as true positives through alignment filtering, and we found
-that both the optimal filter and GUIDANCE were able to successfully
+that both the `optimal' filter and GUIDANCE were able to successfully
 recover these false negatives at high divergence levels, resulting in
@@
 
-Among the aligners and filters tested here, we can recommend the use
-of unfiltered \prankc alignments for the detection of sitewise
+The manual or automated adjustment of alignments has been thought by
+many to be an important step in evolutionary analyses due to fear of a
+high prevalence of misalignment-induced false positives. While this is
+true for some aligners, we find that more accurate alignment
+algorithms result in significantly fewer false positives in the
+subsequent detection of sitewise positive selection. This strongly
+reduces the beneficial effect of alignment filtering, so much so that
+current filtering methods are essentially unable to improve the
+performance of \prankc alignments when analyzed with SLR.
+
+As a result, among the aligners and filters tested here, we recommend
+the use of unfiltered \prankc alignments for the detection of sitewise
 positive selection in all but the most divergent phylogenies. GUIDANCE
@@
 opposed to the elimination of false positives. For an analysis where
-the control of false positives is paramount, the added computational
-expense of running many bootstrap alignment replicates (as is done
-GUIDANCE) may not be worth the possibility of a slight increase in
-power. The development of efficient and effective alignment filters,
-particularly with an eye towards recovering false negatives as
-identified in this study, will be an interesting area for future
-research.
+the control of false positives is the primary concern, the added
+computational expense of running many bootstrap alignment replicates
+(as performed by GUIDANCE) may not be worth the possibility of a
+slight increase in power. The development of efficient and effective
+alignment filters, particularly with an eye towards recovering false
+negatives as identified in this study, will be an interesting area for
+future research.
 
@@
 
+A tangential but interesting observation from our simulations was that
+all three methods we tested for the sitewise detection of positive
+selection were highly conservative when analyzing alignments with a
+mammalian-like $\omg$ distribution. SLR, for example, yielded FPRs
+well below the nominal 5\% level. This was due to a mismatch between
+SLR's null model of neutral evolution ($\omg=1$) and our more
+realistic distribution of non-positive sites (where
+$\bar{\omg}=0.277$). We note that while the ability of SLR and PAML's
+sitewise models to distinguish between neutral evolution and positive
+selection in a well-controlled manner is important, the majority of
+protein sites evolve under moderate purifying selection. Future work
+on methods to adaptively adjust the cutoff thresholds to achieve
+better statistical control under non-neutral (and possibly unknown)
+$\omg$ distributions could yield much greater power to detect positive
+selection while maintaining good control of error rates.
+
 We showed here that even relatively simple evolutionary simulation
@@
 constraint---to further increase the applicability and accuracy of
-such experiments, \textcolor{red}{and we believe that flexible and
-  accessible simulation programs such as INDELible
-  \citep{Fletcher2009INDELible} and PhyloSim \citep{Sipos2011PhyloSim}
-  will play an important role in the quantitative assessment of
-  alignment algorithms and alignment-dependent comaprative analyses.}
+such experiments, and we believe that flexible and accessible
+simulation programs such as INDELible \citep{Fletcher2009INDELible}
+and PhyloSim \citep{Sipos2011PhyloSim} will play an important role in
+the quantitative assessment of alignment algorithms and
+alignment-dependent comparative analyses.
 
@@
 \bibliography{slrsim}
+
 \end{document}

### 2011-07-21 projects/slrsim/paper/supp_slrsim.tex
Paper final draft.

\documentclass{article}
\usepackage{natbib,graphicx,dblfloatfix,multirow,array,threeparttable,xspace}
\usepackage[landscape,hmargin=1.5cm,vmargin=1.5cm]{geometry}

\newcommand{\pranka}{PRANK$_{\textrm{AA}}$\xspace}

\begin{document}

\makeatletter 
\renewcommand{\figurename}{Supplementary Figure}
\renewcommand{\thefigure}{S\@arabic\c@figure}

\begin{figure}[t]
\begin{center}
%\includegraphics[scale=0.65]{supp_fig1.pdf}
\includegraphics[scale=0.8]{supp_fig1.pdf}
\end{center}
\caption{This figure depicts the same simulations and uses the same
  formatting as Figure 4, except that results for MAFFT and \pranka
  have been added to sections (a) and (b) and results using ClustalW +
  GUIDANCE have been added to all sections.}
\label{supp_fig1}
\end{figure}

\end{document}

### 2011-09-05 scripts/mysql_functions.R
Slrsim updates.

@@ -34,7 +34,8 @@ connect <- function(dbname=NULL) {
   return(con)
 }
 
-dbUpdateVars <- function(conn, dbtable, dataframe=NULL, primary, vars=names(dataframe)) { 
+dbUpdateVars <- function(conn, dbtable, dataframe=NULL, primary, vars=colnames(dataframe)) { 
+  print(paste("dbupdateVars for table", dbtable))
   if (!dbExistsTable(conn, dbtable)) { 
     stop("The target table \"", dbtable, "\" doesn't exist in the database \"", dbGetInfo(conn)$dbname, "\"\n\n", call. = FALSE)
   }
@@ -47,14 +48,23 @@ dbUpdateVars <- function(conn, dbtable, dataframe=NULL, primary, vars=names(data
   if (!all(toupper(vars) %in% toupper(names(dataframe)))) { 
     stop("One or more variables don't exist in the source dataframe\n\n", call. = FALSE) 
   } 
-  if (!(toupper(primary) %in% toupper(dbListFields(con, dbtable)))) { 
+  if (!(toupper(primary) %in% toupper(dbListFields(conn, dbtable)))) { 
     stop("The primary key variable doesn't exist in the target table\n\n", call. = FALSE) 
   } 
-  if (!all(toupper(vars) %in% toupper(dbListFields(con, dbtable)))) { 
+  if (!all(toupper(vars) %in% toupper(dbListFields(conn, dbtable)))) { 
+    print(toupper(vars))
+    print(toupper(dbListFields(conn, dbtable)))
     stop("One or more variables don't exist in the target table\n\n", call. = FALSE) 
   } 
 
   all.vars <- unique(c(primary,vars))
+
+  for (i in 1:length(all.vars)) {
+    cur.var <- all.vars[i]
+    if (is.character(dataframe[, cur.var])) {
+      dataframe[, cur.var] <- gsub("'", '', dataframe[, cur.var])
+    }
+  }
 
   if(length(all.vars) > 1) { 
     pastedvars <- paste("'", apply(dataframe[, all.vars], 1, paste, collapse="', '"), "'", sep="") 

### 2011-09-06 projects/slrsim/paper/slrsim.tex
Tiny updates.
Slrsim updates.

@@
 \usepackage{color}
+%\usepackage{ulem}
 
@@
 \begin{document}
+
+\renewcommand\refname{Literature Cited}
 
@@
 \newcommand{\Sw}{Sitewise\xspace}
-\newcommand{\Dr}{{\em Drosophila}\xspace}
+\newcommand{\Dr}{{\emph{Drosophila}}\xspace}
 
 \title{The effects of alignment error and alignment filtering on the \sw detection of positive selection}
-
-\author{{Gregory Jordan and Nick Goldman}\\
-\affiliation{European Bioinformatics Institute, Wellcome Trust Genome Campus, Hinxton, Cambridgeshire, CB10 1SD, United Kingdom}}
+%%\title{The effects of alignment error and alignment filtering on the \sw detection of positive selection in mammals}
+%\title{The effects of alignment error and alignment filtering on the \sw detection of positive selection in vertebrate genomics}
+\author{{Gregory Jordan and Nick Goldman}}
+%\affiliation{European Bioinformatics Institute, Wellcome Trust Genome Campus, Hinxton, Cambridgeshire, CB10 1SD, United Kingdom}}
+\maketitle
 %\authorlist{Jordan and Goldman}
@@
 \section*{Abstract}
-
-The effects of alignment error on many types of evolutionary inference
-are not well understood. Alignment filters are commonly applied to try
-to reduce errors resulting from misalignment, but neither the
-prevalence of such errors nor the ability of filters to reduce them
-has been systematically studied. Focusing on the sitewise detection of
-positive selection, we performed simulation experiments to quantify
-the false positives and false negatives introduced by alignment error
-and the ability of alignment filters to improve performance. We found
-that some aligners led to many false positive results across a wide
-range of divergence levels, while others resulted in very few. False
-negatives were a problem for all aligners, increasing with
-divergence. Of the aligners tested, PRANK consistently performed the
-best and ClustalW performed the worst. The best alignment filters
-showed good ability to reduce the error rates from ClustalW
-alignments, but they failed to improve the performance of PRANK
-alignments under most circumstances. Of the filters tested, GUIDANCE
-performed the best and Gblocks performed the worst, failing to improve
-even the error-prone ClustalW alignments. Controls using the true
-alignment and an optimal filtering method suggested that performance
-improvements could be gained by improving aligners or filters to
-reduce the prevalence of false negatives, especially at higher
-divergence levels and indel rates. As no combination of aligner plus
-filter significantly outperformed the best aligner alone, we conclude
-that for the sitewise evolutionary analysis of most vertebrate, insect
-and fungal phylogenies, the use of unfiltered PRANK alignments is
-recommended.
-
-%}
-
+\textcolor{red}{
+Alignment filters are commonly applied to try to reduce errors
+resulting from misalignment, but the prevalence of such errors and the
+ability of filters to reduce them is not well understood.
+% 
+Focusing on the sitewise detection of positive selection across a wide
+range of divergence levels and indel rates, we performed simulation
+experiments to quantify the false positives and false negatives
+introduced by alignment error and the ability of alignment filters to
+improve performance.
+%
+We found that some aligners led to many false positives while others
+resulted in very few.  False negatives were a problem for all
+aligners, increasing with divergence. Of the aligners tested, PRANK's
+codon-based alignments consistently performed the best and ClustalW
+performed the worst.
+%
+Of the filters tested, GUIDANCE performed the best and Gblocks
+performed the worst. Although some filters showed good ability to
+reduce the error rates from ClustalW and MAFFT alignments, none were
+found to substantially improve the performance of PRANK alignments
+under most \textcolor{red}{conditions}.
+% 
+Our results revealed distinct trends in error rates and power levels
+for aligners and filters within a biologically plausible parameter
+space. Controls using the true alignment and an optimal filtering
+method suggested that performance improvements could be gained by
+improving aligners or filters to reduce the prevalence of false
+negatives, especially at higher divergence levels and indel rates.
+}
 %\keywords{Alignment, positive selection, alignment filtering, indels}
@@
 power of \sw evolutionary inference of positive selection. Our
-approach is similar to that of Anisimova et al.
+approach builds from that of Anisimova et al.
 \citeyearpar{Anisimova2002Accuracy} in that we use simulated protein
@@
 considering each site to be an independent hypothesis test producing a
-positive or negative result. Our experimental design also shares some
-features with Fletcher and Yang's \citeyearpar{Fletcher2010Effect}
-investigation of the performance of branch-site method for detecting
-genewise positive selection under alignment error. Like Fletcher and
-Yang \citeyearpar{Fletcher2010Effect}, we simulate codon alignments
-with indels and calculate inferred alignments using a small but
-diverse sample of aligners. Unlike their study, here we focus on \sw
-detection of positive selection occurring throughout a phylogeny (as
-opposed to genewise detection of selection acting at specific
-branches), we explore a wider range of plausible tree sizes, indel
-rates and divergence levels, and we evaluate the impact of a number of
-alignment filtering methods on the \sw analysis. Given the recent
-proliferation of completed and planned vertebrate genomes, we pay
-special attention to choosing simulation parameters and analysis
-methods similar to those commonly encountered in the \sw analysis of
-vertebrate gene families.
+positive or negative result. \textcolor{red}{As we simulate codon
+  alignments with indels and calculate inferred alignments using a
+  small but diverse sample of aligners and alignment filters, the
+  experimental design also shares some features with Fletcher and
+  Yang's \citeyearpar{Fletcher2010Effect} investigation of the
+  performance of the branch-site method for detecting positive
+  selection under alignment error and Privman et al.'s
+  \citeyearpar{Privman2011Improving} study of the effect of alignment
+  filtering on detecting positive selection in HIV-1 genes. However,
+  the present work differs from previous studies in a number of
+  important ways.}
+
+\textcolor{red}{Unlike Fletcher and Yang, here we focus on \sw
+  detection of sitewise positive selection occurring throughout a
+  phylogeny (as opposed to genewise detection of selection acting at
+  specific branches) and we evaluate the impact of a number of
+  alignment filtering methods on the \sw analysis.}
+
+\textcolor{red}{Privman et al., whose simulations centered on the
+  evolutionary chracteristics of three HIV-1 genes, concluded that
+  alignment filtering improves the performance of positive selection
+  inference by removing false positives. While this may be true for
+  the three genes (and perhaps for HIV-1 in general), HIV-1 is known
+  to evolve with widespread positive selection in the human host
+  \citep{Yang2003Widespread}. Results valid for these three HIV-1
+  genes may not be widely applicable to large-scale vertebrate and
+  mammalian comparative datasets, which exhibit less adaptive
+  evolution \citep{Kosiol2008Patterns} and which comprise a larger
+  diversity of protein structures and a wider range of species
+  divergence levels.}
+
+\textcolor{red}{We hypothesize that these factors may strongly affect
+  the the performance of methods for alignment and detection of
+  selection. Accordingly, our simulations employ parameters similar to
+  those commonly encountered in the \sw analysis of vertebrate gene
+  families and encompass a wide range of biologically plausible indel
+  rates and divergence levels. Although our main purpose is to
+  understand the behaviour of various aligners and filters within our
+  range of simulations, we will make use of the above-noted
+  methodological similarities by pointing out those aspects of our
+  study which agree with or contradict the literature where possible.}
 
@@
 allowing individual residues to be masked instead of entire
-columns. GUIDANCE generates many replicate alignments, each using a
+columns. \textcolor{red}{Privman et al. \citeyearpar{Privman2011Improving} found a
+residue-based filter to be more effective than its column-based
+equivalent, so we filtered residues instead of alignment columns where
+possible}. GUIDANCE generates many replicate alignments, each using a
 slightly perturbed guide tree, with either MAFFT or \pranka as the
@@
 \caption{Sequences were simulated with indels using one of three tree
-  shapes (t-, 17- or 44-taxon) and a range of indel rates and mean
+  shapes (6-, 17- or 44-taxon) and a range of indel rates and mean
   path length (MPL) divergence levels. Alignments were inferred with
@@
 \bibliographystyle{mbe3} \fontsize{9}{10}\selectfont% \itemsep 0pt
-
 \bibliography{slrsim}

### 2011-09-12 projects/2xmammals/analyze_mammals.R
Super-duper mammals alignments.

@@ -1,0 +1,538 @@
+library(plyr)
+library(ggplot2)
+source("~/src/greg-ensembl/scripts/mysql_functions.R")
+
+do.all <- function() {
+  
+}
+
+get.genes <- function() {
+  if (!exists('genes.df', envir=.GlobalEnv)) {
+    con <- connect('gj1_2x_63'); 
+    df <- dbGetQuery(con, 'select * from genes')
+    assign('genes.df', df, envir=.GlobalEnv)
+  }
+  df <- get('genes.df', envir=.GlobalEnv)
+  df
+}
+
+get.genes.split <- function() {
+  if (!exists('genes.split.df', envir=.GlobalEnv)) {
+    genes <- get.genes()
+
+    out.df <- data.frame()
+    clades <- c('p', 'g', 'l', 'm', 'e')
+    
+    for (i in 1:4) {
+      clade <- clades[i]
+      bl.s <- paste(clade, '_slr_total_length', sep='')
+      sites.s <- paste(clade, '_sites', sep='')
+      dnds.s <- paste(clade, '_slr_dnds', sep='')
+      ecdf.f <- ecdf(genes[, dnds.s])
+      cum <- ecdf.f(genes[, dnds.s])
+      cur.genes <- genes
+      cur.genes$parameter_set_id <- i
+      cur.genes$clade <- clade
+      cur.genes$dnds <- genes[, dnds.s]
+      cur.genes$sites <- genes[, sites.s]
+      cur.genes$bl <- genes[, bl.s]
+      cur.genes$cum_dnds <- cum
+      out.df <- rbind(out.df, cur.genes)
+    }
+    assign('genes.split.df', out.df, envir=.GlobalEnv)
+  }
+  df <- get('genes.split.df', envir=.GlobalEnv)
+  df
+}
+
+get.sites <- function() {
+  if (!exists('sites.df', envir=.GlobalEnv)) {
+    con <- connect('gj1_2x_63'); 
+    df <- dbGetQuery(con, 'select * from sites limit 4000000')
+
+    # Remove all gaps and 'single char' sites
+    df <- subset(df, !(note %in% c('all_gaps', 'single_char')))
+
+    # Add p-values
+    df[, 'pval'] <- 1 - pchisq(abs(df[, 'lrt_stat']), 1)
+    
+    # Add p-set factor.
[…]

### 2011-09-12 projects/gorilla/create_orthologs_analysis.pl
Add parallel analysis scripts to Gorilla project, refactor go enrichments.

@@ -9,10 +9,10 @@ use Bio::Greg::Hive::HiveLoaderUtils;
 use Bio::Greg::Hive::ComparaHiveLoaderUtils;
 
 # First create a database for the experiment.
-my $url = 'mysql://ensadmin:ensembl@ens-research/gj1_orthologs';
+my $url = 'mysql://ensadmin:ensembl@ens-research/gj1_subtrees_61';
 my $dba = Bio::EnsEMBL::Compara::DBSQL::DBAdaptor->new( -url => $url );
 
-$url = 'mysql://ensadmin:ensembl@ens-research/' . "gj1_orthologs";
+$url = 'mysql://ensadmin:ensembl@ens-research/' . "gj1_subtrees_61";
 my $h = new Bio::Greg::Hive::ComparaHiveLoaderUtils;
 $h->init($url);
 

### 2011-09-12 projects/alnscore/plot_alnscore.R
Alnscore updates.

@@ -24,7 +24,7 @@ scores.scatter <- function() {
   cur.df <- aln
 #  cur.df$n_seqs <- factor(cur.df$n_seqs)
 
-  stats <- c('sub_tcs', 'sub_sps', 'sub_pbl', 'sub_mmbl')
+  stats <- c('sub_tcs', 'sub_sps', 'sub_pbl', 'sub_cbl')
   comb.df <- data.frame()
   for (stat in stats) {
     new.df <- cur.df
@@ -32,21 +32,6 @@ scores.scatter <- function() {
     new.df$stat <- stat
     comb.df <- rbind(comb.df, new.df)
   }
-
-#  pdf(file="mpl_pbl.pdf", width=10, height=5)
-#  p <- ggplot(comb.df, aes(x=mpl, y=y.val, colour=n_seqs))
-#  p <- p + geom_point(size=1, alpha=1)
-#  p <- p + scale_colour_brewer()
-#  p <- p + facet_grid(stat ~ tree)
-#  print(p)
-#  dev.off()
-
-#  pdf(file="total_bl_pbl.pdf", width=10, height=5)
-#  p <- ggplot(comb.df, aes(x=total_bl, y=y.val, colour=n_seqs))
-#  p <- p + geom_point(size=1, alpha=1)
-#  p <- p + facet_grid(stat ~ tree)
-#  print(p)
-#  dev.off()
 
   cur.df <- subset(comb.df, aligner=='prank' & mpl > 0.9 & mpl < 1.1)
   pdf(file="n_seqs_scatter.pdf", width=10, height=5)
@@ -57,7 +42,7 @@ scores.scatter <- function() {
   print(p)
   dev.off()
 
-  cur.df <- subset(comb.df, n_seqs %in% c(2, 32))
+  cur.df <- subset(comb.df, n_seqs %in% c(2))
   cur.df$aligner <- paste(cur.df$aligner, cur.df$n_seqs, sep=' ')
 
   pdf(file="aln_scatter.pdf", width=10, height=10)
@@ -68,22 +53,12 @@ scores.scatter <- function() {
   print(p)
   dev.off()
 
-  pdf(file="aln_scatter_total_bl.pdf", width=10, height=10)
-  p <- ggplot(cur.df, aes(x=log(total_bl/(2*n_seqs-2)), y=y.val, colour=aligner))
-  p <- p + geom_point(size=1, alpha=1)
[…]

### 2011-09-12 projects/slrsim/combined.plot.R
Add some filtering to Slrsim analysis / plots.

@@ -171,11 +171,23 @@ multi.trees <- function() {
 }
 
 multi.filters <- function() {
+
+  _multi.filters(include.mafft=FALSE)
+  _multi.filters(include.mafft=TRUE, file='filt_fig_multi_supp.pdf')
+}
+
+_multi.filters <- function(include.mafft=FALSE, file='filt_fig_multi.pdf') {
   tbl.a <<- read.csv('~/scratch/gj1_fig_three_a/current/table.csv')
   tbl.b <<- read.csv('~/scratch/gj1_fig_three_b/current/table.csv')
 
   tbl.a$aligner <- 'prankc'
   tbl.b$aligner <- 'clustalw'
+
+  if (include.mafft) {
+    tbl.c <<- read.csv('~/scratch101/gj1_fig_three_e/current/table.csv')
+    tbl.c$aligner <- 'mafft'
+  }
+
 
   reorder.f <- function(tbl) {
     print(levels(tbl$filter))
@@ -188,11 +200,17 @@ multi.filters <- function() {
   }
   tbl.a <- reorder.f(tbl.a)
   tbl.b <- reorder.f(tbl.b)
+  if (include.mafft) {
+    tbl.c <- reorder.f(tbl.c)
+    tbl.c$aln.filt <- factor(tbl.c$filter, labels=paste(tbl.c[1,]$aligner, levels(tbl.c$filter), sep=' '))
+  }
 
   tbl.a$aln.filt <- factor(tbl.a$filter, labels=paste(tbl.a[1,]$aligner, levels(tbl.a$filter), sep=' '))
   tbl.b$aln.filt <- factor(tbl.b$filter, labels=paste(tbl.b[1,]$aligner, levels(tbl.b$filter), sep=' '))
   #tbl.b$aln.filt <- paste(tbl.b$aligner, tbl.b$filter, sep=' ')
+  
   tbl.a <- rbind(tbl.a, tbl.b)
+  
   #print(head(tbl.a))
 
   tbl.a$tdr_at_thresh <- 1 - tbl.a$fdr_at_thresh

### 2011-09-12 projects/slrsim/hmmer_test/hmmer_phylosim.R
A bit of tikering with Hmmer phylosim

@@ -1,8 +1,10 @@
+library(R.oo)
+library(ape)
+library(ggplot2)
 source("~/src/greg-ensembl/projects/phylosim/PhyloSimSource.R")
-source("~/src/greg-ensembl/projects/phylosim/Plots.R")
+source("~/src/greg-ensembl/projects/phylosim/PhyloSimPlots.R")
 
 read.hmmer <- function(file) {
-
   lines <- readLines(file)
   lines.toks <- strsplit(lines, "\\s+")
 
@@ -69,7 +71,7 @@ conv.probs <- function(x) {
 
 profile.to.phylosim.seq <- function(profile) {
   n.sites <- length(profile)
-  #n.sites <- 20
+  #n.sites <- 10  # Shorten the profile for quicker testing...
 
   seq <- CodonSequence(length=n.sites)
   gy <- GY94(kappa = 2)
@@ -90,10 +92,12 @@ profile.to.phylosim.seq <- function(profile) {
   sites.list <- list()
   process.list <- list()
   for (i in 1:n.sites) {
-    print(i)
+    print(paste(i, '/',n.sites))
     cur.site <- profile[[i]]
     
     # Create a separate GY94 process
+    # Is it faster to clone or to create a new process?
+    #cur.gy <- GY94(kappa = 2)
     cur.gy <- clone(gy)
 
     # Create sitewise insertors and deletors.
@@ -151,8 +155,6 @@ profile.to.phylosim.seq <- function(profile) {
   #print(getDeletionTolerance(seq, del))
 
   sampleStates(seq)
-  print(seq)
-  print(Translate(seq, as.character(seq)))
   return(seq)
 }
 
@@ -172,38 +174,132 @@ aa.probs.to.codon.freqs <- function(aa.probs, codons, aas) {
   return(codon.freqs)
 }
 
-do.stuff <- function(pfam.id, phylo=NA, n.leaves=NA) {
-  PSIM_FAST <- TRUE
-
-  if (!is.na(phylo)) {
+load.or.make.tree <- function(tree.file=NA, phylo=NA, n.leaves=NA) {
+  if (!is.na(tree.file)) {
+    tree <- read.tree(tree.file)
+  } else if (!is.na(phylo)) {
     tree <- phylo
   } else {
     tree <- rcoal(n.leaves, br=runif)
     tree.len <- max.length.to.root(tree)
[…]

### 2011-09-12 projects/slrsim/paper/notes.tex

\documentclass{article}
\usepackage{xspace,bm,threeparttable,graphicx,dblfloatfix,multirow,array}
\usepackage[a4paper,hmargin=2.5cm,vmargin=2.5cm]{geometry}

\newcommand{\omg}{\bm{\omega}\xspace}

\begin{document}
\title{Notes on / comparison to Privman et al. alignment filtering paper}
\maketitle

\section{Comparison table}

\begin{table}[h]
\footnotesize \raggedright \tabcolsep 5pt
\begin{tabular}{ m{3cm} m{7cm} m{7cm} }
 & Privman et al.
 & This study
\\
\hline
  Selection method & PAML M7/M8 & SLR, PAML M7/M8\\
  $\omg{}$ distr.
    & 4-category discrete estimated from HIV alignment windows
    & 30-category discrete estimated from mammalian alignments\\
  \% pos-sel sites
    & 4.6\% (gag), 2.9\% (pol), 10.5\% (env)
    & 6\%\\
  Indel distribution
    & Variable rates estimated from HIV alignment windows (mean rate: 0.045 gag, 0.041 pol, 0.139 env)
    & Constant rate per simulation, range from 0 to 0.1\\
  $\omg$ and indel covariation
    & Yes - separate sim parameters for each 20-aa window
    & No - constant indel rate and $\omg$ distribution\\
  Sequence count
    & 20
    & 6, 17, 44\\
  Tree shape
    & Fairly balanced, longest branches at root and tips
    & Balanced 6-species; mammalian / vertebrate shaped 17 and 44\\
  Divergence level
    & (MPL values as input to INDELIBLE; divide by ~1.5 to get dS) 0.497 gag, 0.492 pol, 0.57 env
    & Variable; 0.05 - 2.0 root-to-tip dS\\
  Total branch length (at MPL=0.85)
    & 8.5 gag, 9.1 pol, 9.1 env
    & 3.9 17-taxon, 10.1 44-taxon, 3.74 6-taxon\\
  Simulation replicates
    & 50 replicates (x 3 genes) = 150 alignments
    & 100 replicates per condition, ~20k alignments\\
  Aligners
    & ClustalW, MAFFT, MUSCLE, PRANK (amino acid)
    & ClustalW, MAFFT, PRANK (amino acid), PRANK (codon)\\
  Filters
    & Column-based GUIDANCE, Gblocks, ALISCORE, optimal; residue-based GUIDANCE
    & Column-based Gblocks; residue-based GUIDANCE, T-Coffee, optimal\\
\end{tabular}
\end{table}

\section{Major conclusions from Privman et al.}
\begin{itemize}

\item{Aligners have a strong impact: PRANK works best, then MAFFT/MUSCLE, then ClustalW}
\item{No clear advantage for any of the column-based filtering methods. Residue-based filter showed small improvement (their explanation why: 'reduced loss of information')}
\item{$env$-like simulations yielded worst performance, $pol$-like simulations yielded best performance. Obvious.}
\item{Filtering should be used all the time. From the abstract:
\begin{quote}
Our study shows that the benefit of removing unreliable regions exceeds the loss of power due to the removal of some of the true positively-selected sites.
\end{quote}

From the discussion:

\begin{quote}
Researchers cannot know the FP rate to expect for their data. It is not under control. Conversely, we show that in all three scenarios positive selection can still be detected after filtering. Therefore, filtering provides a more conservative approach...
\end{quote}

}

\end{itemize}

\section{Criticisms of Privman et al.}

\begin{itemize}

\item{Figure 1 - Why does MAFFT perform better than Prank in the pol simulations? Why does Gblocks often yield the best overall performance (e.g., MAFFT+Gblocks is best in the env simulations, slightly better thank Prank+Gblocks and way better than GUIDANCE)}
\item{Figure 2 - Why were MAFFT alignments used instead of Prank, if Prank was discovered to perform the best?}
\item{Figure 3 - again, why use the middle-of-the-road aligner instead of the best one available? What is Prank's performance on the equivalent simulations?}
\item{The 90\% FPR number mentioned in their results section is misleading. It comes from Figure 3 env simulations, which used MAFFT instead of Prank and represents a gene-wide FPR (instead of sitewise FPR). We specifically show that Prank does better than other aligners because it keeps the FPR very low, even at high indel rates (see our Supp. Figure S1B for the best comparison).}

\end{itemize}

\section{How do their simulations fit into our results?}

Although the simulations weren't performed in the exact same way, it seems that the $\omg$ distributions are roughly comparable (though pol has lower, and env has higher, \% pos-sel sites), and gag/pol are around 0.05 mean indel rate, while env is around 0.15 mean indel rate. Divergence is around ~0.3-0.5 dS (there is some uncertainty regarding the exact meaning of their branch lengths). So the divergence is at the low end of our range, and the indel rates are at the medium-to-high end.

Using Privman et al. Figure S3 (ClustalW/gag) we see TPR of ~0.08 at a FPR of 1\%. Our Figure 4c (ClustalW/17-taxon) shows a ~0.3-0.4 TPR at FPR of 1\%, MPL=0.5 and indel=0.05. Our numbers seem to trend higher than Privman et al. for the 'equivalent' simulation.

In terms of the benefit of filtering, we can try to compare our results to their gag simulations (which had a $\omg$ distribution most similar to ours). With ClustalW alignments and Gblocks filtering they find a ~30\% improvement in TPR at 5\% FPR, while we saw no change (MPL=0.4, indel=0.08). For GUIDANCE filtering they find ~100\% improvement, while we saw ~10\% improvement. With Prank alignments, they saw ~10-20\% improvement for all methods. We saw no change for all methods.

Some of this difference could be that they used amino-acid Prank while we used codon Prank. Also, the difference between 1\% and 5\% FPR for the filtering measure could impact things.

More generally, it is plausible that the clustering of regions with high indel rates and positive selection is causing their alignments to be more 'difficult' overall, thus putting their equivalent position in our parameter space further towards the upper-right. Still, even with this explanation a number of inconsistencies remain (see Criticisms).

\section{What to emphasize about our study versus Privman et al. / Fletcher and Yang}

\begin{itemize}

\item{We separated false positive and false negative rates using a constant SLR threshold -- this allowed to separately analyze the sources of error, yielding the insight that all aligners suffer from false negatives, and Prank suffers from fewer false positives.}
\item{We identified consistent trends across a wide range of indel rates / divergence levels}
\item{We measured the impact of filtering when applied to the worst and best aligners, instead of using a middle-performing aligner that should not be used in practice by anyone worried about misalignment.}

\end{itemize}

\section{To-do}

\begin{itemize}

\item{Update introduction with some mention of Privman et al. and why our study differs from theirs}
\item{Add some comparison to Results / Conclusion to Privman et al.'s results}
\item{Subtle changes to manuscript (title / abstract / conclusion) emphasizing the unique / novel aspects of this study}

\end{itemize}

\end{document}

### 2011-09-12 projects/slrsim/paper/slrsim.tex

@@
 \usepackage{color}
-%\usepackage{ulem}
 
@@
 \newcommand{\tpr}{TPR$_{1\%}$\xspace}
+\newcommand{\tprf}{TPR$_{5\%}$\xspace}
 \newcommand{\sw}{sitewise\xspace}
@@
 \title{The effects of alignment error and alignment filtering on the \sw detection of positive selection}
-%%\title{The effects of alignment error and alignment filtering on the \sw detection of positive selection in mammals}
-%\title{The effects of alignment error and alignment filtering on the \sw detection of positive selection in vertebrate genomics}
+%%\title{The effects of alignment error and alignment filtering on the \sw detection of positive selection in vertebrates}
 \author{{Gregory Jordan and Nick Goldman}}
@@
 for aligners and filters within a biologically plausible parameter
-space. Controls using the true alignment and an optimal filtering
+space, and controls using the true alignment and an optimal filtering
 method suggested that performance improvements could be gained by
@@
 phylogenetic trees than unfiltered alignments
-\citep{Dessimoz2010Phylogenetic}. With the application of published
+\citep{Dessimoz2010Phylogenetic}. \textcolor{red}{A recent study suggested that the
+benefit of alignment filtering (in terms of improved accuracy)
+outweighs the cost (in terms of reduced power) when applied to
+detecting positive selection, but the analysis was limited to a small
+range of possible evolutionary scenarios, as discussed below
+\citep{Privman2011Improving}.}  With the application of published
 filtering methods to alignments before testing for positive selection
 becoming standard practice
-\citep{Studer2008Pervasive,Aguileta2009Rapidly}, an analysis of the
-benefits of alignment filtering to the detection of positive selection
+\citep{Studer2008Pervasive,Aguileta2009Rapidly}, \textcolor{red}{continued investigation into the}
+potential benefits of alignment filtering to the detection of positive selection
 would seem well-warranted.
@@
 power of \sw evolutionary inference of positive selection. Our
-approach builds from that of Anisimova et al.
+approach \textcolor{red}{is modeled on that of} Anisimova et al.
 \citeyearpar{Anisimova2002Accuracy} in that we use simulated protein
@@
   selection under alignment error and Privman et al.'s
-  \citeyearpar{Privman2011Improving} study of the effect of alignment
+  study of the effect of alignment
   filtering on detecting positive selection in HIV-1 genes. However,
@@
   specific branches) and we evaluate the impact of a number of
-  alignment filtering methods on the \sw analysis.}
+  alignment filtering methods on the \sw analysis. Thus, the
+  biological hypothesis being investigated is different, and the
+  addition of filtering methods adds another layer of complexity to
+  our analysis.}
 
@@
 
-\textcolor{red}{We hypothesize that these factors may strongly affect
-  the the performance of methods for alignment and detection of
-  selection. Accordingly, our simulations employ parameters similar to
-  those commonly encountered in the \sw analysis of vertebrate gene
-  families and encompass a wide range of biologically plausible indel
-  rates and divergence levels. Although our main purpose is to
-  understand the behaviour of various aligners and filters within our
-  range of simulations, we will make use of the above-noted
-  methodological similarities by pointing out those aspects of our
-  study which agree with or contradict the literature where possible.}
+\textcolor{red}{We hypothesize that the divergence level and indel
+  rate, two highly variable and important evolutionary factors, may
+  strongly affect the performance of methods for alignment and
+  detection of selection. Accordingly, our simulations encompass a
+  wide range of biologically plausible indel rates and divergence
+  levels while fixing other parameters at values typical to those
+  encountered in the \sw analysis of vertebrate gene
+  families. Although our main purpose is to understand the behaviour
+  of various aligners and filters within our range of simulations, we
+  will make use of the above-noted methodological similarities in
+  order to highlight aspects of our results which agree with or
+  contradict the literature where possible.}
 
@@
 of \sw selective pressures estimated from mammalian gene trees
-(\citeyearpar{LindbladToh2011High}; log-normal parameters shown in Table
+(Lindblad-Toh et al. \citeyear{LindbladToh2011High}; log-normal parameters shown in Table
 \ref{table_1}). This distribution, with mean $\omg$ of 0.28 and 6\% of
@@
 performance characteristics, with \prankc always performing best and
-ClustalW performing worst. The same ranking of aligners with respect
-to detecting positive selection was found by Fletcher and Yang
-\citeyearpar{Fletcher2010Effect}; our results corroborate their
-finding and provide evidence that this ranking may be applicable to a
-wider array of evolutionary analyses.
+ClustalW performing worst. \textcolor{red}{The same ranking of
+  aligners with respect to detecting positive selection has been
+  observed in a number of studies
+  \citep{Fletcher2010Effect,Markova-Raina2011High,Privman2011Improving};
+  our results corroborate these findings and provide evidence that
+  this ranking may be applicable across a wide range of divergence
+  levels and indel rates.}
 
@@
 As alignment filters act through the removal of alignment residues or
-columns, a certain amount of reduction in the FPR and TPR was expected
+columns, a certain amount of reduction in both the FPR and TPR was expected
 purely from the decreased amount of information available. For
@@
 filtering performance. Additionally, the prevalence of false negatives
-resulting from misalignment suggested that alignment filters may also
+resulting from misalignment suggested the interesting possibility that alignment filters may also
 improve power by removing false negatives, perhaps by masking out
@@
 achieved by a perfect alignment filter. One exception to the general
-trend was the observation of two simulation conditions with slightly
-elevated FPRs in the true alignment compared to \prankc alignments (at
-an indel rate of 0.04 and MPL of 0.8 and 2.0). This small
-inconsistency may be explained by stochastic variation in false
-positive counts, as the absolute value of the FPR was very low in both
-datasets under those conditions (on the order of $5\times10^{-4}$,
-Figure \ref{fig_4}B). A similar slight FPR elevation was also observed
-at the same indel rate for the optimal, GUIDANCE and T-Coffee filters.
+trend of lower FPR in the true alignment was the observation of two
+simulation conditions with slightly elevated FPRs in the true
+alignment compared to \prankc alignments (at an indel rate of 0.04 and
+MPL of 0.8 and 2.0). This small inconsistency may be explained by
+stochastic variation in false positive counts, as the absolute value
+of the FPR was very low in both datasets under those
+conditions. \textcolor{red}{Figure \ref{fig_4}B shows the FPR to be on
+  the order of $5\times10^{-4}$ under those conditions, and we
+  observed 63 false positives for the true alignment and 67 false
+  positives for the \prankc alignment across all 150 replicates at the
+  indel rate of 0.04 and MPL of 0.8}. A similar slight FPR elevation
+was also observed at the same indel rate for the optimal, GUIDANCE and
+T-Coffee filters.
 
@@
 MPL$>$1.6), well above those found in commonly-analyzed groups of
-species. Thus, the use of unfiltered \prankc alignments would yield
-the best performance for detecting \sw positive selection when
-analyzing protein-coding sequences across a wide range of indel rates
-and commonly encountered sequence divergence levels. The performance
-of the `optimal' filter on \prankc alignments suggests that mild
-further improvements to filtering strategies may be possible, but the
-potential for improvement is small and may be of little value.
+species. \textcolor{red}{Thus, the use of unfiltered \prankc alignments would yield
+largely equivalent performance to GUIDANCE-filtered alignments for detecting
+\sw positive selection when analyzing protein-coding sequences across
+a wide range of indel rates and commonly encountered sequence
+divergence levels.
+On the other hand, GUIDANCE (when judiciously applied, including an upper limit on the amount of sequence data removed) never significantly reduced performance compared to the unfiltered alignments.} Finally, we note that the performance of the `optimal' filter on \prankc
+alignments suggests that mild further improvements to filtering
+strategies may be possible, but the potential for improvement is small
+and may be of little value.
 
@@
 divergence levels, using simulation parameters designed to approximate
-the analysis of mammalian and other typical protein-coding genes. We
+the analysis of typical mammalian protein-coding genes. We
 evaluated the ability of four alignment methods and three alignment
@@
 alignment conditions. We confirmed and extended the findings of
-Fletcher and Yang \citeyearpar{Fletcher2010Effect} regarding the
+Fletcher and Yang \citeyearpar{Fletcher2010Effect}, Markova-Raina et al. \citeyearpar{Markova-Raina2011High}, and Privman et al. \citeyearpar{Privman2011Improving} regarding the
 relative accuracy of different aligners, showing that \prankc had the
@@
 reduces the beneficial effect of alignment filtering, so much so that
-current filtering methods are essentially unable to improve the
+current filtering methods are \textcolor{red}{scarcely able to} improve the
 performance of \prankc alignments when analyzed with SLR.
 
-As a result, among the aligners and filters tested here, we recommend
-the use of unfiltered \prankc alignments for the detection of sitewise
-positive selection in all but the most divergent phylogenies. GUIDANCE
-showed some ability to improve sitewise power under difficult
-alignment conditions at high divergence levels, but this improvement
+As a result, we cannot unequivocally recommend the use of alignment
+filtering in the detection of sitewise positive selection, except
+perhaps for the most divergent and indel-prone sequences, which are
+unlikely to be encountered in analyses focusing on mammalian or
+vertebrate genes. Although GUIDANCE showed some ability to improve the
+error-controlled power under difficult alignment conditions at high
+divergence levels (whereas T-Coffee and Gblocks failed to improve upon
+any \prankc alignments) this improvement was modest in magnitude and
 was achieved largely through the recovery of false negatives as
@@
 (as performed by GUIDANCE) may not be worth the possibility of a
-slight increase in power. The development of efficient and effective
-alignment filters, particularly with an eye towards recovering false
-negatives as identified in this study, will be an interesting area for
-future research.
+slight increase in power. On the other hand, judicious application of
+GUIDANCE never significantly reduced power, so its use would not be
+expected to yield worse results.
+
+\textcolor{red}{Some of our results and conclusions differed from those of Privman et
+al. \citeyearpar{Privman2011Improving}, who found strong improvements
+in error-controlled power by filtering alignments in simulations
+focused on three HIV-1 genes. Although the phylogenetic trees they
+used to guide their simulations contained divergence levels at the low
+end of our tested range (MPLs of 0.38, 0.34 and 0.33 for gag, pol and
+env, respectively; E. Privman, personal communication) and were roughly
+comparable in size and shape to our 17-taxon tree, we failed to find a
+significant benefit for alignment filters at any MPL below 1.2 when
+using \prankc alignments. Interestingly, the authors also found
+Gblocks to be roughly comparable in performance to GUIDANCE, while we
+found Gblocks' performance to be very poor. Some of these
+discrepancies may be due to differences in the details of our
+simulations or filtering procedures, but in the end our results are
+largely complementary: Privman et al. showed that filtering can be
+beneficial for detecting positive selection, especially in the case of
+fast-evolving (but fairly closely-related) sequences, while we have
+shown that divergence levels and indel rates have a significant impact
+on the accuracy and sensitivity of different aligners.}
 
@@
 and filters. The trends we observed were consistent across a wide
-range of parameter values, suggesting that they reflect fundamental
-differences in each method's ability to align or filter sequences as
-opposed to artifacts due to the relative simplicity of our
-simulations.
+range of parameter values and tree sizes, suggesting that they reflect
+fundamental differences in each method's ability to align or filter
+sequences as opposed to artifacts due to the relative simplicity of
+our simulations.
 
@@
 
-GJ was funded by a Gates Cambridge Trust Scholarship and is a member
-of Darwin College, University of Cambridge.
+The authors would like to thank Eyal Privman for providing HIV-1 tree
+data and useful discussion. GJ was funded by a Gates Cambridge Trust
+Scholarship and is a member of Darwin College, University of
+Cambridge.
 

### 2011-09-12 projects/slrsim/paper/supp_slrsim.tex

@@
 
+\newcommand{\mtwo}{PAML M2a\xspace}
+\newcommand{\meight}{PAML M8\xspace}
 \newcommand{\pranka}{PRANK$_{\textrm{AA}}$\xspace}
+\newcommand{\prankc}{PRANK$_{\textrm{C}}$\xspace}
+\newcommand{\omg}{\bm{\omega}\xspace}
+\newcommand{\tpr}{TPR$_{1\%}$\xspace}
+\newcommand{\tprf}{TPR$_{5\%}$\xspace}
+\newcommand{\sw}{sitewise\xspace}
+\newcommand{\Sw}{Sitewise\xspace}
+\newcommand{\Dr}{{\emph{Drosophila}}\xspace}
 
@@
 %\includegraphics[scale=0.65]{supp_fig1.pdf}
-\includegraphics[scale=0.8]{supp_fig1.pdf}
+\includegraphics[scale=0.7]{supp_fig1.pdf}
 \end{center}
@@
   have been added to sections (a) and (b) and results using ClustalW +
-  GUIDANCE have been added to all sections.}
+  GUIDANCE and MAFFT + GUIDANCE have been added to all sections.}
 \label{supp_fig1}
@@
 
+\begin{figure}[t]
+\begin{center}
+%\includegraphics[scale=0.65]{supp_fig1.pdf}
+\includegraphics[scale=0.55]{supp_fig2.pdf}
+\end{center}
+\caption{This figure depicts largely the same simulations and uses the
+  same formatting as Figure 5 except for two changes. First,
+  results for MAFFT have been added to section (b) and results for
+  \prankc have been moved to section (c). Second, the rightmost column
+  has been added, showing the true positive rate (TPR) at a 5\% false
+  positive rate (FPR) threshold (labeled \tprf).}
+\label{supp_fig2}
+\end{figure}
+
 \end{document}

### 2011-12-05 projects/slrsim/paper/response.txt
Add files from final version.

Associate Editor: Holland, Barbara

(0A) It might be nice to cite the David Morrison paper "A framework for phylogenetic sequence alignment" as it has a nice discussion of the different reasons people make alignments and makes the good point that the best alignment software for downstream phylogenetic inference might be different from the best aligner for doing homology searches, or in this case detecting positive selection.

>> Response 0A: We have added a sentence highlighting that aligners may be designed with different downstream applications in mind, citing the aforementioned paper (page 2, red text).

(0B) I wasn't quite clear on the difference between 'over-alignment' and 'non-alignment of homologous codons' on pg 2, line 24-25

>> Response 0B: We have changed the text (page 2, red text) to clarify the situation being explained in that paragraph. The goal was to concisely explain two plausible scenarios for why how false negatives (e.g., truly positively-selected sites not identified due to misalignment) might occur.

(0C) on page 7, line 28 do you really mean 'Bayes Empirical Bayes' or is one of the Bayes unnecessary?

>> Response 0C: Yes, the method implemented by PAML is termed 'Bayes Empirical Bayes', as written in the title of the cited Yang et al. 2005 paper.

(0D) I got myself completely confused in the results when I couldn't find where MPL had been defined and assumed for a page or so that it stood for maximum performance level - I did eventually find that it was mean path length, but it might be worth repeating the definition at the start of this section.

>> Response 0D: Good point -- this relatively non-standard acronym is now redefined at the start of the Results section (page 9, red text).

Reviewer: 1

(1A) My only observation is that positive selection is often inferred using branch-site models. That is, it is being tested for a specific tree branch. It could have been interesting to test the effect of the alignment program and filtering on this type of inference.

>> Response 1A: We agree that branch-site models are popular and potentially more powerful for detecting certain adaptive events. However, alignment error in branch-site models was covered at some depth by Fletcher and Yang (2010, first cited on p. 3 of our manuscript). Given our result on sites-based models which showed that filtering has a limited ability to improve the power over unfiltered PRANK alignments, we wouldn't necessarily expect filtering to show a significantly different result for the branch-site model. On the more practical side, we should note that a significant amount of computation was already required to run this extensive set of simulations using the relatively quick SLR method -- to explore an equivalent parameter space evaluating the branch-site test (which is notoriously slow for large trees and alignments) would be rather prohibitive. We feel that any further extension of our approach to branch-site models is better left to a separate dedicated study.

In the submitted manuscript, we did briefly compare our results to those from Fletcher and Yang in the third full paragraph on p.16, noting that the false positive rates from PRANK were higher in Fletcher and Yang's branch-site tests than in our simulations under roughly similar conditions. We have added a sentence to the end of that paragraph to further emphasize the point (page 17, red text).

Reviewer: 2

(2A) I only want to raise two minor points.  First, the number of aligners that PRANK is tested against is limited to MaFFT, and ClustalW which is popular but known to be suboptimal.  I think the current paper is interesting and useful as it is, but its impact would be enhanced if the set of aligners were more comprehensive, and include e.g. ProbCons or T-Coffee.

>> Response 2A: We focused on MAFFT, ClustalW and PRANK in the manuscript in order to frame the discussion in terms of aligners which (in our hands, for detecting sitewise positive selection) showed the greatest range of performance. However, as the reviewer points out, the inclusion of additional aligners such as the consistency-based ProbCons or T-Coffee might provide further benefit, and so we have added results from those two aligners (ProbCons and T-Coffee) in Supplementary Figure 1. We chose to limit the discussion of those aligners in the main text to avoid increasing the length of the manuscript, but we mention their overall performance in the Results section (Pages 10 and 12, red text) and note that the complete set of results can be found in Supplementary Figure S1.

(2B) Second, the authors themselves point out that the assumption of a uniform indel rate is not a good reflection of reality (p. 17 line 40).  Their dismissal of the importance of this feature is not backed up by data, and is not in keeping with the thoroughness of the rest of the paper.  Clustered indels in regions under systematic reduced constraint may lead to increased rates of alignment errors in already more divergent sequence, conceivably leading to higher rates of spurious postitive selection inferences.

>> Response 2B: It was not our intent to come across as dismissive of the importance of the variation in rates of insertion and deletion in proteins (and covariation with selection pressures). We believe this type of heterogeneity is extremely important in the analysis and interpretation real data. However, we deliberately decided not to include indel-rate heterogeneity in our simulations for two main reasons:

  (1) The correlation between distribution of selective constraints and indel rates is not well understood. Privman et al. took an interesting approach to this problem, estimating the dN/dS distribution and indel rate in sliding windows along the length of HIV-1 proteins, but it is unclear to what extent this pattern (which in the case of HIV-1 did reveal a situation similar to that mentioned by the reviewer) generalizes across all proteins. Thus, we did not feel confident basing our entire simulation strategy on a potentially misleading or non-representative model of heterogeneity.

  (2) The use of uniform indel rates also has an advantage in the interpretation of results. Even with a good understanding of patterns of heterogeneity, the power and error rates resulting from simulations incorporating heterogeneity across the length of the sequence are less widely applicable than the results from uniform simulations. This is because the FPR estimated from a model with a specific pattern of indel rate variation is only relevant for proteins sharing that specific pattern, as many of the FPs likely came from those regions under relaxed constraint. On the other hand, our results based on uniform indel rates characterize the general performance of each aligner at a fixed alignment "difficulty". The matrix of FPRs and TPRs could be used as a look-up table to guide researchers analyzing heterogeneous proteins: if a researcher suspects that a part of her protein of interest is evolving with a high rate of insertion and deletion, our results calculated at that high indel rate could be taken as an upper bound on the overall error rate of sitewise analysis on her protein.

We have expanded and refined the relevant text in the Conclusions section (page 18, red text), making more explicit our deliberate decision not to include indel rate heterogeneity in our simulations and indicating how one might use our matrix of error rates and power levels to qualitatively assess the expected performance in a protein with segments evolving under a variety of evolutionary patterns.

### 2011-12-05 projects/slrsim/paper/739_supp_nographics.tex
Add files from final version.

--- 2011-09-12 projects/slrsim/paper/supp_slrsim.tex
@@
 \documentclass{article}
-\usepackage{natbib,graphicx,dblfloatfix,multirow,array,threeparttable,xspace}
-\usepackage[landscape,hmargin=1.5cm,vmargin=1.5cm]{geometry}
+\usepackage{natbib,graphicx,dblfloatfix,multirow,array,threeparttable}
+\usepackage{xspace,bm,amsmath,amssymb}
+\usepackage[nolists,tablesfirst,nomarkers]{endfloat}
+\usepackage[landscape,hmargin=2.5cm,vmargin=2.5cm]{geometry}
+\usepackage{color}
+\usepackage{booktabs}
+\usepackage{multirow}
+
+\definecolor{red}{rgb}{1,0,0}
+
+\linespread{1.5}
+
+\setlength{\bibsep}{0.0pt}
+
+\begin{document}
+
+\renewcommand\refname{Literature Cited}
 
@@
 \newcommand{\Dr}{{\emph{Drosophila}}\xspace}
-
-\begin{document}
+\newcommand{\tcr}[1]{#1}
+\newcommand{\tcrr}[1]{\textcolor{red}{#1}}
+\newcommand{\nsyn}{nonsynonymous\xspace}
+\newcommand{\syn}{synonymous\xspace}
+\newcommand{\nh}{non-homologous\xspace}
+%\newcommand{\hidegraphics}[2]{\includegraphics[#1]{#2}}
+\newcommand{\hidegraphics}[2]{}
 
@@
 \begin{center}
-%\includegraphics[scale=0.65]{supp_fig1.pdf}
-\includegraphics[scale=0.7]{supp_fig1.pdf}
+\hidegraphics{scale=0.65}{supp_fig1.pdf}
 \end{center}
@@
   formatting as Figure 4, except that results for MAFFT and \pranka
-  have been added to sections (a) and (b) and results using ClustalW +
-  GUIDANCE and MAFFT + GUIDANCE have been added to all sections.}
+  have been added to sections (A) and (B) and results using ProbCons,
+  T-Coffee, ClustalW + GUIDANCE, and MAFFT + GUIDANCE have been added
+  to all sections.}
 \label{supp_fig1}
@@
 \begin{center}
-%\includegraphics[scale=0.65]{supp_fig1.pdf}
-\includegraphics[scale=0.55]{supp_fig2.pdf}
+
+\hidegraphics{scale=0.55}{supp_fig2.pdf}
 \end{center}
@@
   same formatting as Figure 5 except for two changes. First,
-  results for MAFFT have been added to section (b) and results for
-  \prankc have been moved to section (c). Second, the rightmost column
+  results for MAFFT have been added to section (B) and results for
+  \prankc have been moved to section (C). Second, the rightmost column
   has been added, showing the true positive rate (TPR) at a 5\% false

### 2012-02-06 projects/2xmammals/aln.R
Various updates.

@@ -1,0 +1,353 @@
+aln.restrict.to.tree <- function(aln, tree) {
+  tip.seqs <- tree$tip.label
+  aln.restrict.to.seqs(aln, tip.seqs)
+}
+
+aln.restrict.to.seqs <- function(aln, seqs) {
+  aln.names <- rownames(aln)
+  overlap <- intersect(aln.names, seqs)
+  aln <- as.matrix(aln[which(aln.names %in% overlap),])
+  aln
+}
+
+aln.remove.allgap.columns <- function(aln) {
+  nongap.cols <- apply(aln, 2, function(x) {
+    sum(!grepl('(---|-)', x))
+  })
+  
+  aln[, nongap.cols > 0]
+}
+
+aln.remove.ngap.columns <- function(aln, n=2) {
+  nongap.cols <- apply(aln, 2, function(x) {
+    sum(!grepl('(---|-)', x))
+  })
+  
+  aln[, nongap.cols >= n]
+}
+
+aln.remove.anygap.columns <- function(aln) {
+  gap.cols <- apply(aln, 2, function(x) {
+    sum(grepl('(---|-)', x))
+  })
+  
+  aln[, gap.cols == 0]
+}
+
+aln.is.codon <- function(aln) {
+  aln <- aln.tx(aln, n.char='Z')
+  if (any(aln == 'X')) {
+    return(FALSE)
+  } else {
+    return(TRUE)
+  }
+}
+
+aln.tx <- function(aln, n.char='X') {
+  library(phylosim)
+  assign("PSIM_FAST", TRUE, envir=.GlobalEnv)
+  # Translates a codon alignment into proteins.
+  ca <- CodonAlphabet()
+
+  char.length <- stringLength(aln[1,1])
+  if (char.length == 1) {
+    pep.aln <- matrix(nrow=nrow(aln), ncol=ncol(aln)/3)
+  } else {
+    pep.aln <- matrix(nrow=nrow(aln), ncol=ncol(aln))
+  }
+  rownames(pep.aln) <- rownames(aln)
+  for (i in 1:nrow(pep.aln)) {
+    codons <- paste(aln[i,],collapse='')
+    for (j in 1:ncol(pep.aln)) {
+      cdna.lo <- (j-1)*3+1
+      cdna.hi <- (j-1)*3+3
+      codon <- substr(codons, cdna.lo, cdna.hi)
+      if (codon == '---') {
+        pep.aln[i,j] <- '-'
[…]

### 2012-02-06 projects/2xmammals/analyze_genes.R

@@ -1,10 +1,12 @@
 uname  <- Sys.getenv("USER")
 if (uname == 'gj1') {
+  source("~/src/greg-ensembl/projects/2xmammals/aln.R")
   source("~/src/greg-ensembl/projects/2xmammals/analyze_mammals.R")
 #  source("~/src/greg-ensembl/scripts/go_enrichments.R")
   source("~/src/greg-ensembl/scripts/liftOver.R")
   source("~/src/greg-ensembl/projects/phylosim/PhyloSimPlots.R")
 } else {
+  source("~/lib/greg-ensembl/projects/2xmammals/aln.R")
   source("~/lib/greg-ensembl/projects/2xmammals/analyze_mammals.R")
 #  source("~/lib/greg-ensembl/scripts/go_enrichments.R")
   source("~/lib/greg-ensembl/scripts/liftOver.R")
@@ -52,38 +54,6 @@ write.aln <- function(x, ff) {
   }
 }
 
-write.all.alns <- function(folder, genes=NULL, ...) {
-  if (is.null(genes)) {
-    genes <- get.genes()
-  }
-  
-  for (i in 1:nrow(genes)) {
-    cur.gene <- genes[i,]
-    gene.name <- cur.gene$gene_name
-    ref.taxon <- cur.gene$ref_taxon_id
-    if (is.na(gene.name)) {
-      gene.name <- cur.gene$ref_gene_id
-    }
-    print(sprintf("%s (%d of %d)", gene.name, i, nrow(genes)))
-    #print(cur.gene)
-    data.prefix <- cur.gene$data_prefix
-
-    aln <- get.aln(cur.gene$data_id, ...)
-
-    dir.create(folder)
-    if (ref.taxon != 9606) {
-      dir.create(file.path(folder, 'no_human'))
-      data.dir <- paste('no_human/', data.prefix, sep='')
-    } else {
-      data.dir <- paste(data.prefix, sep='')
-    }
-    dir.create(file.path(folder, data.dir))
-    aln.f <- paste(file.path(folder, data.dir), '/', gene.name, '.fasta', sep='')
-    write.aln(aln, file=aln.f)
-    print(sprintf("%s (%d of %d)", aln.f, i, nrow(genes)))
-  }
-}
-
 save.alns <- function() {
   con <- connect(dbname())
   cmd <- "select * from genes where ref_taxon_id=9606 and chr_name like '%chr%';"
[…]

### 2012-02-06 projects/2xmammals/analyze_mammals.R

@@ -1,7 +1,7 @@
-library(GenomicRanges)
 library(R.oo)
 library(ape)
 library(plyr)
+
 uname  <- Sys.getenv("USER")
 if (uname == 'gj1') {
   source("~/src/greg-ensembl/scripts/mysql_functions.R")
@@ -36,13 +36,16 @@ scratch.f <- function(f) {
 
 project.f <- function(f) {
   ifebi(
-    paste("~/src/greg-ensembl/projects/2xmammals/", f, sep=''),
-    paste("~/lib/greg-ensembl/projects/2xmammals/", f, sep='')
+    paste("~/lib/greg-ensembl/projects/2xmammals/", f, sep=''),
+    paste("~/src/greg-ensembl/projects/2xmammals/", f, sep='')
   )
 }
 
 scratch <- function() {
-  "~/scratch/gj1_2x_63_alt/current/data/"
+  ifebi(
+    "/homes/greg/scratch/gj1_2x_63_alt/current/data/",
+    "/nfs/users/nfs_g/gj1/scratch/gj1_2x_63_alt/current/data/"
+  )  
 }
 
 get.genes <- function() {
@@ -61,6 +64,26 @@ get.genes <- function() {
     assign('genes.df', df, envir=.GlobalEnv)
   }
   df <- get('genes.df', envir=.GlobalEnv)
+
+  # Add a column for the aln location.
+  df$aln_file <- paste(df$data_prefix, '/', df$gene_name, '.fasta', sep='')
+
+  # Genes with no human gene, or on a bad chromosome, get a chr_name of 'none'.
+  hum <- subset(df, ref_taxon_id==9606)
+  hum <- subset(hum, grepl('chr', chr_name))
+  no.hum <- subset(df, ref_taxon_id!=9606 | !grepl('chr', chr_name))
+  if (nrow(no.hum) > 0) {
+    no.hum$chr_name <- 'none'
+    no.hum$aln_file <- paste('no_human', no.hum$aln_file, sep='/')
+  }
+  
+  if (nrow(hum) == 0) {
+    df <- no.hum
+  } else if (nrow(no.hum) == 0) {
+    df <- hum
+  } else {
+    df <- rbind(hum, no.hum)
+  }
   df
 }
 
@@ -101,6 +124,74 @@ get.genes.split <- function() {
   }
   load(split.genes.f)
   genes.split.df
+}
+
+collect.sites <- function(pset, filter='default', test=F) {
[…]

### 2012-02-06 projects/2xmammals/estimate_trees.R

@@ -1,0 +1,287 @@
+bsub.estimate.m0 <- function() {
+  newicks.df <- get.subset.newicks()
+
+  species.groups <- newicks.df$name
+  #species.groups <- c('Primates')
+
+  subdirs <- get.output.subdirs()
+  subdirs <- c('concat_nongaps_20')
+
+  out.f <- output.folder(T, T, T)
+  for (subdir in subdirs) {
+    for (species.group in species.groups) {
+      for (n.codons in c(1e4, 1e5, 5e5, 1e6, 2e6)) {
+        aln.in <- 'all_combined.fasta'
+        #aln.in <- 'alns_1.fasta'
+        #aln.in <- '../00/EIF3A.fasta'
+        xtra <- paste(out.f, subdir, species.group, aln.in, n.codons, TRUE, sep=' ')
+        bsub.function('estimate_m0_tree', mem=16, extra.args=xtra, queue='long')
+      }
+    }
+  }
+}
+
+bsub.estimate.branch <- function() {
+  newicks.df <- get.subset.newicks()
+
+  species.groups <- newicks.df$name
+  species.groups <- c('Primates')
+
+  subdirs <- get.output.subdirs()
+  subdirs <- c('concat_nongaps_20')
+
+  out.f <- output.folder(T, T, T)
+  for (subdir in subdirs) {
+    for (species.group in species.groups) {
+#      for (n.codons in c(1e4, 1e5, 5e5, 1e6, 2e6)) {
+      for (n.codons in c(1e4, 1e5)) {
+        aln.in <- 'all_combined.fasta'
+        xtra <- paste(out.f, subdir, species.group, aln.in, n.codons, TRUE, sep=' ')
+        bsub.function('estimate_branch_tree', mem=16, extra.args=xtra, queue='long')
+      }
+      return()
+    }
+  }
+}
+
+
+test.estimate.m0 <- function() {
+  base.dir <- output.folder(T, T, T)
+  subdir <- 'concat_nongaps_20'
+  species.group <- 'Primates'
+  #aln.in <- '../00/EIF3A.fasta'
+  aln.in <- 'alns_1.fasta'
+  n.codons <- 1000
+  estimate.m0.tree(base.dir, subdir, species.group, aln.in, n.codons, clean=T)
+}
+
+test.estimate.branch <- function() {
+  base.dir <- output.folder(T, T, T)
[…]

### 2012-02-06 projects/2xmammals/sites_scripts.R

@@ -2,11 +2,17 @@ uname  <- Sys.getenv("USER")
 if (uname == 'gj1') {
   source("~/src/greg-ensembl/projects/2xmammals/analyze_mammals.R")
   source("~/src/greg-ensembl/projects/2xmammals/analyze_genes.R")
+  source("~/src/greg-ensembl/projects/2xmammals/analyze_substitutions.R")
+  source("~/src/greg-ensembl/projects/2xmammals/output_data.R")
+  source("~/src/greg-ensembl/projects/2xmammals/estimate_trees.R")
   source("~/src/greg-ensembl/projects/2xmammals/calc_recomb.R")
   source("~/src/greg-ensembl/scripts/xtable_utils.R")
 } else {
   source("~/lib/greg-ensembl/projects/2xmammals/analyze_mammals.R")
   source("~/lib/greg-ensembl/projects/2xmammals/analyze_genes.R")
+  source("~/lib/greg-ensembl/projects/2xmammals/analyze_substitutions.R")
+  source("~/lib/greg-ensembl/projects/2xmammals/output_data.R")
+  source("~/lib/greg-ensembl/projects/2xmammals/estimate_trees.R")
   source("~/lib/greg-ensembl/projects/2xmammals/calc_recomb.R")
   source("~/lib/greg-ensembl/scripts/xtable_utils.R")
 }
@@ -73,9 +79,16 @@ collect_genes <- function(pset, filter='default', n.indices=0, test=F, subset.in
   print(" done!")
 }
 
-write_alns <- function(chr.s, remove.paralogs=T, mask.clusters=T) {
+write_alns <- function(chr.s, remove.paralogs, mask.clusters, 
+  mask.nongaps, jobs.per.chr, job.index) {
   remove.paralogs <- as.logical(remove.paralogs)
   mask.clusters <- as.logical(mask.clusters)
+  mask.nongaps <- as.logical(mask.nongaps)
+  jobs.per.chr <- as.integer(jobs.per.chr)
+  job.index <- as.integer(job.index)
+
+  # Remove all existing functions and objects
+  rm(list=ls(pattern='get', envir=.GlobalEnv), envir=.GlobalEnv)
 
   print("  loading file...")
   out.f <- scratch.f(sprintf("export_%s.Rdata", chr.s))
@@ -84,98 +97,38 @@ write_alns <- function(chr.s, remove.paralogs=T, mask.clusters=T) {
   print("  writing alns...")
   attach(ensembl_alns)
 
[…]

### 2012-02-06 projects/slrsim/paper/739_main_nographics.tex

--- 2011-09-12 projects/slrsim/paper/slrsim.tex
@@
 \usepackage{color}
+\usepackage{booktabs}
+\usepackage{multirow}
+\usepackage{authblk}
 
@@
 \newcommand{\Dr}{{\emph{Drosophila}}\xspace}
+%\newcommand{\tcr}[1]{\textcolor{red}{#1}}
+%\newcommand{\tcrr}[1]{\textcolor{red}{#1}}
+\newcommand{\tcr}[1]{#1}
+\newcommand{\tcrr}[1]{#1}
+\newcommand{\nsyn}{nonsynonymous\xspace}
+\newcommand{\syn}{synonymous\xspace}
+\newcommand{\nh}{non-homologous\xspace}
+%\newcommand{\hidegraphics}[2]{\includegraphics[#1]{#2}}
+\newcommand{\hidegraphics}[2]{}
 
 \title{The effects of alignment error and alignment filtering on the \sw detection of positive selection}
-%%\title{The effects of alignment error and alignment filtering on the \sw detection of positive selection in vertebrates}
-\author{{Gregory Jordan and Nick Goldman}}
-%\affiliation{European Bioinformatics Institute, Wellcome Trust Genome Campus, Hinxton, Cambridgeshire, CB10 1SD, United Kingdom}}
+\date{October 26, 2011}
+\author[1]{Gregory Jordan}
+\author[1,*]{Nick Goldman}
+\affil[1]{EMBL-EBI, Wellcome Trust Genome Campus, Hinxton, Cambridgeshire, CB10 1SD, United Kingdom}
+\affil[*]{To whom correspondence should be addressed: Nick Goldman, Tel:~+44~(0)1223~492~530; Fax:~+44~(0)1223~494~468; E-mail:~goldman@ebi.ac.uk}
+
 \maketitle
-%\authorlist{Jordan and Goldman}
-%\shorttitle{Effect of Alignment Error on Detecting Sitewise Selection}
-
-%\mbeabstract{% 
+
+\subsection*{Keywords}
+Alignment error, positive selection, codon models, alignment filtering
+
+\subsection*{Running head}
+Alignment error and detecting sitewise selection
+
+\subsection*{Abbreviations used}
+MPL---mean path length; TPR---true positive rate; FPR---false positive rate
+
+\newpage
+
 \section*{Abstract}
-\textcolor{red}{
-Alignment filters are commonly applied to try to reduce errors
-resulting from misalignment, but the prevalence of such errors and the
-ability of filters to reduce them is not well understood.
+When detecting positive selection in proteins, the prevalence of
+errors resulting from misalignment and the ability of alignment
+filters to mitigate such errors are not well understood, but filters
+are commonly applied to try to avoid false positive results.
 % 
@@
 resulted in very few.  False negatives were a problem for all
-aligners, increasing with divergence. Of the aligners tested, PRANK's
+aligners, increasing with sequence divergence. Of the aligners tested, PRANK's
 codon-based alignments consistently performed the best and ClustalW
@@
 found to substantially improve the performance of PRANK alignments
-under most \textcolor{red}{conditions}.
+under most conditions.
 %
@@
 for aligners and filters within a biologically plausible parameter
-space, and controls using the true alignment and an optimal filtering
-method suggested that performance improvements could be gained by
-improving aligners or filters to reduce the prevalence of false
-negatives, especially at higher divergence levels and indel rates.
-}
+space. With the best aligner, a low false positive rate was maintained
+even with extremely divergent, indel-prone sequences. Controls using
+the true alignment and an optimal filtering method suggested that
+performance improvements could be gained by improving aligners or
+filters to reduce the prevalence of false negatives, especially at
+higher divergence levels and indel rates.
+
 %\keywords{Alignment, positive selection, alignment filtering, indels}
-
-%\email{greg@ebi.ac.uk}
-
-%\mbestyle
 
@@
 one might expect alignment error to result mainly in an increased
-number of false positives, as the spurious alignment of nonhomologous
-codons on average would result in a high number of apparent
-non-synonymous substitutions and a low number of synonymous
-substitutions (since a randomly chosen pair of codons are more likely
-to be nonsynonymous than synonymous). However, false negatives may
-also be introduced, either through over-alignment of synonymous codons
-(increasing the synonymous substitution rate) or non-alignment of
-homologous codons (causing reduced power due to less evolutionary
-information). Since different aligners employ a variety of algorithms,
-evolutionary models, and heuristic optimizations
+number of false positives, as the spurious alignment of \nh codons on
+average would result in a high number of apparent \nsyn substitutions
+and a low number of \syn substitutions (since two randomly chosen
+codons are more likely to be \nsyn than \syn).  \tcr{However, false
+  negatives may also be introduced, either through the introduction of
+  \syn but \nh codons into a positively-selected site (thus reducing
+  power due to an inflated \syn substitution rate) or through the
+  failure to align truly homologous codons at a positively-selected
+  site (reducing power due to less evidence for positive selection at
+  that site).}  Since different aligners employ a variety of
+algorithms, evolutionary models, and heuristic optimizations
 \citep{Notredame2007Recent}, each program may be more or less prone to
@@
 variations in the nature and magnitude of its impact on the detection
-of positive selection.
-
-\input{table1.txt}
-
-The protein structure and the evolutionary divergence of a dataset may
-also contribute to the effects of alignment error. Differently
-structured protein regions show variable tolerance to biological
-indels, with indels more common in extracellular and transmembrane
-proteins than in highly folded enzymes and housekeeping genes
-\citep{delaChaux2007DNA}. This suggests that well-folded protein
-regions will experience fewer biological indels---and will therefore
-be less susceptible to alignment error---than less structured
-regions.
+of positive selection. \tcr{Different aligners may also be designed
+  for different downstream applications, such as phylogenetic
+  inference or functional annotation \citep{Morrison2009}, making the
+  optimal choice of aligner potentially dependent on the way in which
+  the resulting alignment will be used. Here we focus on the sitewise
+  detection of positive selection.}
+
+\begin{table}
+\caption{Parameter Values Used in Simulations}
+\begin{threeparttable}
+\footnotesize
+\begin{tabular}{lllllllll}
+\toprule
+ \multicolumn{3}{c}{Tree} & \multicolumn{3}{c}{Insertions and Deletions} & \multicolumn{3}{c}{$\omega$ Distribution} \\
+\cmidrule(r){1-3} \cmidrule(r){4-6} \cmidrule(r){7-9}
+ & & & Size & Mean Length & & & & \\
+Taxa & Source & MPL & Distribution & (Std. Dev.) & Rate & Shape & Mean & $p (\omega>1)$ \\
+\midrule
+6 & Artificial & & power law & & & lognormal & & \\
+17 & $\beta$-globin & 0.05--2.0 & decay: 1.8 & 3.33 (5.51) & 0--0.2 & log mean: -1.864 & 0.277 & 0.06 \\
+44 & Vertebrates & & max length: 40 & & &  log SD: 1.201 & & \\
+\bottomrule
+\label{table_1}
+\end{tabular}
+\begin{tablenotes}
+\item [] N{\sc ote}.---{\hskip -.1pt}MPL is the mean path length of the tree in units of substitutions per synonymous site.
+Indel lengths are measured in units of codons, and the indel rate is defined as the number of insertion \& deletion events per substitution.
+\end{tablenotes}
+\end{threeparttable}
+\end{table}
+
+\tcr{In addition, we expect the protein structure and the
+  evolutionary divergence of a dataset to contribute to the effects of
+  alignment error}. Differently structured protein regions show
+  variable tolerance to biological indels, with indels more common in
+  extracellular and transmembrane proteins than in highly folded
+  enzymes and housekeeping genes \citep{delaChaux2007DNA}. This
+  suggests that well-folded protein regions will experience fewer
+  biological indels---and will therefore be less susceptible to
+  alignment error---than less structured regions.
 
@@
 phylogenetic trees than unfiltered alignments
-\citep{Dessimoz2010Phylogenetic}. \textcolor{red}{A recent study suggested that the
-benefit of alignment filtering (in terms of improved accuracy)
-outweighs the cost (in terms of reduced power) when applied to
-detecting positive selection, but the analysis was limited to a small
-range of possible evolutionary scenarios, as discussed below
-\citep{Privman2011Improving}.}  With the application of published
-filtering methods to alignments before testing for positive selection
-becoming standard practice
-\citep{Studer2008Pervasive,Aguileta2009Rapidly}, \textcolor{red}{continued investigation into the}
-potential benefits of alignment filtering to the detection of positive selection
-would seem well-warranted.
+\citep{Dessimoz2010Phylogenetic}. A recent study using a variety of
+filters suggested that the benefit of alignment filtering (in terms of
+improved accuracy) outweighs the cost (in terms of reduced power) when
+applied to detecting positive selection \citep{Privman2011Improving},
+but this analysis was limited to a small range of possible
+evolutionary scenarios as discussed below.  With the application of
+published filtering methods to alignments before testing for positive
+selection becoming standard practice
+\citep{Studer2008Pervasive,Aguileta2009Rapidly}, continued
+investigation of potential benefits of alignment filtering to the
+detection of positive selection seems well-warranted.
 
@@
 power of \sw evolutionary inference of positive selection. Our
-approach \textcolor{red}{is modeled on that of} Anisimova et al.
-\citeyearpar{Anisimova2002Accuracy} in that we use simulated protein
-alignments to evaluate methods for detecting \sw positive selection,
-considering each site to be an independent hypothesis test producing a
-positive or negative result. \textcolor{red}{As we simulate codon
-  alignments with indels and calculate inferred alignments using a
-  small but diverse sample of aligners and alignment filters, the
-  experimental design also shares some features with Fletcher and
-  Yang's \citeyearpar{Fletcher2010Effect} investigation of the
-  performance of the branch-site method for detecting positive
-  selection under alignment error and Privman et al.'s
-  study of the effect of alignment
-  filtering on detecting positive selection in HIV-1 genes. However,
-  the present work differs from previous studies in a number of
-  important ways.}
-
-\textcolor{red}{Unlike Fletcher and Yang, here we focus on \sw
-  detection of sitewise positive selection occurring throughout a
-  phylogeny (as opposed to genewise detection of selection acting at
-  specific branches) and we evaluate the impact of a number of
-  alignment filtering methods on the \sw analysis. Thus, the
-  biological hypothesis being investigated is different, and the
-  addition of filtering methods adds another layer of complexity to
-  our analysis.}
-
-\textcolor{red}{Privman et al., whose simulations centered on the
-  evolutionary chracteristics of three HIV-1 genes, concluded that
-  alignment filtering improves the performance of positive selection
-  inference by removing false positives. While this may be true for
-  the three genes (and perhaps for HIV-1 in general), HIV-1 is known
-  to evolve with widespread positive selection in the human host
-  \citep{Yang2003Widespread}. Results valid for these three HIV-1
-  genes may not be widely applicable to large-scale vertebrate and
-  mammalian comparative datasets, which exhibit less adaptive
+approach builds on those of Anisimova et al.
+\citeyearpar{Anisimova2002Accuracy}, Fletcher and Yang
+\citeyearpar{Fletcher2010Effect}, and Privman et
+al. \citeyearpar{Privman2011Improving}, using simulated protein
+alignments including insertions and deletions to evaluate methods for
+detecting \sw positive selection. Furthermore, we incorporate a
+diverse sample of aligners and alignment filters into an experimental
+design that differs from previous ones in a number of important ways.
+
+We focus on sitewise detection of positive selection occurring
+throughout a phylogeny and evaluate the impact of a number of
+alignment filtering methods on the \sw analysis. Thus, the biological
+hypothesis being investigated is different from that studied by
+Fletcher and Yang \citeyearpar{Fletcher2010Effect}, who focused on
+genewise selection acting at specific branches.  Privman et
+al. \citeyearpar{Privman2011Improving} recently published a related
+paper considering the evolutionary chracteristics of three HIV-1
+genes. They concluded that alignment filtering improves the
+performance of positive selection inference by reducing false positive
+results. While this may be true for these three genes (and perhaps for
+HIV-1 in general), HIV-1 is known to evolve with widespread positive
+selection in the human host \citep{Yang2003Widespread}. Results valid
+for these genes may not be widely applicable to large-scale vertebrate
+and mammalian comparative datasets, which exhibit less adaptive
 evolution \citep{Kosiol2008Patterns} and which comprise a larger
 diversity of protein structures and a wider range of species
-  divergence levels.}
-
-\textcolor{red}{We hypothesize that the divergence level and indel
-  rate, two highly variable and important evolutionary factors, may
-  strongly affect the performance of methods for alignment and
-  detection of selection. Accordingly, our simulations encompass a
-  wide range of biologically plausible indel rates and divergence
-  levels while fixing other parameters at values typical to those
-  encountered in the \sw analysis of vertebrate gene
-  families. Although our main purpose is to understand the behaviour
-  of various aligners and filters within our range of simulations, we
-  will make use of the above-noted methodological similarities in
-  order to highlight aspects of our results which agree with or
-  contradict the literature where possible.}
+divergence levels.
+
+We hypothesize that the divergence level and indel rate, two important
+evolutionary factors which are highly variable within and between
+different genomes, may strongly affect the performance of methods for
+alignment and detection of selection. Accordingly, our simulations
+encompass a wide range of biologically plausible indel rates and
+divergence levels while fixing other parameters at values typical to
+those encountered in the \sw analysis of vertebrate gene families.
 
@@
 
-\input{table2.txt}
+\begin{table}[p]
+\caption{Genome-wide Divergence Estimates for Commonly Analyzed Eukaryotes}
+\begin{center}
+\begin{threeparttable}
+
+\defcitealias{Hillier2004Sequence}{}
+\defcitealias{Siepel2005Evolutionarily}{}
+
+\footnotesize \tabcolsep 3pt
+\begin{tabular}{
+l r r l
+}
+\toprule
+ Species
+ & Pairwise $dS$
+ & Root-to-tip $dS$
+ & Reference
+\\
+\midrule
+   Human-Chimp & 0.01 & (0.005) & \citealp{Nei2010Neutral}
+\\ Human-Mouse & 0.43 & (0.215) & \citealp{Nei2010Neutral}
+\\ Human-Mouse & 0.5 - 0.8 & (0.25 - 0.4) & \citealp{Ogurtsov2004Indel}
+\\ Human-Chicken & 0.9 & (0.45) & \citealp{Nei2010Neutral}
+\\ Human-Chicken & 1.66 & (0.83) & Hillier et al., 2004 \citetalias{Hillier2004Sequence}
+\\ Human-Zebrafish & 1.38 & (0.69) & \citealp{Nei2010Neutral}
+\\ Vertebrates & -- & 0.75 & Siepel et al., 2005 \citetalias{Siepel2005Evolutionarily} 
+\\ Drosophila & -- & 1.0 & Siepel et al., 2005
+\\ Yeasts & -- & 1.25 & Siepel et al., 2005
+\\
+\bottomrule
+\label{table_2}
+\end{tabular}
+\begin{tablenotes}
+
+\item [] N{\sc ote}.---{\hskip -.1pt}The root-to-tip $dS$ is
+equivalent to the MPL (mean path length) used in our simulations. For
+two-species comparisons where the pairwise $dS$ was given, the
+root-to-tip $dS$ was calculated as half of the pairwise $dS$ and is
+included in parentheses.
+
+\end{tablenotes}
+\end{threeparttable}
+\end{center}
+\end{table}
 
@@
 lineages in the tree. We simulated alignments with MPL divergence
-between 0.05 and 2.0 synonymous substitutions per synonymous site,
+between 0.05 and 2.0 \syn substitutions per \syn site,
 spanning the range of evolutionary divergences observed in several
@@
 \begin{center}
-%\includegraphics[scale=0.5]{fig1.pdf}
-\includegraphics[scale=0.8]{fig1.pdf}
+\hidegraphics{scale=0.8}{fig1.pdf}
 \end{center}
@@
   were scaled to equal mean path length (MPL) divergence levels for
-  simulation. (a) A 6-taxon artificial tree used in previous
+  simulation. (A) A 6-taxon artificial tree used in previous
   simulations
-  \citep{Anisimova2001Accuracy,Massingham2005Detecting}. (b) A tree
+  \citep{Anisimova2001Accuracy,Massingham2005Detecting}. (B) A tree
   estimated from $\beta$-globin genes of 17 vertebrates and used in
   previous empirical analyses and simulation studies
-  \citep{Anisimova2001Accuracy,Anisimova2002Accuracy}. (c) The
+  \citep{Anisimova2001Accuracy,Anisimova2002Accuracy}. (C) The
   44-species tree used by the ENCODE project
@@
 The distribution of \sw selective pressures (embodied by the parameter
-$\omg$, the ratio of the rate of non-synonymous substitution to the
-rate of synonymous substitution) was modeled with a log-normal
+$\omg$, the ratio of the rate of \nsyn substitution to the
+rate of \syn substitution) was modeled with a log-normal
 distribution derived from a maximum-likelihood fit to a large dataset
@@
 simulation to correct for the difference between our definition of
-branch lengths as the number of synonymous substitutions per
-synonymous site ($dS$) and INDELible's interpretation of branch length
+branch lengths as the number of \syn substitutions per
+\syn site ($dS$) and INDELible's interpretation of branch length
 as the average number of substitutions per codon ($t$)
@@
 $t=3(NdN+SdS) = 3dS(\bar{\omg{}}N+S)$, where $N$ and $S$ are the
-proportion of nonsynonymous and synonymous sites and $\bar{\omg}$ is
+proportion of \nsyn and \syn sites and $\bar{\omg}$ is
 the mean $\omg$ across all sites. $S$ is approximately 0.3 when
@@
 
-Alignments were inferred using four alignment algorithms chosen for
+Alignments were inferred using six alignment algorithms chosen for
 their widespread use or demonstrated accuracy: ClustalW v1.82
-\citep{Thompson1994ClustalW}, MAFFT \citep{Katoh2005MAFFT} and two
-variants of PRANK \citep{Loytynoja2005From} based on an amino acid
-model (subsequently referred to as \pranka{}) or an empirical codon
-model (subsequently referred to as \prankc{}). Unaligned amino acid
+\citep{Thompson1994ClustalW}, MAFFT \citep{Katoh2005MAFFT}, \tcr{ProbCons
+\citep{Do2005a}, T-Coffee \citep{Notredame2000TCoffee}} and two variants of
+PRANK \citep{Loytynoja2008PhylogenyAware} based on an amino acid model
+(subsequently referred to as \pranka{}) or an empirical codon model
+(subsequently referred to as \prankc{}). Unaligned amino acid
 sequences were given as input to all alignment programs (except
@@
 
-GUIDANCE and T-Coffee produce individual scores for each residue,
+GUIDANCE and T-Coffee filters produce scores for each residue,
 allowing individual residues to be masked instead of entire
-columns. \textcolor{red}{Privman et al. \citeyearpar{Privman2011Improving} found a
+columns. Privman et al. \citeyearpar{Privman2011Improving} found a
 residue-based filter to be more effective than its column-based
-equivalent, so we filtered residues instead of alignment columns where
-possible}. GUIDANCE generates many replicate alignments, each using a
-slightly perturbed guide tree, with either MAFFT or \pranka as the
-bootstrap aligner. The program then assigns to each residue from the
-input alignment a score from 0 to 1 based on how consistently it was
-placed in the replicate alignments. In order to maximize the
-similarity between the input aligner and the bootstrap aligner, we ran
-GUIDANCE with 100 MAFFT replicates when filtering ClustalW alignments
-and with 30 \pranka replicates when filtering \prankc
-alignments. T-Coffee calculates the residue-wise consistency between
-an input multiple alignment and independently calculated pairwise
-alignments \citep{Notredame2003Using}, rounding and normalizing
-residue scores into integers between 0 and 9. T-Coffee was run using
-its default settings and the {\em evaluate\_mode
+equivalent, and we opted to filter residues instead of alignment
+columns where possible. GUIDANCE generates many replicate alignments,
+each using a slightly perturbed guide tree, with either MAFFT or
+\pranka as the bootstrap aligner. The program then assigns to each
+residue from the input alignment a score from 0 to 1 based on how
+consistently it was placed in the replicate alignments. In order to
+maximize the similarity between the input aligner and the bootstrap
+aligner, we ran GUIDANCE with 100 MAFFT replicates when filtering
+ClustalW alignments and with 30 \pranka replicates when filtering
+\prankc alignments. T-Coffee calculates the residue-wise consistency
+between an input multiple alignment and independently calculated
+pairwise alignments \citep{Notredame2003Using}, rounding and
+normalizing residue scores into integers between 0 and 9. T-Coffee was
+run using its default settings and the {\em evaluate\_mode
   --output=score\_ascii} command-line parameters to output alignment
@@
 \begin{center}
-%\includegraphics[scale=0.75]{fig2.pdf}
-\includegraphics[scale=1.25]{fig2.pdf}
+\hidegraphics{scale=1.25}{fig2.pdf}
 \end{center}
@@
 \subsection*{The Performance of Three Methods for Detecting \Sw Positive Selection}
-We first evaluated the ability of three \sw methods, \mtwo,
-\meight and SLR, to accurately estimate \sw $\omg$ values and to
-detect positive selection under a range of tree lengths in the absence
-of alignment error. Figure \ref{fig_2} shows the TPR, FPR, \tpr{} and \sw $\omg$
-correlation over a range of MPLs for each of the three simulation
-trees.
+We first evaluated the ability of three \sw methods, \mtwo, \meight
+and SLR, to accurately estimate \sw $\omg$ values and to detect
+positive selection under a range of tree lengths in the absence of
+alignment error. Figure \ref{fig_2} shows the TPR, FPR, \tpr{} and \sw
+$\omg$ correlation over a range of mean path lengths
+\tcr{(MPL, defined as the mean root-to-tip branch
+  length across all lineages)} for each of the three simulation trees.
 
@@
 decreasing at higher values. The downward trend in the 6-taxon tree
-was likely due to the impact of saturation of synonymous sites in the
+was likely due to the impact of saturation of \syn sites in the
 very long branches present in such a sparse tree at high divergence
@@
 textured lines, Figure \ref{fig_3}) analyzed using three different
-aligners (ClustalW, MAFFT, and \prankc) and the true alignment. (We
-omitted results for \pranka alignments from Figure \ref{fig_3} in
-order to reduce visual clutter; a full comparison of all aligners
-tested is shown in Figure \ref{fig_4}C and discussed in the next
-section.) For the indel simulations, the indel rate here was held
-constant at 0.1 indel event per substitution.
+aligners (ClustalW, MAFFT, and \prankc) and the true
+alignment. \tcrr{(Results for ProbCons, T-Coffee and \pranka
+  alignments are generally of intermediate quality, with ProbCons and
+  T-Coffee showing slightly higher TPR, slightly higher FPR, and very
+  similar \tpr{} compared to MAFFT, and \pranka showing performance
+  levels superior to these but inferior to \prankc. These results are
+  omitted from Figure \ref{fig_3} in order to reduce visual clutter;
+  \tpr results for \pranka are shown in Figure \ref{fig_4}C and
+  discussed in the next section, and a comparison of results from all
+  aligners tested can be found in Supplementary Figure S1.)} For the
+indel simulations, the indel rate here was held constant at 0.1 indel
+event per substitution.
 \begin{figure}[t]
 \begin{center}
-%\includegraphics[scale=0.75]{fig3.pdf}
-\includegraphics[scale=1.25]{fig3.pdf}
+\hidegraphics{scale=1.25}{fig3.pdf}
 \end{center}
@@
 performance characteristics, with \prankc always performing best and
-ClustalW performing worst. \textcolor{red}{The same ranking of
-  aligners with respect to detecting positive selection has been
-  observed in a number of studies
+ClustalW performing worst. The same ranking of aligners with respect
+to detecting positive selection has been observed in a number of
+studies
 \citep{Fletcher2010Effect,Markova-Raina2011High,Privman2011Improving};
-  our results corroborate these findings and provide evidence that
-  this ranking may be applicable across a wide range of divergence
-  levels and indel rates.}
+our results corroborate these findings and provide evidence that this
+ranking may be consistent across a wide range of divergence levels and
+indel rates.
 
@@
 \begin{center}
-%\includegraphics[scale=0.8]{fig4.pdf}
-\includegraphics[scale=0.8]{fig4.pdf}
+\hidegraphics{scale=0.8}{fig4.pdf}
 \end{center}
@@
   shapes (6-, 17- or 44-taxon) and a range of indel rates and mean
-  path length (MPL) divergence levels. Alignments were inferred with
+  path length (MPL) divergence levels. Alignments are inferred with
   one of four aligners (ClustalW, MAFFT, \pranka{}, \prankc) and
@@
   given (indel rate, MPL) pair as measured by one of three summary
-  statistics: (a) the true positive rate (TPR) at the recommended
-  cutoff threshold, (b) the false positive rate (FPR) at the
-  recommended cutoff threshold, or (c) the TPR at a 1\% FPR
-  threshold. Results for MAFFT and \pranka are omitted from (a) and
-  (b); as in (c) they show characteristics intermediate between
+  statistics: (A) the true positive rate (TPR) at the recommended
+  cutoff threshold, (B) the false positive rate (FPR) at the
+  recommended cutoff threshold, or (C) the TPR at a 1\% FPR
+  threshold. Results for MAFFT and \pranka are omitted from (A) and
+  (B); as in (C) they show characteristics intermediate between
   ClustalW and \prankc.}
@@
 heatmaps of the TPR and FPR for ClustalW, \prankc and the true
-alignment (Figure \ref{fig_4}A,B) and a heatmap of the
-error-controlled \tpr for all aligners tested (Figure
-\ref{fig_4}C). (MAFFT and \pranka were omitted from Figure
-\ref{fig_4}A--B to save space, but their performance fell between that
-of ClustalW and \prankc for all performance measures, with \pranka
-slightly outperforming MAFFT; a comprehensive set of TPR, FPR, and
-\tpr results can be found in Supplementary Figure S1.)
-The results from Figure \ref{fig_3}, which were simulated with an
-indel rate of 0.1, correspond to the middle row of each panel in
-Figure \ref{fig_4}; rows above and below the middle row represent
-higher and lower indel rates, respectively. Similarly, the bottom row
-of each panel in Figure \ref{fig_4} was simulated with an indel rate
-of zero and corresponds to the `No Indels' data in Figure \ref{fig_3}.
+alignment (Figure \ref{fig_4}A, B) and a heatmap of the
+error-controlled \tpr for all aligners tested (Figure \ref{fig_4}C).
+\tcrr{(MAFFT and \pranka are omitted from Figure \ref{fig_4}A, B and
+  ProbCons and T-Coffee are entirely omitted from Figure \ref{fig_4}
+  to save space. The performance of all these aligners fell between
+  that of ClustalW and \prankc for all our measures. \pranka slightly
+  outperforms MAFFT, and ProbCons and T-Coffee show similar
+  performance to MAFFT. A comprehensive set of TPR, FPR, and \tpr
+  results can be found in Supplementary Figure S1.)}  The results from
+Figure \ref{fig_3}, which were simulated with an indel rate of 0.1,
+correspond to the middle row of each panel in Figure \ref{fig_4}; rows
+above and below the middle row represent higher and lower indel rates,
+respectively. Similarly, the bottom row of each panel in Figure
+\ref{fig_4} was simulated with an indel rate of zero and corresponds
+to the `No Indels' data in Figure \ref{fig_3}.
 
@@
 \begin{center}
-%\includegraphics[scale=0.65]{fig5.pdf}
-\includegraphics[scale=0.65]{fig5.pdf}
+\hidegraphics{scale=0.65}{fig5.pdf}
 \end{center}
@@
   of indel rates and mean path length (MPL) divergence
-  levels. Alignments were inferred using (a) ClustalW or (b) \prankc,
+  levels. Alignments were inferred using (A) ClustalW or (B) \prankc,
   either left unfiltered or filtered with one of four alignment
@@
 performance through the introduction of false negatives and false
-positives, we tested whether alignment filtering methods
-could reduce error rates and improve the power of \sw detection of
-positive selection. Using sequences simulated from the 17-taxon tree
-and a range of indel rates and divergence levels, we calculated
-inferred alignments using ClustalW and \prankc and applied four
-filtering methods before performing the \sw analysis. Since we wished
-to determine whether alignment filters either improved or worsened the
+positives, we tested whether alignment filtering methods could reduce
+error rates and improve the power of \sw detection of positive
+selection. Using sequences simulated from the 17-taxon tree and a
+range of indel rates and divergence levels, we calculated inferred
+alignments using ClustalW and \prankc and applied four filtering
+methods before performing the \sw analysis. Since we wished to
+determine whether alignment filters either improved or worsened the
 error rates and power of \sw analysis, we measured the ratio of each
@@
 unfiltered alignments. These relative values are presented in Figure
-\ref{fig_5}.
+\ref{fig_5}. \tcrr{(Filtering results for \pranka, ProbCons and
+  T-Coffee were not calculated, and results for MAFFT are omitted from
+  Figure \ref{fig_5} to save space. The gain or loss in performance
+  resulting from filtering MAFFT alignments was generally intermediate
+  to that resulting from filtering ClustalW or \prankc alignments. A
+  comprehensive set of filtering results, including MAFFT alignments
+  and \tprf values, can be found in Supplementary Figure S2.)}
 
@@
 of the FPR was very low in both datasets under those
-conditions. \textcolor{red}{Figure \ref{fig_4}B shows the FPR to be on
-  the order of $5\times10^{-4}$ under those conditions, and we
-  observed 63 false positives for the true alignment and 67 false
-  positives for the \prankc alignment across all 150 replicates at the
-  indel rate of 0.04 and MPL of 0.8}. A similar slight FPR elevation
-was also observed at the same indel rate for the optimal, GUIDANCE and
-T-Coffee filters.
+conditions. Figure \ref{fig_4}B shows the FPR to be on the order of
+$5\times10^{-4}$ under those conditions; in total, we observed 63
+false positives for the true alignment and 67 false positives for the
+\prankc alignment at the indel rate of 0.04 and MPL of 0.8 across all
+150 replicates (comprising ca. 75,000 analyzed sites). A similar
+slight FPR elevation was also observed at the same indel rate for the
+optimal, GUIDANCE and T-Coffee filters.
 
@@
 false negative at a positively-selected site: either misalignment of
-one or more nonhomologous codons causing the positive signal to be
+one or more \nh codons causing the positive signal to be
 masked, or non-alignment of homologous codons causing the amount of
@@
 ability to improve sitewise performance, behaving qualitatively
-similarly to the `optimal' filter. In order to compare the TPR, FPR
-and \tpr values obtained with filtered ClustalW alignments to those
-obtained with unfiltered alignments, we ran ClustalW with GUIDANCE
-filtering using all three trees and the same range of indel rates and
-MPLs as used for Figure \ref{fig_4}; these results are included in
-Supplementary Figure S1. GUIDANCE filtering generally
-improved the performance of ClustalW alignments, resulting in sitewise
-performance comparable to (but not better than) unfiltered MAFFT
-alignments.
+similarly to the optimal filter.
+
+\tcrr{Since the performance measurements shown in Figure \ref{fig_5}
+  are expressed relative to values obtained with unfiltered
+  alignments, they do not allow for easy comparison of the absolute
+  performance of any combination of aligner and filter. To more
+  directly compare the absolute TPR, FPR and \tpr values obtained with
+  filtered ClustalW and MAFFT alignments to those obtained with other
+  unfiltered alignments, we simulated additional datasets with
+  ClustalW and MAFFT alignments and GUIDANCE filtering using all three
+  trees and the same range of indel rates and MPLs as used for Figure
+  \ref{fig_4}. These results are included in Supplementary Figure
+  S1. Comparing absolute performance, we find that GUIDANCE filtering
+  generally improved the \tpr for ClustalW and MAFFT alignments,
+  largely due to strong FPR reductions in regions of high indel rates
+  and low divergence levels. The resulting \tpr performance for
+  filtered ClustalW alignments was comparable to (but not better than)
+  unfiltered MAFFT alignments, and the \tpr for filtered MAFFT
+  alignments was slightly better than unfiltered MAFFT alignments and
+  substantially worse than unfiltered \pranka alignments.}
 
 Filtering was less beneficial when applied to the more accurate
-\prankc alignments, with T-Coffee reducing performance and GUIDANCE
-yielding only mild \tpr improvements. Importantly, GUIDANCE only
-showed improved performance at high divergence levels (e.g.,
+\prankc alignments, with T-Coffee filtering reducing performance and
+GUIDANCE yielding only mild \tpr improvements. Importantly, GUIDANCE
+only showed improved performance at high divergence levels (e.g.,
 MPL$>$1.6), well above those found in commonly-analyzed groups of
-species. \textcolor{red}{Thus, the use of unfiltered \prankc alignments would yield
-largely equivalent performance to GUIDANCE-filtered alignments for detecting
-\sw positive selection when analyzing protein-coding sequences across
-a wide range of indel rates and commonly encountered sequence
-divergence levels.
-On the other hand, GUIDANCE (when judiciously applied, including an upper limit on the amount of sequence data removed) never significantly reduced performance compared to the unfiltered alignments.} Finally, we note that the performance of the `optimal' filter on \prankc
-alignments suggests that mild further improvements to filtering
-strategies may be possible, but the potential for improvement is small
-and may be of little value.
+species. Thus, the use of unfiltered \prankc alignments would yield
+largely equivalent performance to GUIDANCE-filtered alignments for
+detecting \sw positive selection when analyzing protein-coding
+sequences at most commonly encountered divergence levels. Equally
+important is the observation that GUIDANCE (when judiciously applied,
+including an upper limit on the amount of sequence data removed) did
+not significantly reduce performance compared to the unfiltered
+alignments. Put simply, filtering neither significantly hurt nor
+significantly improved performance. Finally, we note that the
+performance of the `optimal' filter on \prankc alignments suggests
+that mild further improvements to filtering strategies may be
+possible, but the potential for improvement is small and may be of
+little value.
 
@@
 the analysis of typical mammalian protein-coding genes. We
-evaluated the ability of four alignment methods and three alignment
+evaluated the ability of six alignment methods and three alignment
 filtering methods to produce alignments for detecting positively
@@
 alignment conditions. We confirmed and extended the findings of
-Fletcher and Yang \citeyearpar{Fletcher2010Effect}, Markova-Raina et al. \citeyearpar{Markova-Raina2011High}, and Privman et al. \citeyearpar{Privman2011Improving} regarding the
-relative accuracy of different aligners, showing that \prankc had the
-best performance and ClustalW had the worst performance for subsequent
-sitewise analysis. Notably, our simulations found that ClustalW
-produced more sitewise false positives than MAFFT or \prankc even at
-low MPLs, suggesting that its use should be avoided even when
+Fletcher and Yang \citeyearpar{Fletcher2010Effect}, Markova-Raina et
+al. \citeyearpar{Markova-Raina2011High}, and Privman et
+al. \citeyearpar{Privman2011Improving} regarding the relative accuracy
+of different aligners, showing that \prankc had the best performance
+and ClustalW had the worst performance for subsequent sitewise
+analysis. Notably, our simulations found that ClustalW produced more
+sitewise false positives than any other aligner tested even at low
+divergence levels, suggesting that its use should be avoided even when
 analyzing closely-related sequences. \prankc, on the other hand,
-resulted in very low FPRs even at higher MPLs. In particular, when the
+resulted in very low FPRs even at higher divergences. In particular, when the
 number of sequences in the tree was large, \prankc{}'s sitewise FPRs
@@
 
-The very low FPRs observed for \prankc alignments conflicted somewhat
+The very low FPRs observed for \prankc alignments conflict somewhat
 with the results of Fletcher and Yang
@@
 false positives when misalignment causes apparent positive selection
-along only the foreground branch, while the SLR sitewise test would produce
-false positives only when misalignment causes a signal of positive
-selection strong enough to overpower the non-positive signal
+along only the foreground branch, while the SLR sitewise test would
+produce false positives only when misalignment causes a signal of
+positive selection strong enough to overpower the non-positive signal
 throughout the tree. This effect stems from the different biological
@@
 either of these tests to detect positive selection within an
-alignment.
+alignment. \tcr{For the detection of positive selection in highly
+  divergent or indel-prone sequences, the use of sitewise models
+  instead of the branch-site test may be a sensible alternative,
+  sacrificing some sensitivity for better error control.}
 
@@
 reduces the beneficial effect of alignment filtering, so much so that
-current filtering methods are \textcolor{red}{scarcely able to} improve the
+current filtering methods are scarcely able to improve the
 performance of \prankc alignments when analyzed with SLR.
@@
 error-controlled power under difficult alignment conditions at high
-divergence levels (whereas T-Coffee and Gblocks failed to improve upon
-any \prankc alignments) this improvement was modest in magnitude and
+divergence levels (whereas T-Coffee and Gblocks filtering failed to improve upon
+any \prankc alignments), this improvement was modest in magnitude and
 was achieved largely through the recovery of false negatives as
@@
 computational expense of running many bootstrap alignment replicates
-(as performed by GUIDANCE) may not be worth the possibility of a
-slight increase in power. On the other hand, judicious application of
-GUIDANCE never significantly reduced power, so its use would not be
-expected to yield worse results.
-
-\textcolor{red}{Some of our results and conclusions differed from those of Privman et
+(as performed by GUIDANCE) may not be offset by the possibility of a
+slight increase in power. However, GUIDANCE never significantly
+reduced power, so its use would not be expected to yield worse
+results.
+
+Some of our conclusions differ from those of Privman et
 al. \citeyearpar{Privman2011Improving}, who found strong improvements
@@
 end of our tested range (MPLs of 0.38, 0.34 and 0.33 for gag, pol and
-env, respectively; E. Privman, personal communication) and were roughly
-comparable in size and shape to our 17-taxon tree, we failed to find a
-significant benefit for alignment filters at any MPL below 1.2 when
-using \prankc alignments. Interestingly, the authors also found
-Gblocks to be roughly comparable in performance to GUIDANCE, while we
-found Gblocks' performance to be very poor. Some of these
+env, respectively; E. Privman, personal communication) and were
+roughly comparable in size and shape to our 17-taxon tree, we failed
+to find a significant benefit for alignment filters at any MPL below
+1.2 when using \prankc alignments. Interestingly, the authors also
+found Gblocks to be roughly comparable in performance to GUIDANCE,
+while we found Gblocks' performance to be very poor. Some of these
 discrepancies may be due to differences in the details of our
@@
 shown that divergence levels and indel rates have a significant impact
-on the accuracy and sensitivity of different aligners.}
+on the performance of different aligners and filters.
 
@@
 spatial or temporal variability in the rate of indel formation or in
-the distribution of selective pressures \citep{Whelan2008Spatial}, but
-we do not expect that such heterogeneity would affect our main
+the distribution of selective pressures (e.g.; \citealt{Whelan2008Spatial}). We
+do not expect that such heterogeneity would affect our main
 conclusions regarding the relative performance of different aligners
-and filters. The trends we observed were consistent across a wide
+and filters: the trends we observed were consistent across a wide
 range of parameter values and tree sizes, suggesting that they reflect
@@
 
-However, the apropriateness of our simulation scheme should be
-critically considered when evaluating our specific power and error
-rate estimates in the context of real-world data analysis. Functional
-differences between genes may influence the $\omg$ distribution and
-indel rate---with some genes or domains showing lower or higher
-tolerance to indels and fewer or more neutrally-evolving sites than
-modeled in our simulations---making false positive results more or
-less likely. Second, species-level differences may also be important,
-as the efficacy of natural selection is highly dependent on effective
-population size \citep{Ellegren2009Selection}. For example, proteins
-evolving in \Dr species with a high effective population size should
-experience stronger positive and purifying selection than in mammals,
-potentially leading to increased power and reduced error rates when
-compared to our simulations based on a mammalian-like $\omg$
-distribution.
+\tcr{However, such heterogeneity is clearly important to the evolution
+  of mammalian proteins \citep{Fay2003}. Many proteins contain combinations of
+  structured domains and unstructured regions along their length,
+  resulting in a discontinuous mix of different structure- and
+  function-related evolutionary pressures. False positives may be more
+  prominent in small unconstrained regions, raising FPR levels above
+  those predicted by simulations with a uniform pattern of
+  evolution. Functional differences between genes may also influence
+  the $\omg$ distribution, with some genes or domains showing fewer or
+  more neutrally-evolving sites than modeled in our simulations,
+  making false positive results either less or more likely,
+  respectively. As such, the appropriateness of our simulation scheme
+  should be critically considered when evaluating our specific power
+  and error rate estimates in the context of real-world data
+  analysis.}
+
+\tcr{In the case of a protein evolving with a mix of insertion and
+  deletion rates, our results based on uniform rates could be used to
+  identify tentative upper and lower bounds for the overall error
+  rate. For example, if a small region of a protein is evolving with
+  higher divergence and indel rates than the rest, then the error
+  rates within the less-constrained region should be comparable to our
+  results based on proteins with uniformly high divergence and indel
+  rate. Correspondingly, the overall error rate for the protein would
+  fall somewhere between those observed in our homogeneous simulations
+  corresponding to the least difficult and most difficult regions of
+  the alignment. Critical to such an analysis is the accurate
+  estimation of the local indel rate, for which a number of methods
+  are currently available or under development
+  \citep{Holmes2005,Cartwright2009Problems}.}
+
+Species-level differences may also have an effect on error rates in
+detecting positive selection, as the efficacy of natural selection is
+highly dependent on effective population size
+\citep{Ellegren2009Selection}. For example, proteins evolving in \Dr
+species with a high effective population size should experience
+stronger positive and purifying selection than in mammals, potentially
+leading to increased power and reduced error rates when compared to
+our simulations based on a mammalian-like $\omg$ distribution.
 

### 2012-02-08 projects/2xmammals/tree.R
Add tree.R

@@ -1,0 +1,530 @@
+# Extracts the NHX annotations from a tree and returns a list with the annotations and the
+# tree string (with NHX stuff removed).
+read.nhx.tree <- function(str) {
+  if (file.exists(str)) {
+    # We're reading a file -- load the file into a string.
+    str <- readLines(str)
+  }
+
+  nhx.matches <- gregexpr("[,\\)\\( ]?([^,\\)\\(]+)?(:?\\d*\\.?\\d*)?\\[.*?\\]", str)
+  matches <- nhx.matches[[1]]
+  match.pos <- as.numeric(matches)
+  match.len <- attr(matches, 'match.length')
+
+  if (match.len[1] == -1) {
+    return(read.tree(text=str))
+  }
+
+  match.pos <- match.pos + 1
+  match.len <- match.len - 1
+
+  #print(match.pos)
+  nhx.strings <- substring(str, match.pos, match.pos+match.len-1)
+  #print(nhx.strings)
+
+  labels <- gsub("\\[&&NHX.*", "", nhx.strings)
+  labels.no.bl <- gsub(":.*", "", labels)
+
+  #print(labels.no.bl)
+
+  # Go through and splice the stripped-down labels back into the string.
+  for (i in 1:length(match.pos)) {
+    new.label <- gsub(labels.no.bl[i], paste('zzz',i,'zzz',sep=''), labels[i])
+    str <- paste(substr(str, 0, match.pos[i]-1 ), new.label, substr(str, match.pos[i] + match.len[i], nchar(str)), sep='')
+    match.pos <- match.pos - match.len[i] + nchar(new.label)
+  }
+  
+  # Parse the Phylo object from the cleaned-up string.
+  #print(str)
+  tree <- read.tree(text=str)
+
+  #print(str(tree))
+
+  # Create a list of NHX annotations keyed by the node ID.
+  tag.values <- gsub(".*\\[&&NHX:(.*)\\]", "\\1", nhx.strings)
+  tagval.list <- strsplit(tag.values, ":")
+  names(tagval.list) <- labels.no.bl
+
+  library(plyr)
+  map.list <- llply(tagval.list, function(x) {
+    list.out <- list()
+    cur.list <- strsplit(x, "=")
+    if (length(cur.list) > 0) {
+      for (i in 1:length(cur.list)) {
+        vec <- cur.list[[i]]
+        list.out[vec[1]] = vec[2]
+      }
+    }
+    return(list.out)
+  })
+
+  # Replace the labels with the true labels.
[…]

### 2012-02-08 projects/2xmammals/analyze_filters.R
Sanger update.

@@ -588,7 +588,46 @@ taxid.to.alias <- function(taxids, binomials=F, include.internals=F, save.df=T) 
     df <- get(df.key, envir=.GlobalEnv)
   }
 
+  # Add some manual fix-ups.
+  tax.ids <- c(
+    '1000000',
+    '1000001',
+    '1000002',
+    '1000003',
+    '1000004',
+    '1000005',
+    '1000006',
+    '1000007',
+    '1000008',
+    '1000009',
+    '1000010'
+  )
+  tax.lbls <- c(
+    'Human/Chimp ancestor',
+    'Squirrel/Rodent ancestor',
+    'Cow/Dolphin ancestor',
+    'Horse/etc ancestor',
+    'Bat/Carnivore ancestor',
+    'Horse/Bat ancestor',
+    'Cow/Dolphin/Alpaca ancestor',
+    'Procavia/Loxodontia ancestor',
+    'Atlantogenata',
+    'Laurasia/Euarchontoglires ancestor',
+    'Primate/Tupaia ancestor'
+    )
+  df <- rbind.fill(df, data.frame(
+    taxon_id = tax.ids,
+    name = tax.lbls,
+    stringsAsFactors=F
+  ))
+  df <- rbind.fill(df, data.frame(
+    taxon_id = '314295',
+    name = 'Hominoidea',
+    stringsAsFactors=F
+  ))
+
   #print(df)
+  #print(taxids)
 
   if (missing(taxids)) {
     print("  Missing some taxids!")

### 2012-02-08 projects/2xmammals/analyze_genes.R

@@ -1,13 +1,8 @@
 uname  <- Sys.getenv("USER")
 if (uname == 'gj1') {
-  source("~/src/greg-ensembl/projects/2xmammals/aln.R")
-  source("~/src/greg-ensembl/projects/2xmammals/analyze_mammals.R")
-#  source("~/src/greg-ensembl/scripts/go_enrichments.R")
   source("~/src/greg-ensembl/scripts/liftOver.R")
   source("~/src/greg-ensembl/projects/phylosim/PhyloSimPlots.R")
 } else {
-  source("~/lib/greg-ensembl/projects/2xmammals/aln.R")
-  source("~/lib/greg-ensembl/projects/2xmammals/analyze_mammals.R")
 #  source("~/lib/greg-ensembl/scripts/go_enrichments.R")
   source("~/lib/greg-ensembl/scripts/liftOver.R")
   source("~/lib/greg-ensembl/projects/phylosim/PhyloSimPlots.R")

### 2012-02-08 projects/2xmammals/analyze_substitutions.R

@@ -146,20 +146,17 @@ nsyn.plot <- function(pset.id=1, test=F, clean=F) {
   print(out.df)
 
   out.df <- ddply(subs.sites, .(site_id), function(x) {
-    n.nsyn <- sum(x$mut_nsyn)
-    n.syn <- sum(x$mut_syn)
-    nsyn.sites <- subset(x, mut_nsyn == 1)
-    nsyn.taxa <- paste(sort(unique(nsyn.sites$taxon_alias)), collapse=', ')
-    n.nsyn.taxa <- length(unique(nsyn.sites$taxon_id))
+    x <- subset(x, mut_nsyn == 1)
+    nsyn.taxa <- paste(sort(unique(x$taxon_alias)), collapse=', ')
+    n.nsyn.taxa <- length(unique(x$taxon_id))
     data.frame(
-      nsyn = n.nsyn,
-      syn = n.syn,
       nsyn.taxa = nsyn.taxa,
-      n.nsyn.taxa = n.nsyn.taxa
+      n.nsyn.taxa = n.nsyn.taxa,
+      stringsAsFactors=F
     )
   })
-  print(table(out.df$nsyn))
-  print(table(out.df$nsyn.taxa))
+  print(tail(sort(table(out.df$nsyn.taxa)), n=20))
+  print(table(out.df$n.nsyn.taxa))
 
   xx <- subset(subs.sites, mut_nsyn == 1)
   unique.positions <- as.factor(xx$site_id)

### 2012-02-08 projects/2xmammals/estimate_m0_tree.pl

@@ -58,6 +58,11 @@ if (!-e $file_out || $clean) {
   }
   print "  running PAML $model...\n";
   my $treeI = Bio::EnsEMBL::Compara::TreeUtils->to_treeI($tree);
+  foreach my $node ($treeI->nodes) {
+    if (!$node->is_leaf) {
+      $node->name('');
+    }
+  }
   my $results = Bio::Greg::Codeml->branch_model_likelihood( $treeI, $aln, $phylo->worker_temp_directory, $params );
   $lines = $results->{lines};
 
@@ -93,7 +98,8 @@ foreach my $node (@nodes) {
   $node->remove_tag('id');
   $node->remove_tag('branch_label');
 }
-my $nhx_str = $codeml_tree->as_text('nhx');
+Bio::EnsEMBL::Compara::TreeUtils->transfer_annotations($codeml_tree, $treeI);
+my $nhx_str = $treeI->as_text('nhx');
 print $nhx_str."\n";
 
 open(OUT, '>', $nhx_out);

### 2012-02-08 projects/2xmammals/estimate_trees.R

@@ -1,21 +1,26 @@
 bsub.estimate.m0 <- function() {
   newicks.df <- get.subset.newicks()
 
-  species.groups <- newicks.df$name
-  #species.groups <- c('Primates')
+  psets <- newicks.df$pset_id
+  #psets <- c(1)
 
   subdirs <- get.output.subdirs()
   subdirs <- c('concat_nongaps_20')
 
   out.f <- output.folder(T, T, T)
   for (subdir in subdirs) {
-    for (species.group in species.groups) {
-      for (n.codons in c(1e4, 1e5, 5e5, 1e6, 2e6)) {
+    for (pset in psets) {
+      for (n.codons in c(1e3)) {
+        test <- FALSE
         aln.in <- 'all_combined.fasta'
         #aln.in <- 'alns_1.fasta'
         #aln.in <- '../00/EIF3A.fasta'
-        xtra <- paste(out.f, subdir, species.group, aln.in, n.codons, TRUE, sep=' ')
-        bsub.function('estimate_m0_tree', mem=16, extra.args=xtra, queue='long')
+        xtra <- paste(out.f, subdir, pset, aln.in, n.codons, test, sep=' ')
+        
+        required.mem <- n.codons / 100000 * 3
+        required.mem <- ceiling(required.mem)
+        required.mem <- max(12, required.mem)
+        bsub.function('estimate_m0_tree', mem=required.mem, extra.args=xtra, queue='long')
       }
     }
   }
@@ -24,20 +29,24 @@ bsub.estimate.m0 <- function() {
 bsub.estimate.branch <- function() {
   newicks.df <- get.subset.newicks()
 
-  species.groups <- newicks.df$name
-  species.groups <- c('Primates')
+  psets <- newicks.df$pset_id
+  psets <- c(1)
 
   subdirs <- get.output.subdirs()
   subdirs <- c('concat_nongaps_20')
 
   out.f <- output.folder(T, T, T)
   for (subdir in subdirs) {
-    for (species.group in species.groups) {
+    for (pset in psets) {
 #      for (n.codons in c(1e4, 1e5, 5e5, 1e6, 2e6)) {
-      for (n.codons in c(1e4, 1e5)) {
+      for (n.codons in c(1e3)) {
[…]

### 2012-02-08 projects/2xmammals/sites_scripts.R

@@ -1,5 +1,7 @@
 uname  <- Sys.getenv("USER")
 if (uname == 'gj1') {
+  source("~/src/greg-ensembl/projects/2xmammals/aln.R")
+  source("~/src/greg-ensembl/projects/2xmammals/tree.R")
   source("~/src/greg-ensembl/projects/2xmammals/analyze_mammals.R")
   source("~/src/greg-ensembl/projects/2xmammals/analyze_genes.R")
   source("~/src/greg-ensembl/projects/2xmammals/analyze_substitutions.R")
@@ -8,6 +10,8 @@ if (uname == 'gj1') {
   source("~/src/greg-ensembl/projects/2xmammals/calc_recomb.R")
   source("~/src/greg-ensembl/scripts/xtable_utils.R")
 } else {
+  source("~/lib/greg-ensembl/projects/2xmammals/aln.R")
+  source("~/lib/greg-ensembl/projects/2xmammals/tree.R")
   source("~/lib/greg-ensembl/projects/2xmammals/analyze_mammals.R")
   source("~/lib/greg-ensembl/projects/2xmammals/analyze_genes.R")
   source("~/lib/greg-ensembl/projects/2xmammals/analyze_substitutions.R")
@@ -118,12 +122,12 @@ concat_alns <- function(remove.paralogs, mask.clusters, mask.nongaps, n.jobs=1, 
   concat.alns(remove.paralogs, mask.clusters, mask.nongaps, n.jobs, subset.index)
 }
 
-estimate_m0_tree <- function(out.f, subdir, species.group, aln.in, n.codons, clean=F) {
-  estimate.m0.tree(out.f, subdir, species.group, aln.in, n.codons, clean)
+estimate_m0_tree <- function(out.f, subdir, pset, aln.in, n.codons, clean=F) {
+  estimate.m0.tree(out.f, subdir, pset, aln.in, n.codons, clean)
 }
 
-estimate_branch_tree <- function(out.f, subdir, species.group, aln.in, n.codons, clean=F) {
-  estimate.branch.tree(out.f, subdir, species.group, aln.in, n.codons, clean)
+estimate_branch_tree <- function(out.f, subdir, pset, aln.in, n.codons, clean=F) {
+  estimate.branch.tree(out.f, subdir, pset, aln.in, n.codons, clean)
 }
 
 collect_alns <- function(chr.s, test=F) {

### 2012-02-08 projects/subclades/lsf_job.R
Add subclades directory.

@@ -1,0 +1,53 @@
+uname  <- Sys.getenv("USER")
+lib <- ifelse(uname == 'gj1', 'src', 'lib')
+
+source(sprintf("~/%s/greg-ensembl/projects/subclades/subclades_sim.R", lib))
+
+args <- commandArgs(trailingOnly=T)
+main <- function() {
+  if (!is.na(args[1])) {
+    fn.name <- args[1]
+    print(paste("Calling", fn.name))
+
+    jobindex <- Sys.getenv('LSB_JOBINDEX')
+    if (jobindex != '' && as.integer(jobindex) > 0) {
+      args <- c(args, as.integer(jobindex))
+    }
+    print(args)
+
+    error.f <- function(e) {
+      print("###ERROR###")
+      print(as.character(e))
+    }
+
+    tryCatch(
+      do.call(fn.name, as.list(args[2:length(args)])),
+      error = error.f
+    )
+  }
+}
+
+ifebi <- function(a, b) {
+  ifelse(Sys.getenv("USER") == 'greg', return(a), return(b))
+}
+
+bsub.function <- function(fn_name, queue='normal', mem=3, extra.args='', jobarray=NULL, jobarray_id=fn_name) {
+  array_s = ''
+  if (!is.null(jobarray)) {
+    array_s <- paste('-J ', jobarray_id, '[1-', jobarray, ']', sep='')
+  }
+  args_s <- paste(fn_name, ' ', extra.args, sep='')
+  cmd <- ifebi(
+    sprintf('bsub -q research -M%d000 %s -o "/homes/greg/scratch/lsf_logs/%s_%%J_%%I.txt" "R-2.12.0 --vanilla --args %s < lsf_job.R"',
+      mem, array_s, fn_name, args_s
+    ),
+    sprintf('bsub -q %s -R "select[mem>%d000] rusage[mem=%d000]" -M%d000000 %s -o "/nfs/users/nfs_g/gj1/scratch/lsf_logs/%s_%%J_%%I.txt" "/software/R-2.13.0/bin/R --vanilla --args %s < lsf_job.R"',
+      queue, mem, mem, mem,
+      array_s, fn_name, args_s
+    )
+  )
+  print(cmd)
+  system(cmd)
+}
+
+main()

### 2012-02-08 projects/subclades/subclades_sim.R

@@ -1,0 +1,40 @@
+lib <- ifelse(uname == 'gj1', 'src', 'lib')
+source(sprintf("~/%s/greg-ensembl/projects/2xmammals/aln.R", lib))
+source(sprintf("~/%s/greg-ensembl/projects/2xmammals/tree.R", lib))
+
+work.f <- function(str) {
+  lib <- ifelse(uname == 'gj1', 'src', 'lib')
+  normalizePath(sprintf("~/%s/greg-ensembl/projects/subclades/%s", lib, str))
+}
+
+scratch.f <- function(str) {
+  lib <- ifelse(uname == 'gj1', 'src', 'lib')
+  normalizePath(sprintf("~/scratch/subclades/%s", dir))
+}
+
+bsub.sim.alns <- function() {
+  species.groups <- c('Primates', 'Glires', 'Laurasiatheria', 'Atlantogenata')
+  pop.sizes <- c(1e4, 5e4, 1e5, 5e5, 1e6)
+  total.bls <- c(0.25, 0.5, 1, 2, 3)
+
+  for (species in species.groups) {
+    for (pop.size in pop.sizes) {
+      for (total.bl in total.bls) {
+        xtra <- paste(species, pop.size, total.bl, sep=' ')
+        bsub.function('sim.aln', extra.args=xtra)
+        return()
+      }
+    }
+  }
+}
+
+sim.aln <- function(species, pop.size, total.bl) {
+  print(species)
+  print(as.numeric(pop.size))
+  print(as.numeric(total.bl))
+
+  tree.f <- work.f(paste(species, '.nh', sep=''))
+  tree <- read.tree(tree.f)
+
+  print(tree)
+}
